# <yrChapterNumber>50. </yrChapterNumber> Where to find datasets

***Note that the following was compiled circa `r format(Sys.Date(), "%b %d, %Y")`.***

There are many many sources of data that can be found online. These 
resources are constantly being updated and new ones are constantly 
becoming available. The following are just some ideas to get started with
your search for data. 


## Datasets in R packages

There are many R packages that come with datasets.
By "dataset" I mean any data that is referenced in an R variable. 
This can be a dataframe, a vector, a list, etc. etc. etc.
Below are just a few example R packages that contain datasets. 
There are many, many others. Also make sure to see the section below about
R pacakges that offer access to APIs.

| Package name | Description |
|--------------|-------------|
| nycflights13 | data about all flights coming into or out of NYC in 2013. |
| Lahman | Extensive data about baseball teams, players, etc. |
| babynames | data provided by USA Social Security administration about births from 1880 through 2017 |
| stringr | Contains vectors fruits,sentences,words that the help uses in examples. |
| etc ... | ... |


To use the data in a package you should
first install the package. As an example we'll use the 
"nycflights13" package which 
contains data about all airline flights that came into or out of 
New York City in 2013. 

```{r}
#| warning: false

if(!require("nycflights13")){install.packages("nycflights13");require("nycflights13");}
```

We can now view which datasets are in the package with the following command:

```{r}
data(package="nycflights13")
```

To see the list of datasets in ALL packages that you have installed, you 
can use the following command

```{r}
data(package=rownames(installed.packages()))
```

To access the data you can use the syntax
PACKAGE_NAME::VARIABLE_NAME. Alternatively you can load the package
by using the library or require functions and then access the data
by just using the variable name.

```{r}
# This will work after you've installed the nycflights13 package
# even before you've loaded it.

head(nycflights13::flights)   # show the first few rows


# Alternatively if you have already loaded the package with the 
# library or require functions (which we did above) you can access
# the flights variable without specifying nycflights::

head(flights)   # show the first few rows
```

- nycflights13 package
  
  * install.packages("nycflights13")  # install the package
    library(nycflights13)             # load the package
      
    help(package="nycflights13")    # see what is included in the package
    


    
## Project Gutenberg

<https://www.gutenberg.org/>

This website hosts complete text of many many books whose copyright has expired.
If you are planning to work with "text" files, then you should search 
for the "UTF-8" or "ASCII" or "Plain Text" versions.

## Search for "Open Data"

<https://www.google.com/search?q=open+data>

Search for "open data" and MANY websites, especially government
based websites will come up. For example:
  
  * <https://data.gov/> - US Government Open Data
  
  * <https://data.ny.gov/> - New York ***State***
  
  * <https://opendata.cityofnewyork.us/> - New York ***City***
  
  * <https://data.nasa.gov/>

  * many, many others ...

## Google dataset search

Google search engine for datasets: <https://datasetsearch.research.google.com/>  

## Kaggle

<https://www.kaggle.com/datasets>  

This is a popular website for machine learning and data science 
competitions, projects and tutorials. There are many datasets hosted
on this platform




