# <yrChapterNumber>21. </yrChapterNumber> JSON Parsing Practice in R

<!--
This was created in Claude.ai
URL: https://claude.ai/chat/209c9b49-139c-4577-b5dc-1aaa8a9329e9
TOPIC: Parsing JSON Data in R: Practice Exercises
-->

<style>
h3 { color:blue; }
</style>

These exercises ask you to retrieve specific information from JSON
that has been processed by the fromJSON() function.

Each exercise includes some JSON and then specific questions that ask 
you to retrieve specific data from that JSON.

The questions are rated easy,medium,harder based on the following:

**easy questions**: data is readily accessible by 
simply using list, dataframe and matrix notation

**medium questions**: you might need additional functions such as unlist() or others.

**harder questions**: require to you to use either apply functions or loops to solve the problem.

```{r}
library(jsonlite)
```

## Exercise: Basic JSON Object Access

Answer the questions below by referring to
the following JSON response from an API.

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'weather_data'
weather_data <- fromJSON('{
  "location": "Seattle",
  "current": {
    "temperature": 12.5,
    "conditions": "Cloudy",
    "wind": {
      "speed": 15.5,
      "direction": "NW"
    }
  },
  "forecast": [
    {"day": "Monday", "high": 14, "low": 8},
    {"day": "Tuesday", "high": 16, "low": 9},
    {"day": "Wednesday", "high": 15, "low": 7}
  ]
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(weather_data)

# THE DATA
weather_data
```
</details>

### Question (easy): Write R code to extract the current temperature.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
weather_data$current$temperature
# or alternatively
weather_data[["current"]][["temperature"]]
```
</details>

### Question (easy): Write R code to create a vector of all forecasted high temperatures.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
weather_data$forecast$high
# or
weather_data$forecast[,"high"]
```
</details>

## Exercise: E-commerce Product Data

Answer the questions below by referring to the following JSON response from an e-commerce API.

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'product_data'
product_data <- fromJSON('{
  "store": "TechStore",
  "products": [
    {
      "id": "p123",
      "name": "Wireless Headphones",
      "price": 89.99,
      "stock": 45,
      "colors": ["black", "white", "blue"],
      "specs": {
        "battery": "20 hours",
        "connectivity": "Bluetooth 5.0",
        "weight": "250g"
      }
    },
    {
      "id": "p124",
      "name": "Smart Watch",
      "price": 199.99,
      "stock": 28,
      "colors": ["black", "silver"],
      "specs": {
        "battery": "48 hours",
        "connectivity": "Bluetooth 5.0",
        "weight": "45g"
      }
    },
    {
      "id": "p125",
      "name": "Wireless Charger",
      "price": 29.99,
      "stock": 120,
      "colors": ["black", "white"],
      "specs": {
        "input": "QC 3.0",
        "output": "15W",
        "weight": "80g"
      }
    }
  ],
  "last_updated": "2024-02-20"
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(product_data)

# THE DATA
product_data
```
</details>

### Question (easy): Write R code to extract the name and price of the second product.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
# Extract name
product_data$products$name[2]
# or
product_data$products[2, "name"]

# Extract price
product_data$products$price[2]
# or
product_data$products[2, "price"]

# Combined as a named vector
c(name = product_data$products$name[2], price = product_data$products$price[2])
```
</details>

### Question (easy): Write R code to extract all available colors for the Wireless Headphones product.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
# Find the row index for the Wireless Headphones
headphone_index <- which(product_data$products$name == "Wireless Headphones")

# Extract the colors
product_data$products$colors[[headphone_index]]
```
</details>

### Question (easy): Write R code to create a data frame showing each product's name, price, and stock.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
data.frame(
  name = product_data$products$name,
  price = product_data$products$price,
  stock = product_data$products$stock
)

# Alternatively
product_data$products[, c("name", "price", "stock")]
```
</details>

## Exercise: Restaurant Menu API

Answer the questions below by referring to the following JSON response from a restaurant menu API.

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'menu_data'
menu_data <- fromJSON('{
  "restaurant": {
    "name": "Flavor Haven",
    "location": "123 Main St",
    "hours": "11am-10pm",
    "menu": {
      "appetizers": [
        {"item": "Garlic Bread", "price": 5.99, "vegetarian": true},
        {"item": "Buffalo Wings", "price": 9.99, "vegetarian": false},
        {"item": "Mozzarella Sticks", "price": 7.99, "vegetarian": true}
      ],
      "main_courses": [
        {"item": "Margherita Pizza", "price": 14.99, "vegetarian": true, "sizes": ["small", "medium", "large"]},
        {"item": "Beef Burger", "price": 12.99, "vegetarian": false, "sides": ["fries", "salad"]},
        {"item": "Fettuccine Alfredo", "price": 13.99, "vegetarian": true, "options": ["add chicken"]}
      ],
      "desserts": [
        {"item": "Chocolate Cake", "price": 6.99, "vegetarian": true},
        {"item": "Ice Cream", "price": 4.99, "vegetarian": true, "flavors": ["vanilla", "chocolate", "strawberry"]}
      ]
    }
  }
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(menu_data)

# THE DATA
menu_data
```
</details>

### Question (easy): Write R code to extract the name and location of the restaurant.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
# Extract name
menu_data$restaurant$name

# Extract location
menu_data$restaurant$location

# Combined as a named vector
c(name = menu_data$restaurant$name, location = menu_data$restaurant$location)
```
</details>

### Question (easy): Write R code to extract all vegetarian items from the appetizers section.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
# Filter the appetizers data frame for vegetarian items
vegetarian_apps <- menu_data$restaurant$menu$appetizers[menu_data$restaurant$menu$appetizers$vegetarian == TRUE, ]

# Just the names of vegetarian appetizers
vegetarian_app_names <- menu_data$restaurant$menu$appetizers$item[menu_data$restaurant$menu$appetizers$vegetarian == TRUE]

# Show the result
vegetarian_apps
# or
vegetarian_app_names
```
</details>

### Question (easy): Write R code to calculate the average price of all main courses.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
mean(menu_data$restaurant$menu$main_courses$price)
```
</details>

## Exercise: Nested Arrays

Answer the questions below by referring to
the following JSON response from an API.

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'library_data'
library_data <- fromJSON('{
  "library": {
    "name": "Central Library",
    "books": [
      {
        "title": "Data Science Basics",
        "authors": ["Smith, J.", "Jones, K."],
        "categories": ["programming", "statistics"],
        "ratings": [4.5, 4.8, 4.2]
      },
      {
        "title": "Statistics Using R",
        "authors": ["Smith, J.", "Jones, K."],
        "categories": ["statistics", "R"],
        "ratings": [4.5, 4.8, 4.2]
      },
      {
        "title": "R Programming Guide",
        "authors": ["Wilson, M."],
        "categories": ["programming", "R"],
        "ratings": [4.7, 4.6, 4.9, 4.5]
      }
    ]
  }
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(library_data)

# THE DATA
library_data
```
</details>

### Question (medium): Write code to create a vector of all unique book categories 
across all books.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
unique(unlist(library_data$library$books$categories))
```
</details>

### Question (harder): get named vector of ratings for each book

Calculate the average rating for each book and store it 
in a named vector where names are book titles.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
bookTitles = sapply(library_data$library$books$ratings, mean)
names(bookTitles) = library_data$library$books$title
bookTitles
```
</details>

## Exercise: Complex Nested Structures

Answer the questions below by referring to
the following JSON response from a social media API.  
The API was called with the following URL: 
https://api.somesite.com/userinfo/Alex Chen

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'social_data'
# The response includes info about a single user "Alex Chen"
social_data <- fromJSON('{
  "user": {
    "id": "u123",
    "name": "Alex Chen",
    "posts": [
      {
        "id": "p1",
        "content": "Learning R!",
        "timestamp": "2024-02-15",
        "comments": [
          {
            "user": "Carol Wu",
            "text": "Check out tidyverse",
            "likes": 5
          },
          {
            "user": "Bob Smith",
            "text": "Great choice!",
            "likes": 3
          },
          {
            "user": "Carol Wu",
            "text": "tidyverse makes many things easier",
            "likes": 2
          }
        ]
      },
      {
        "id": "p2",
        "content": "JSON parsing is fun",
        "timestamp": "2024-02-16",
        "comments": [
          {
            "user": "Mike Jones",
            "text": "Try jsonlite",
            "likes": 4
          }
        ]
      },
      {
        "id": "p3",
        "content": "Coffee is key :)",
        "timestamp": "2024-02-16",
        "comments": [
          {
            "user": "Carol Wu",
            "text": "I totally agree!",
            "likes": 1
          },
          {
            "user": "Mike Jones",
            "text": "yessss",
            "likes": 0
          }
        ]
      }

    ]
  }
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(social_data)

# THE DATA
social_data
```
</details>

### Question (harder): Create a data frame containing each comment's user and number of likes.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
# Method 1 - extract vectors separately and combine into a dataframe
users = unlist(sapply(social_data$user$posts$comments, function(x) x$user))
likes = unlist(sapply(social_data$user$posts$comments, function(x) x$likes))
data.frame(user=users, likes=likes)

# Method 2: Using do.call to combine list of data frames
do.call(rbind, lapply(social_data$user$posts$comments, function(comments) {
  comments
}))[, c("user", "likes")]

# Method 3: Using tidyverse if available
library(tidyverse)
bind_rows(social_data$user$posts$comments)[,c("user","likes")]
```
</details>

### Question (harder): function returning all comments by specific user

Create a function that takes the social_data object and a 
username as parameters, then returns all comments made by that user.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
find_user_comments <- function(data, username) {
  
  # CommentsListDf is a list of dataframes. Each dataframe in the
  # list contains the columns: user, text, likes
  commentsListDf = data$user$posts$comments
  
  # commentsList is a list of vectors. The vectors contain just the comments for 
  # the specified username. There is one vector in the list from each of
  # the original dataframes.
  commentsList = lapply(commentsListDf, function(df){ df[df$user==username, "text"]})
  
  # Combine all of the vectors in the list into a single vector.
  return( unlist(commentsList) )
}

# Test the function
find_user_comments(social_data, "Carol Wu")
```
</details>

## Exercise: Working with Lists of Data Frames

Answer the questions below by referring to
the following JSON response from a fitness tracking API:

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see JSON</summary>
```{r}
# API response stored in variable 'fitness_data'
fitness_data <- fromJSON('{
  "user_id": "f789",
  "tracking_data": {
    "daily_steps": [
      {"date": "2024-02-14", "steps": 8432, "active_minutes": 45},
      {"date": "2024-02-15", "steps": 10234, "active_minutes": 62},
      {"date": "2024-02-16", "steps": 7321, "active_minutes": 38}
    ],
    "workouts": [
      {
        "type": "running",
        "sessions": [
          {"date": "2024-02-14", "duration": 30, "distance": 5.2},
          {"date": "2024-02-14", "duration": 25, "distance": 3.9},
          {"date": "2024-02-16", "duration": 25, "distance": 4.1}
        ]
      },
      {
        "type": "cycling",
        "sessions": [
          {"date": "2024-02-14", "duration": 30, "distance": 9.1},
          {"date": "2024-02-15", "duration": 45, "distance": 15.3}
        ]
      }
    ]
  }
}')
```
</details>

<details style="background-color:#DDD;padding:5px;">
<summary>Click to see fromJSON() output</summary>
```{r}
# THE STRUCTURE
str(fitness_data)

# THE DATA
fitness_data
```
</details>

### Question (harder): Create a data frame with total distance covered for each type of workout.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```{r}
workout_summary <- data.frame(
  type = fitness_data$tracking_data$workouts$type,
  total_distance = sapply(fitness_data$tracking_data$workouts$sessions, 
                         function(x) sum(x$distance))
)

workout_summary
```
</details>

<!-- YRTODO 
4.2 has a bug. I'm not even sure what the response is supposed to look like.
Perhaps ask in the chat from claude.ai.

Bonus question I didn't even look at this yet.

Fix both of these.


### Question: function to get all list of all activities for specific date

Create a function that takes the fitness_data object and a date as parameters, 
then returns a list containing all fitness activities 
(both steps and workouts) for that date.

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```r
get_daily_activities <- function(data, target_date) {
  # Get steps for the date
  daily_steps <- subset(data$tracking_data$daily_steps,
                       date == target_date)
  
  # Get workouts for the date
  workouts <- lapply(data$tracking_data$workouts, function(workout) {
    sessions <- subset(workout$sessions, date == target_date)
    if (nrow(sessions) > 0) {
      cbind(type = workout$type, sessions)
    } else {
      NULL
    }
  })
  
  # Remove NULL entries from workouts
  workouts <- Filter(Negate(is.null), workouts)
  
  list(
    steps = daily_steps,
    workouts = workouts
  )
}

# Test the function
get_daily_activities(fitness_data, "2024-02-14")
```
</details>

## Bonus Challenge

Create a function that recursively traverses any JSON object and returns a 
character vector of all possible access paths to leaf nodes (scalar values). 
For example:

```r
# Example usage:
get_json_paths(weather_data)
# Should return something like:
# [1] "location"
# [2] "current$temperature"
# [3] "current$conditions"
# [4] "current$wind$speed"
# [5] "current$wind$direction"
# ...
```

<details style="border:1px solid blue;padding:5px;">
<summary>Click for answer</summary>
```r
get_json_paths <- function(obj, prefix = "") {
  paths <- character()
  
  if (is.list(obj)) {
    for (name in names(obj)) {
      new_prefix <- if (prefix == "") name else paste0(prefix, "$", name)
      paths <- c(paths, get_json_paths(obj[[name]], new_prefix))
    }
  } else if (is.data.frame(obj)) {
    for (col in names(obj)) {
      new_prefix <- if (prefix == "") col else paste0(prefix, "$", col)
      paths <- c(paths, new_prefix)
    }
  } else {
    paths <- prefix
  }
  
  return(paths)
}
```
</details>

This challenge helps students understand the recursive nature of JSON structures and reinforces their understanding of how R represents these nested data structures internally.

-->