# <yrChapterNumber>14. </yrChapterNumber> regular expressions (regex)

```{r}
rm(list=ls())
```

## Intro to Regular Expressions (Also Known As "regex")

Regular expressions (AKA "regex") are patterns that are used to identify specific
types of data. They can be used to search through and slice and dice
character (i.e. textual) data in a variety of ways. The best 
way to understand is just to dive right in, which we will do below. 

Before we get started it's important to note that 
regular expressions are not specic to R. Rather regular expressions
are a concept that are available to use in most programming languages
(e.g. Python, Java, etc.) as well as in many text editors and in other
technical environments (e.g. Bash). There are slightly different versions
of regular expression syntax found in different environments. We will 
be focusing mainly on how to use regular expressions in R. However, this
knowledge carries over directly to using regex in other languages and
technical environments. 

## Some resources for learning regular expressions

The following is one of the many online tutorials.

* <https://ryanstutorials.net/regular-expressions-tutorial/>

The following R help pages are relevant to regular expressions.
However, they are hard to understand without a little intro first. I 
recommend that you go through the material in this book first or in 
other tutorials. Then you can refer to the help pages when necessary.

**Relevant R help pages**

```{r}
#| eval: false

?regex     
?grep      
?strsplit  
```

R's stringr package also includes several functions that can 
be used with regex. For now though, we will focus mainly on the "Base R" regex
functions. Once you understand those, you will be very prepared to understand 
functions in the stringr package that utilize regular expressions. 

## Data for examples

Before we start, let's define some data to be used with examples in this file.
*(NOTE: I made up N. American apple and S. Korean Fig so that I can use*
*them in some examples.)

### fruit vector

```{r}
#| file: createData-vecFruit-v001.R
#| echo: true
```

```{r}
fruit
```

### addresses vector

```{r}
#| file: createData-vecAddresses-v001.R
#| echo: true
```

```{r}
addresses

# show each address, one per line
cat(addresses, sep="\n")
```

### famousQuotes dataframe

```{r}
#| file: createData-famousQuotesDf-v001.R
#| echo: true
```

## The stringr package

### stringr::words vector

The stringr package includes some data that we will use in the examples. 

```{r}
#-------------------------------------------------------------------------.
# NOTE - most of the examples in this file were created using the data above.
# The stringr package also contains some data that can be used to experiment
# with these functions. 
#-------------------------------------------------------------------------.

if(!require(stringr)){install.packages("stringr");require(stringr)}

stringr::words
head(stringr::words, 100)
```

## grep function

```{r}
# Show all words that
# "start with a p, end with a y (with anything in the middle)"
grep(stringr::words, pattern="^p.*y$", value=TRUE)

# Starts with a p, ends with a y, nothing in the middle.
# Only matches "py". 
# There are no words that match.
grep(stringr::words, pattern="^py$", value=TRUE)

# match any word that start with p, ends with y and has a single
# character between them
grep(stringr::words, pattern="^p.y$", value=TRUE)

# match any word that start with p, ends with y and
# has exactly two characters between them.
grep(stringr::words, pattern="^p..y$", value=TRUE)

# match any word that start with p, ends with y and
# has exactly four characters between them.
grep(stringr::words, pattern="^p....y$", value=TRUE)

# match any sequence of characters between the p and the y
grep(stringr::words, pattern="^p.*y$", value=TRUE)

# starts with a p
grep(stringr::words, pattern="^p", value=TRUE)
```

## grep and grepl

```{r}
# When value=FALSE grep returns the positions in the vector of 
# values that matched
grep(stringr::words, pattern="^p.*y$", value=TRUE)
grep(stringr::words, pattern="^p.*y$", value=FALSE)

# default is value=FALSE
grep(stringr::words, pattern="^p.*y$", value=FALSE)
```


```{r}
#@ grep and grepl
#@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
#@ grep stands for "Globally search for a Regular Expression and Print the result"
#@
#@ Grep will search through the entries in a character vector and display those
#@ entries that match a specified pattern (see examples below). These patterns
#@ are known as regular expressions or "regex".
#@
#@ The history of grep started with a a command that was used on the Unix operating
#@ system. It has been adapted for use with many programming environments. R has
#@ its own version.
#@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@


# grep ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# grep returns character values or the indexes (i.e. position numbers) 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Find all fruit whose name contains the letter "h"
grep(pattern="h", x=fruit, value=TRUE)   # value=TRUE, show the acutal values that match the pattern 
grep(pattern="h", x=fruit, value=FALSE)  # value=FALSE, show the index (ie. position) of the values that match 


# grepl ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# grepl returns logical values (i.e. TRUE/FALSE vectors)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

grepl(pattern="h", x=fruit)    # find which values include an "h"
```

## Summary: 3 ways to use grep or grepl

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# - grep ( regexPattern , value=TRUE)  # returns the actual values that match  
# - grep ( regexPattern , value=FALSE) # returns the index numbers of the values that match  
# - grepl ( regexPattern )              # returns a logical vector that indicate which values match  
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# For now, let's focus on grep(... , value=TRUE) as it is easier to understand the results. 


# The pattern is searched for in the entire entry ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The pattern is considered "matched" if it appears anywhere in the data value.
# For example:   grep("h", fruit)
#
# returns all fruit that contain an "h", no matter whether the h is at the 
# beginning, end or middle of the word.
#
# You can change this behavior with the ^ and $ metacharacters (see below)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

## Spaces are NOT ignored. 

Spaces count as part of the pattern. They are NOT ignored.

```{r}
# fruits that contain a space
grep(pattern=" ", x=fruit, value=TRUE) # all fruit that contain a space
```

```{r}
# search for "k "  (i.e. k followed by a space)
grep("k ", fruit, value=TRUE) # "black cherry"

# search for "k"  (i.e. without a space - JUST a "k")
grep("k", fruit, value=TRUE) # "prickly pear"  "black cherry"  "kumquat"

# search for "ck" 
grep("ck", fruit, value=TRUE) # "prickly pear"  "black cherry"  "kumquat"

```

## regex patterns do NOT understand "numbers"

Digits are NOT treated as numbers. They are treated the same as any
other character. Therefore grep("12", SOME_VECTOR) will match any value
that contains a 1 followed by a 2, including "123" and "34321234".


```{r}
addresses # show all the addresses

grep("23", addresses, value=TRUE)  # matches anything that contains 23
```

## case sensitivity

By default, R's version of grep is case sensitive.

There are a few different approaches for changing the default behavior
to instead search case-INsensitively.

### (a) case INsensitive searches - use ignore.case argument

The first way - use ignore.case = TRUE. See the code below.

```{r}
grep("H",fruit, value=TRUE)  # contains a capital "H"
grep("h",fruit, value=TRUE)  # contains a lowercase "h"
grep("h", fruit, value=TRUE, ignore.case=TRUE) # contains AnY h
grep("H", fruit, value=TRUE, ignore.case=TRUE) # same thing
```

### (b) case INsensitive searches - character classes - e.g. [aA]

Another way to search for for both CAPITAL and lowercase characters, e.g. [Hh]
For example, [hH] indicates that h or H is valid to be matched. 
We will describe the exact meaning of the [square brackets] in a lot more 
detail below.

```{r}
grep("[hH]", fruit, value=TRUE)
```

### (c) case INsensitive searches - use toupper() and tolower()

another way using R's toupper or tolower functions

```{r}
msg = "She said 'Hello' to Joe."
msg
toupper(msg)
tolower(msg)

grep("h", tolower(fruit), value=TRUE)
```

## str_view() from the stringr package

The str_view function from the stringr package can be very helpful
when you're trying to understand a regular expression. str_view shows
exactly what parts of a string match the pattern. See the example below.

```{r}
# str_view is part of the stringr package
library(stringr)

greetings = c("hi there", "yo dude", "shalom", "bon jour")
cat(greetings, sep="\n")

# match the letter h in each greeting
str_view(greetings, "h")
```

## sub and gsub functions

sub and gsub functions are used to "substitute" the text that was matched 
by a regular expression with other text. 

The difference between sub() and gsub() is that in a single character value,
sub() function only substitutes the first part of the character value that 
matched the regex. By contrast, the gsub() function replaces EVERY part of 
the character value that matched the regex. (the "g" in "gsub" stands for
"global"). See the examples below.

IMPORTANT - both sub and gsub return the ENTIRE vector with only the 
values matched the regex being changed. This is different from the grep and
grepl functions that returned only those entries in the vector that
matched the regular expression.

```{r}
#@ sub and gsub functions ####
#@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
#@
#@ sub (SOME_REGEX_PATTERN, REPLACMENT, SOME_VECTOR)
#@    sub returns a new vector. The return value is the same as SOME_VECTOR
#@    except that the FIRST match of the pattern in each entry of SOME_VECTOR
#@    is replaced with REPLACEMENT - see the examples below.
#@
#@ gsub (SOME_REGEX_PATTERN, REPLACMENT, SOME_VECTOR)
#@    same as sub but ALL matches of the pattern are replaced (not just the
#@    first in each entry of the the vector - see the exmaples below
#@
#@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
```

```{r}
# QUESTION
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# replace the first letter "e" that appears in any fruit with the letter "X"
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# ANSWER
sub(pattern="e", replacement="X", x=fruit)  # "applX"    "N. AmXrican apple"     etc
```

```{r}
# QUESTION
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# replace ALL of the "e"s that appears in any fruit with the letter "x"
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# ANSWER
gsub(pattern="e", replacement="X", fruit)   # "applX"    "N. AmXrican applX"     etc
```

```{r}
# QUESTION
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# remove all spaces from the addresses
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# ANSWER
gsub(pattern=" ", replacement="", addresses)   # "12345SesameStreet"   "OneMicro$oftWay"   etc.
```

We will revisit sub and gsub later with more complex examples ...

## strsplit function

`strsplit()` is used to split a string based on a "delimeter" that appears
between the different values. This "delimeter" can be a regular expression.
We'll come back to strsplit later, but let's introduce it here.

```{r}
sentences = c("He said hi. She said bye. We went to the park.",
              "I like ice cream! Do you? Sue likes pizza.")
sentences
```

```{r}
#------------------------------------------------------------------------.
# QUESTION - 
# Use strsplit to split the values in the sentences vector by 
# splitting based on spaces. Assign the result to the varible "sentenceWords".
#
# Write code to get the 3rd "word" from the 1st entry in the sentences 
# vector.
#------------------------------------------------------------------------.

# ANSWER
sentenceWords = strsplit(sentences, split=" ")
sentenceWords

# Notice that the result is a LIST:
str(sentenceWords)

# Show the 3rd word in the 1st sentence
sentenceWords[[1]][3]
```

#### --- practice ---

```{r}
#------------------------------------------------------------------------.
# QUESTION - split each entry in the sentences variable into individual 
# sententces. 
# 
# WARNING - the value of the split argument is interpreted as a
# regular expression pattern. Be careful.
#------------------------------------------------------------------------.

# ANSWER

# 1st attempt - doesn't work. 
strsplit(sentences, ".")

# This doesn't work since the 2nd argument is a regular expression.
# The following will split based on periods.
sentences
strsplit(sentences, "\\.")

# Use a "regular expression" to instead split on any of a period, 
# question mark, or exclamation point.
sentences
strsplit(sentences, "[.?!]")  # split on any one of .?!
```

## Other functions: regmatches, regexec, regexpr, gregexpr

The following are other function in Base R that use regex. 
These are a little more advanced. It's probably better to try researching
these functions after first understanding the material presented in 
this section.

You can search online or see the R documentation for more info about these functions.

* regmatches
* regexec
* regexpr
* gregexpr

## stringr functions 

The stringr package includes many functions for use with character vectors.
One example is str_length, which is very similar to the nchar() function
in Base R.

```{r}
# The str_length function is part of the stringr package.
# To use it you must install stringr (or install tidyverse, which is a 
# collection of packages one of which is stringr)
str_length(c("abc", "hello", "I like ice cream!"))

# This function is very similar to the
# nchar function that is built into base R.
nchar(c("abc", "hello", "I like ice cream!"))  
```

The stringr package also includes numerous functions that make use
of regular expressions. The following is a table of the stringr functions
and the Base R functions that can be used to accomplish similar things.

| stringr                            | Base R                    | Description                            |
|------------------------------------|---------------------------|----------------------------------------|
| `str_detect()`                     | `grepl()`                 | Returns TRUE/FALSE if pattern is found |
| `str_extract(), str_extract_all()` | `regmatches()`            | Extract matching patterns              |
| `str_match(), str_match_all()`     | `regexec(), regmatches()` | Extract matched groups                 |
| `str_replace(), str_replace_all()` | `sub(), gsub()`           | Replace matched patterns               |
| `str_split()`                      | `strsplit()`              | Split string on pattern                |
| `str_subset()`                     | `grep(value = TRUE)`      | Keep strings matching pattern          |
| `str_locate(), str_locate_all()`   | `regexpr(), gregexpr()`   | Find positions of matches              |
| `str_count()`                      | `lengths(regmatches())`    | Count pattern occurrences             |

