# <yrChapterNumber>16. </yrChapterNumber> Escaping meta-characters and regex dialects

```{r}
#| file: 0200200-regex-setup.R
#| include: false
```

## Searching with regular expressions in a "text editor"

A "text editor" is a program that is used to edit "text files". A text file
can only contain "plain text" - i.e. no pictures, no music, only one font, 

RStudio's text editor

The text editor that is in RStudio can be used to create many different
types of files. For example, it can be used to create 
both "R Script files" (i.e. .R files) and "Quarto Documents" (i.e. .qmd files).
In addition, it can be used to create "plain text files". To do so, choose
the following menu choices from RStudio's menu: "File | New File | Text File"


```{r}
# NOTE: regular expressions are used in many different languages and environments.
# In general in regular expression in OTHER environments, 
# if you want to actually match a metacharacter (e.g. period, parentheses,
# caret, dollar sign, etc) you precede the metacharacter with a backslash.
#
# For example, you can do this in RStudio's text editor - just type ctrl-f or cmd-f
# and click "regex" checkbox. Then type your regular expression into the 
# search box.
# For example
# Try searching the addresses.txt file for the following in the RStudio text
# editor:
#
#   one|1
#   .
#   \.
#   $
#   \$
#
# Matching meta characters requires that you "escape" the meta-character
# by preceding it with a backslash e.g.  \.
```

## Matching meta-characters

```{r}
# Matching meta-characters ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# NOTE: regular expressions are used in many different languages and environments.
# In general in regular expression in OTHER environments, 
# if you want to actually match a metacharacter (e.g. period, parentheses,
# caret, dollar sign, etc) you precede the metacharacter with a backslash.
#
# For example, you can do this in RStudio's text editor - just type ctrl-f or cmd-f
# and click "regex" checkbox. Then type your regular expression into the 
# search box.
#
# For example
# Try searching the addresses.txt file for the following in the RStudio text
# editor:
#
#   one|1
#   .
#   \.
#   $
#   \$
#
# Matching meta characters requires that you "escape" the meta-character
# by preceding it with a backslash e.g.  \.
#
#
#
# When writing regular expression patterns in R you must use TWO \\'s to escape a metacharacter ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# HOWEVER, in R, since character values already use a backslash
# such as \n for a new line, you must use TWO backslashes in the regex
# pattern. The first backslash escapes the 2nd backslash from R
# so that R's character values don't interpret it in a special way.
# See the examples below.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

### Reminder of how backslashes (\) are used in R ####

```{r}
# Reminder of how backslashes (\) are used in R ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Remember that R uses backslashes to change the meaning (or to "escape" 
# the meaning) of the character that follows the backslash. 
# For example in the following cat command, \n, is displayed as a 
# "newline character" and \t is displayed as a tab.

cat("Hello\nJoe\thow are you\n\ndoing?\n\tI'm fine.")

# Similarly in the following cat command the \" escapes the meaning
# of the quote. It no longer implies the end of the quotation. The 
# meaning of \" is simply to include a quotation mark as part of the 
# text.

cat("Lincoln said \"Four score and seven years ago today...\"")

# If the following line were not commented it would cause an error
# because the quotation is not actually closed due to the \ before the
# final quotation mark.
#
#cat("This is a backslash: \")   # ERROR

# The following works correctly. Note that \\ is needed to escape
# the normal meaning of the backslash character!
#cat("This is a backslash: \\")   # ERROR

#cat("This is a period \.")   # ERROR \. is NOT an R escape sequence

# You must use TWO backslashes in R's regular expressions ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# In R you must use two backslashes in a regex pattern to 
# escape a metacharacter.


grep("\\.", addresses, value=TRUE)  # all addresses that contain a period

grep(".", addresses, value=TRUE)  # All the addresses

stuff = c("", "apple", "", "banana")
stuff
grep(".", stuff, value=FALSE)

grep(".", stuff, value=TRUE)

# This is an ERROR in R but would be correct in other 
# languages or environment that use regular expressions
#grep("\.", addresses, value="TRUE")  # ERROR - R doesn't recognize \.

# Without the backslash you will find all addresses that contain
# at least a single character (i.e. all the addresses)
grep(".", addresses, value="TRUE")  

grep("\\$", addresses, value=TRUE)   # addresses that contain a dollar sign

grep("$", addresses, value=TRUE)    # all addresses - why?? - they all have an ending
```

### To search for an actual backslash you must use 4 backslashes in the pattern

```{r}
# To search for an actual backslash you must use 4 backslashes in the pattern ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Searching for an actual backslash in the data can be tricky. 
# Remember, one of our addresses had a backslash in it. Let's find it.
#
# To look for a single backslash in the data you must use FOUR backslashes.
# Just as R character values need to "escape" a backslash with a 2nd backslash,
# so too do regular expressions need to escape a backslash with a 2nd backslash.
# Therefore if you want to write a regular expression in R that searches for
# a backslash, you must write FOUR backslashes in a row. The first two resolve
# to a single backslash. The 3rd and 4th resolve to a single backslash. Then finally
# the two single backslashes are used in the regular expression to match a 
# single actual backslash in the data.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

answer = grep("\\\\", addresses, value=TRUE)  # look for a single backslash in the data

answer

cat(answer)



stuff="\\\\"
stuff
cat(stuff)
```

## Matching QUOTES

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Matching QUOTES
#
# "Quotation marks" are NOT meta-characters in regular expressions. They
# have no special meaning in a regular expression. However, as with all 
# R code you must make sure to use a single backslash if the quotation mark
# is inside of quotation marks (e.g.  "\"" ) - see the example below.
#
# Note that when using R's regular expression functions, regex
# meta-characters, such as the period or ^ for which you want to remove 
# the special meaning require a DOUBLE backslash (as explained above).
#
# A regex pattern in VS Code (or a similar editor) that includes " or ' 
# would not need any backslashes since these aren't regex meta characters.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
quoteStuff = c("Joe says great stuff.", "Franklin said \"a penny saved ...\"")
quoteStuff
cat(quoteStuff, sep="\n")
grep ("\"", quoteStuff, value=TRUE) #  "Franklin said \"a penny saved ...\""


grep ("\\.", quoteStuff, value=TRUE) #  "Franklin said \"a penny saved ...\""
```

## Different "flavors" or "dialects" of regular expressions. ####

```{r}
# Different "flavors" or "dialects" of regular expressions. ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Regular expressions have been around for a long time. Different "dialects"
# of regular expressions have popped up over the years. 
#
# Some programming languages and tools use slightly different "rules" 
# for regular expressions. This can be frustrating. However, the basic set
# of regular expresion rules remains the same for most programming languages 
# and tools. 
#
# Regular expressions first became popular with the Unix operating system in 
# the 1970s. There were many different versions of Unix being marketed by 
# different companies, each with slight differences. POSIX is a standard that
# defines how things should be done in a standard way across all the different
# versions of Unix. POSIX addresses regular expressions too. 
#
# POSIX introduced "named character classes" as described below. R will 
# recognize these. 
#
# Other additions to the regular expression notation were introduced by 
# the once very popular Perl programming language. You can get these features
# to work in R by specifying perl=TRUE as one of the arguments for grep
# and other functions in R that work with regular expressions. 
# For more details about perl regular expressions, see ?regex.
#
# As we said above, regular expressions are NOT totally standardized across all 
# languages and environments. For example (as of Feb 10, 2022)
# there are subtle differences between the rules for regular expressions
# that are used in R and those that are used in the 
# Visual Studio Code (VS Code) text editor. You can see a summary of the 
# rules used by VS Code here:
#   https://docs.microsoft.com/en-us/visualstudio/ide/using-regular-expressions-in-visual-studio?view=vs-2022
#
# Although there may be some differences between different languaes and 
# environments, the vast majority of regular expression meta characters
# work the same across the different environments. 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

### [[:digit:]] vs \\d - different shorthand notations for character classes

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Two different shorthand notations for character classes
# - POSIX named character classes , e.g. [[:alnum:]]  [[:digit:]]   etc. 
# - backslash shortcuts , e.g. \s \S \d \D  etc.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

#### "POSIX" named character classes, e.g. [[:digit:]] ####

```{r}
# "POSIX" named character classes, e.g. [[:alnum:]] ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# NOTE: These are available in R. 
#
#       They currently are NOT available in VSCode
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Bracket notation in regular expressions (e.g. [aeiou] or [0-9]) are known as 
# character classes. 
#
# You can use several "named character classes" as shorthand for some common
# character classes. These are shown below. Notice the [[double brackets]]
# We'll explain more about the the [[double brackets]] below.
#
# [[:upper:]]   same as [A-Z]
# [[:lower:]]   same as [a-z]
# [[:space:]]   same as [ \r\n\t]
# [[:punct:]]   all "special" characters, eg. !@#$% etc...
# [[:digit:]]   same as [0-9]
# [[:alpha:]]   same as [a-zA-Z]
# [[:alnum:]]   same as [a-zA-Z0-9]
#
# The [[double brackets]] shown above are necessary since these
# "named character classes" must actually be placed inside a pair of 
# [square brackets]. For example, you can also use the named
# character classes inside a larger character class.
#
# For example the following will match any single character
# from the following list:    -,+,*,/,(,),0,1,2,3,4,5,6,7,8,9
#
#   [-+*/()[:digit:]]    is the same as [-+*/()0-9]
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

### --- practice ---

```{r}
# QUESTION
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Search for addresses that contain at least one digit. Use a POSIX
# named character class.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# ANSWER 
grep("[[:digit:]]", addresses, value=TRUE)   # uses POSIX named character classes

str_view(addresses, "[[:digit:]]")

# NOTE - the pattern "[:digit:]" with one set of [brackets] does NOT work.
#
# Since there is only one set of [brackets], the pattern matches any one of
# the characters that are between the [brackets], i.e. match
# one of the characters ":", "d", "i", "g", "i", "t" or ":"
# This is equivalent to "[:digt]"  (I removed the 2nd ":" and the 2nd "i" as
# they are repetitive.)

# THIS DOESN'T WORK! - see note above
grep("[:digit:]", addresses, value=TRUE)  # looks for one of the following :,d,i,g,i,t,:


grep("^[N[:digit:]]", addresses, value=TRUE)   # same as [N0-9]
```

```{r}
# QUESTION:
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Write a command that replaces any sequence of digits or mathematical 
# operators with the text "<<MATH-EXPRESSION>>"
#
# You can use the following "mathStuff" variable to test your answer.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

mathStuff <- c("What is 3+2 ? Do you know the answer?",
               "99.5 desgrees in Farenheit is 99.5*(5/9)-32 degrees in Celcius")

mathStuff
```

```{r}
# ANSWER

gsub("[-+*/().[:digit:]]+", "<<MATH-EXPRESSION>>", mathStuff)
```

```{r]}
# QUESTION:
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Write a grep command that matches punctuation and letters, but not numbers.
# You can use the following data to test your answer. Use POSIX named
# character classes.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

stuff = c("1234",  # This should NOT match since it doesn't contain letters or punctuation
          "12.34", # This SHOULD match since it contains punctuation.
          ".",     # This SHOULD match since it contains punctuation.
          "hi")    # This SHOULD match since it contains at least one letter
```

```{r}
# ANSWER

# The following will match any punctuation or letters but not numbers
grep("[[:punct:][:alpha:]]", stuff, value=TRUE)   #   "."    "hi"
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Search for fruits that contain spaces using the POSIX
# named character classes for spaces
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# ANSWER
grep("[[:space:]]", fruit, value=TRUE)
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Search fruit for those that contain punctuation (e.g. periods, commas, etc)
# using the POSIX named character classes
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# ANSWER

grep ("[[:punct:]]", fruit, value=TRUE)
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Show fruit that contain either an x,y,z or some punctuation.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# ANSWER

# NOTE that there are TWO sets of brackets. The POSIX named character
# class, [:punct:], is itself inside a set of [brackets].

grep("[xyz[:punct:]]", fruit, value=TRUE)
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Search all ADDRESSES for those that contain punctuation (e.g. periods,
# commas, etc) or actual digits (e.g. 0123456789) using POSIX named
# character classes
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# ANSWER

grep("[[:punct:][:digit:]]", addresses, value=TRUE)

# This also works
grep("[[:punct:]0-9]", addresses, value=TRUE)
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Search all ADDRESSES for those that contain some punctuation that 
# comes immediately after the letter t. Use POSIX named classes.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

```{r}
# ANSWER
grep("[tT][[:punct:]]", addresses, value=TRUE)

# ANOTHER WAY
grep("[tT][^[:alnum:]]", addresses, value=TRUE)
```

```{r}
# QUESTION 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# answer the previous question without using POSIX named classes.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

grep("[tT][,.!?]", addresses, value=TRUE)
```

#### backslash shortcuts for character classes, e.g. \s \S \d \D etc.

```{r}
# backslash shortcuts for character classes, e.g. \s \S \d \D etc. ####
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The following are also shorthand notation that you can use for some
# characer classes.
#
# Note that in R you must use a double backslash, e.g. \\s instead of \s
#
# \s   is the same as [ \n\t\r]  also same as [[:space:]]
#      It matches anything which is considered whitespace.
#      This could be a space, tab, line break etc.
#
# \S   is the same as [^ \n\t\r]
#      It matches the opposite of \s, that is anything which is not considered
#      whitespace.
#
# \d   is the same as [0-9]  (ie. it matches a single digit)  same as [[:digit:]]
#
# \D   is the same as [^0-9] (i.e. it matches a single NON-digit)
#
# \w - matches anything which is considered a word character. That is
#      [A-Za-z0-9_]. Note the inclusion of the underscore character '_'. This is
#      because in programming and other areas we regularly use the underscore as part
#      of, say, a variable or function name.
#
# \W - matches  [^A-Za-z0-9_] the opposite of \w, that is anything which is not considered a
#      word character.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```

