# <yrChapterNumber></yrChapterNumber> Working with Excel files

```{r}
#| warning: false

if(!require(rvest)){install.packages("rvest");require(rvest)}
if(!require(stringr)){install.packages("stringr");require(stringr)}
if(!require(dplyr)){install.packages("dplyr");require(dplyr)}
if(!require(lubridate)){install.packages("lubridate");require(lubridate)}
if(!require(kableExtra)){install.packages("kableExtra");require(kableExtra)}
```

There are many R packages that work with Excel files. Some of these packages
are easier to use while some are more complex but offer more functionality. 

There are several R packages that can be used to read and write data between R 
and Excel. However, many of these packages are limited in the extent of what
they can do. For example many packages, can only read the results of Excel 
formulas in R but cannot read the actual formulas into R.

When it comes to more advanced Excel features, such as charts and 
pivot tables, only some R packages offer any functionality.

The best way to learn about these packages is to search for current info.
Once you've identified a package that might be appropriate for your task it
would be a good idea to look at the help for the package 
(i.e. `help(package="SOME_PACKAGE")` as well as 
any vignettes that might be available - a good place to find these is 
on the CRAN website.

## Some suggested packages

### readxl

If all you want to do is to read the data from an Excel file into R, a 
good choice is the readxl pakcage that is part of the tidyverse.
See the tidyverse website or CRAN for more info. 

### 
If you are 

## A list of Excel related packages

The set of R packages on CRAN can be found here:
<https://cran.r-project.org/web/packages/available_packages_by_name.html>

To get a 
Below is the list of these packages (as of `r format(Sys.Date(), format="%b %d, %Y")`)
that contain the word "Excel" or the text "xls" or "xlsx" (i.e. the filename extensions for Excel files)
in either the pacakage name or the description

::: {.content-hidden}
```{r}
if(!require(rvest)){install.packages("rvest");require(rvest)}

page = read_html("https://cran.r-project.org/web/packages/available_packages_by_name.html")

pkgs = page %>% 
  html_element("table") %>%
  html_table()

names(pkgs) = c("pkg", "description")

excelPkgs = pkgs[ grepl("excel",pkgs$pkg,ignore.case=TRUE) |
            grepl("excel",pkgs$description,ignore.case=TRUE) , ]

widthPkgName = max(nchar(excelPkgs$pkg))

excelPkgs$description = stringr::str_wrap(excelPkgs$description, width = 78 - widthPkgName)

as.data.frame(excelPkgs)

#grep("excel", lines, ignore.case = TRUE, value=TRUE)
```
:::

## Control Excel (old versions on Windows) from R

<https://bert-toolkit.com/>


## with dplyr

```{r}
#| paged.print: false
#| echo: false

page = read_html("https://cran.r-project.org/web/packages/available_packages_by_name.html")

pkgs = page %>% 
  html_element("table") %>%
  html_table() %>%
  rename(c(pkg=1, description=2)) %>%
  filter(
    str_detect(pkg,regex("Excel", ignore_case=TRUE)) | 
      str_detect(description,regex("Excel",ignore_case=TRUE)) |
      str_detect(description,regex("xls",ignore_case=TRUE))
    ) %>%
  mutate(description = str_wrap(description, width = 70 - max(str_width(pkg)))) %>%
  kableExtra::kbl(caption="Excel packages") %>%
  kable_styling(bootstrap_options=c("striped","hover","condensed","responsive",full_width=FALSE,position="center",fixed_thead=TRUE))

pkgs

```
