# <yrChapterNumber>10. </yrChapterNumber> Using pipes, i.e. %>% and |>

## Intro to the "pipe", i.e. %>% from the magrittr package

```{r}
# The pipe symbol, i.e. %>% that we will be discussing below
# is part of the magrittr package. You can install/load magrittr directly.
# Alternatively, magrittr is automatically installed/loaded with most of the
# tidyverse packages. 
if(!require(magrittr)) {install.packages("magrittr");require(magrittr);}

# For more info see:
#
# help(package="magrittr")`

```

## Using the %>% operator.

The %>% operator is used to send the output of one command into the 
into a 2nd commmand. Specifically, the info 
on the left side of the %>% is sent into the first argument 
of the function on the right hand side of the %>%
See the example below.

```{r}
# The following is code that does NOT use a "pipe".
x = seq(4,6)
rep(x, 3)


# The following code uses a "pipe", i.e. %>%, to send
# the info from the left of the %>% into the first argument
# of the function that appears on the right hand side
# of the %>%
seq(4, 6) %>% rep(3)

# We can extend this further
seq(4, 6)  %>%  rep(3)  %>%  sum()

# To make it easier to read, we can put each function
# call on a separate line.
seq(4, 6) %>%
  rep(3) %>%
  sum()

# Notice in the code above that the %>% appears at the end of 
# each line. This is necessary to indicate to R that the command
# isn't finished yet.

# Without pipes, we could have coded the same thing using the
# following R code. This still works but the pipes make it easier
# to read and understand the code.

x = seq(4,6)
y = rep(x, 3)
sum(y)
```

## Passing a value to an argument that is not the 1st argument

If you want to send a value to an argument that is not the first,
you can use the period as shown in the code below.

```{r}
# define a vector of numbers
nums = c(1,3)

# This is without the pipe:
rep("hello", sum(nums))

# This is with the pipe. We use the . in the rep function to show
# where the output of the pipe should go.

sum(nums) %>%
  rep("hello", .)

```

## You can also use the |> operator

The %>% was and is still part of the magrittr package. The pipe was seen
as such a useful tool that as of R version 4.1 "base-R" started to include 
its own version of the pipe. The base-R version of the pipe is |> while
the magrittr version of the pipe is %>%

The pipe symbol that is now built into R (without needing to load the
magrittr package) provides basic pipe functionality. However,
if you want extra features (e.g. using the . to show where the data
should be placed) then you still need to use the magrittr version 
of the pipe.


```{r}

# The following is the new base-R version of the pipe
seq(4, 6) |>
  rep(3) |>
  sum()

# Unfortunately you cannot use |> to send the output to any location
# other than the 1st argument. For example:
nums = c(1,3)

# this doesn't work:
sum(nums) |>        # using the new base-R pipe symbol |>
  rep("hello", .)   

# but this does
sum(nums) %>%       # using magrittr pipe symbol %>%
  rep("hello", .)   

```
