# <yrChapterNumber>15. </yrChapterNumber> stringr tutorial

This is an overview of the stringr package that is part of the "tidyverse"
family of packages.  The info in this section comes from this youtube playlist:
<https://www.youtube.com/watch?v=oIu5jK8DeX8&list=PLiC1doDIe9rDwsUhd3FtN1XGCV2ES1xZ2>{target="_blank"}

<div style="font-style: italic; margin-left: 2em;"> 
See these resources for more info about the entire tidyverse
family of packages.

* <https://r4ds.hadley.nz/>{target="_blank"}
* <https://tidyverse.org/>{target="_blank"}

See these links for more info about the stringr package

* <https://stringr.tidyverse.org/>{target="_blank"}

See these links for more info about other related tidyverse packagaes

* tibble package: <https://tibble.tidyverse.org/>
* magrittr: <https://magrittr.tidyverse.org/>
* lubridate package: <https://lubridate.tidyverse.org/>
* hms package: <https://hms.tidyverse.org/>
* dplyr package: <https://dplyr.tidyverse.org/>
  - Also see this playlist for info about dplyr:
    <https://www.youtube.com/watch?v=THGFXV4RW8U&list=PLiC1doDIe9rC8RgWPAWqDETE-VbKOWfWl>
</div>

```{r}
#| echo: false
#| output: false

# Load in some data for use in some of the lessons
#library(RSQLite)
#library(tidyverse)

sub <- "MachineLearning" 
#db <- src_sqlite('../input/reddit-comments-may-2015/reddit-comments-may-2015/database.sqlite', create = F)

# Load the desired subset of data from the database
#db_subset <- db %>% 
#             tbl('May2015') %>% 
#             filter(subreddit == sub)
          
#data <- data.frame(db_subset)[, c("author","score","body")]

CRANrepos = c(
    "https://mirror.las.iastate.edu/CRAN/",       # Iowa State University, Iowa
    "http://ftp.ussg.iu.edu/CRAN/",               # Indiana University, Indiana
    "https://repo.miserver.it.umich.edu/cran/", # University of Michigan
    "https://cran.wustl.edu/",              # Washington University, Missouri
    "https://archive.linux.duke.edu/cran/",     # Duke University, NC
    "https://cran.case.edu/",                   # Case Western Reserve University, OH
    "https://ftp.osuosl.org/pub/cran/",         # Oregon State University
    "http://lib.stat.cmu.edu/R/CRAN/",    # Carnegie Mellon University, PA
    "https://cran.mirrors.hoobly.com/",         # Hoobly Classifieds, PA
    "https://mirrors.nics.utk.edu/cran/") # Nat. Inst. 4 Computational Sci, TN


if (!require(tidyverse)){install.packages("tidyverse", repos=CRANrepos);require(tidyverse)}


data = read.csv("data3.csv")
names(data) = c("author","score","body")
```

## Setting up the package for use

### Download the code with install.packages() function

Before using functions from any package, you must install the package using
the install.packages() function (see more below).

The install.packages() function, downloads the code for the R 
package(s) to your computer from a "repository" known as 
CRAN - the Comprehensive R Archive Network. CRAN is actually supported by 
different websites that are funded and managed by different institutions.
Each of these sources of package info is known as a CRAN "mirror".
If you don't specify where the the install.packages function should get the 
package, a window might popup asking your to choose a "CRAN mirror". 
Alternatively, you can specify the mirror you want in the **repos** argument
directly in the call to install.packages. This is what we do below.

You can install the stringr package with the following command.

```{r}
# The following is the list of major CRAN mirrors. You do NOT need to do this 
# part. It is done for this website since the code to generate this website
# is NOT run interactively and if a popup should appear asking us to choose 
# a "mirror" the website would not be generated correctly. Therefore we
# use this code to specify a list of mirror from which install.packages()
# can choose one. If the first one is not available (e.g. it is "down") then
# install.packages() will try the 2nd one, etc.

CRANrepos = c(
    "https://mirror.las.iastate.edu/CRAN/",       # Iowa State University, Iowa
    "http://ftp.ussg.iu.edu/CRAN/",               # Indiana University, Indiana
    "https://repo.miserver.it.umich.edu/cran/", # University of Michigan
    "https://cran.wustl.edu/",              # Washington University, Missouri
    "https://archive.linux.duke.edu/cran/",     # Duke University, NC
    "https://cran.case.edu/",                   # Case Western Reserve University, OH
    "https://ftp.osuosl.org/pub/cran/",         # Oregon State University
    "http://lib.stat.cmu.edu/R/CRAN/",    # Carnegie Mellon University, PA
    "https://cran.mirrors.hoobly.com/",         # Hoobly Classifieds, PA
    "https://mirrors.nics.utk.edu/cran/") # Nat. Inst. 4 Computational Sci, TN

# NOTE: If you are running this command interactively you can leave out the
# repos argument.
install.packages("stringr", repos=CRANrepos)
```

<em>
**Note** that stringr is part of the `tidyverse` family of packages.
You can install any tidyverse package by itself or install the entire 
set of tidyverse packages with the command `install.packages("tidyverse")`
</em>

### Calling functions from stringr

At this point the code is downloaded to your computer. 
You can now use the functions in stringr but you will have to prefix each
call to a function with `stringr::`

For example the stringr::str_length() function returns the length of the 
strings (i.e. character values) that are passed to it. 

```{r}
stuff = c("Hi", "there.", "How are you?")
stringr::str_length(stuff)
```

Notice that I must include `stringr::`

```{r}
#| warning: true
#| error: true

# Determine the length of each value in the vector
stuff = c("Hi", "there.", "How are you?")
str_length(stuff)

```

### library(stringr)  or   require(stringr)

You can use the library or require commands to avoid needing to write stringr::

```{r}
#| warning: true
#| error: true

library(stringr)   # require(stringr) will also work 

stuff = c("Hi", "there.", "How are you?")
str_length(stuff)
```


<em>
**Note** that stringr is part of the `tidyverse` family of packages.
If you've already installed the entire tidyverser set of packages
with `install.packages("tidyverse")` you could then call `library(tidyverse)`
to "load" the entire tidyverse into your R session.
</em>


## stringr: Basic String Manipulation

```{r}
chr_data <- c("Data", "Daft", "YouTube", "channel",
             "learn", "and", "have", "FUN!")
```

```{r}
# Check the length of a string
str_length("Hi there! How are you?")
str_length(chr_data)
```

```{r}
# Convert string letters to uppercase
str_to_upper(chr_data)
```

```{r}
# Convert string letters to lowercase
str_to_lower(chr_data)
```

```{r}
# Convert string to title (first letter uppercase)
str_to_title(chr_data)
```

```{r}
# Convert string to sentence (only first letter of first word uppercase)
str_to_sentence("make me into a SENTENCE!")
```

```{r}
# Trim whitespace
str_trim("  Trim Me!   ")
```

```{r}
# Pad strings with whitespace
str_pad("Pad Me!", width = 15, side="both")
```

```{r}
# Truncate strings to a given length
str_trunc("If you have a long string, you might want to truncate it!", 
          width = 50)
```

## stringr: Split and Join Strings

```{r}
# Split strings
str_split("Split Me!", pattern = " ")

food <- c(
  "apples and oranges and pears and bananas",
  "pineapples and mangos and guavas"
)

stringr::str_split(food, " and ")
```

```{r}
# Join strings (equivalent to base R paste())
str_c("Join", "Me!", sep="_")

# Join strings (equivalent to base R paste())
str_c(c("Join", "vectors"), c("Me!", "too!"), sep="_")
```

```{r}
# Collapse a vector of strings into a single string
str_c(c("Turn", "me", "into", "one", "string!"), collapse= " ")
```

```{r}
# Convert NA values in character vector to string "NA"
str_replace_na(c("Make", NA, "strings!"))
```

## stringr: Sorting Strings

```{r}
sort_data <- c("sort", "me", "please!")

# Get vector of indicies that would sort a string alphabetically
str_order(sort_data)
```

```{r}
# Use discovered ordering to extract data in sorted order
sort_data[str_order(sort_data)]
```

```{r}
# Directly extract sorted strings
str_sort(sort_data)
```

```{r}
# Extract in reverse sorted order
str_sort(sort_data, decreasing = TRUE)
```

## stringr: String Interpolation

```{r}
first <- c("Luke", "Han", "Jean-Luc")
last <- c("Skywalker", "Solo", "Picard")

# Interpolate (insert variable values) into strings with str_glue()
str_glue("My name is {first}. {first} {last}.")
```

```{r}
minimum_age <- 18
over_minimum <- c(5, 17, 33)

# Interpolate the result of an execution into a string
str_glue("{first} {last} is {minimum_age + over_minimum} years old.")
```

```{r}
num <- c(1:5)

# Interpolate the result of function calls
str_glue("The square root of {num} is {round(sqrt(num), 3)}.")
```

```{r}
fuel_efficiency <- 30

# Interpolate strings using data from a data frame
mtcars %>% rownames_to_column("Model") %>%
         filter(mpg > fuel_efficiency) %>%
         str_glue_data("The {Model} gets {mpg} mpg.")
```

## stringr: String Matching

```{r}
head(data,8)
```

```{r}
# Detecting the presence of a pattern in strings
str_detect(data$body[1:100], pattern="deep")
```

```{r}
# Get the indicies of matched strings
str_inds <- str_which(data$body[1:100], pattern="deep")
str_inds
```

```{r}
# Extract matched strings using detected indicies
data$body[str_inds]
```

```{r}
# Count the number of matches
str_count(data$body[1:100], "deep")
```

```{r}
# Get the position of matches
str_locate_all(data$body[1], "deep")
```

```{r}
# Get a list of the first match found in each string as a vector
str_extract(data$body[1:3], "deep|the|and")
```

```{r}
# Get a list of the first match found in each string as matrix
str_match(data$body[1:3], "deep|the|and")
```

```{r}
# Get a list of the all matches found in each string as list of matricies
str_match_all(data$body[1:3], "deep|the|and")
```

## stringr: Subset and Replace Strings

```{r}
head(data,8)
```

```{r}
# Get a string subset based on character position
str_sub(data$body[1], start=1, end=100)
```

```{r}
# Get a string subset based on words
word(data$body[1], start=1, end=10)
```

```{r}
# Get the strings that contain a certain pattern
str_subset(data$body[1:100], pattern="deep")
```

```{r}
# Replace a substring with a new string by substring position
str_sub(data$body[1], start=1, end=100) <- str_to_upper(str_sub(data$body[1], 
                                                                start=1, 
                                                                end=100))
str_sub(data$body[1], start=1, end=100)
```

```{r}
# Replace first occurrence of a substring with a new string by matching
str_replace(data$body[1], pattern="deep|DEEP", replacement="multi-layer")
```

```{r}
# Replace all occurrences of a substring with a new string by matching
str_replace_all(data$body[1], pattern="deep|DEEP", replacement="multi-layer")
```


## stringr: Viewing Strings

```{r}
# Basic printing
print(data$body[1:10])
```

```{r}
deep_learning_posts <- data$body[str_which(data$body, "deep learning")]

# View strings in HTML format with the first occurence of a pattern highlighted
str_view(deep_learning_posts, pattern="deep")
```

```{r}
# View strings in HTML format with the first all occurences highlighted
str_view_all(deep_learning_posts, pattern="deep")
```

```{r}
# Format strings into paragraphs of a given width with str_wrap()
wrapped <- str_wrap(data$body[str_which(data$body, "deep learning")][1], 
                    width = 50)
wrapped 
```

```{r}
# Print wrapped string with output obeying newlines
wrapped %>% cat()
```

```{r}
# Display wrapped paragraph as HTML, inserting paragraph breaks
str_wrap(data$body[str_which(data$body, "deep learning")][1], width = 50) %>%
str_replace_all("\n", "<br>") %>%
str_view_all(pattern = "deep")
```

