# <yrChapterNumber>25. </yrChapterNumber> "Cleaning" data

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

Often data comes to us in a "messy" format. One of the "less glamorous" but
very important jobs of anyone
working with data is to know how to "clean up" "messy" data. In other words
to put the data into a common format. 

For example, suppose that employees in a company submit expense reports
every month for travel and other business expenses. It would be best if the
data entry system (e.g. a website) enforced a standard format for the data.
However, this doesn't always happen. Sometimes the system tries to enforce
a specific format but people "misuse" the system leading to 
inconsistencies between different entries. Sometimes data from different
data entry systems are combined into a single data set.

The data below represents expenses from various employees. There are many
inconsistencies between different entries. For example, 
some names are in "first last" format and some names are in "last, first" format. 
There are many other 
formatting inconsistencies that are listed below.

```{r}
library(tibble)
expenses = tribble(
  ~employee,           ~date,          ~amount,          ~category,         ~comments,
  "Sue Smith",         "1/19/2023",    "59.99",         "food",            "lunch with client",
  "Schwartz, Joe",     "01/19/2023",   "$27.00",        "office supplies", "paper for printer",
  "mike david harris", "2023-01-19",   "25",            "Office Stuff",    NA,
  "Dr. A. Davis",      "19/1/2023",    "five hundred",  "FOOD",            "NA",
  "Dr Jones",          "1/19/23",      "1,234.56",      "office suppl.",   "chairs",
  "S. Jones Jr",       "19/1/23",      "1000",          "Office supplies", "desk",
  "Conway, Ella Sr.",  "Jan 19, 2023", "$35.23",        "LUNCH",           "---",
  "Pat Williams",      "2/20/2023",    "150",           "travel",          "train ticket",
  "Chen, Wei",         "20/02/2023",   "$89.50",        "Travel",          "hotel",
  "jennifer ann martinez", "2023-02-21", "42",          "FOOD",            NA,
  "Ms. B. Kim",        "21/2/2023",    "two hundred",  "office supplies", "stapler",
  "J. Thompson",      "2/21/23",      "2,500.00",      "TRAVEL",          "flight",
  "R. Patel Ph.D",     "21/2/23",      "75",            "office suppl.",   "notebooks",
  "Garcia, Maria Jr.", "Feb 22, 2023", "$120.00",       "food",            "team dinner",
  "Dr. Evans",         "3/01/2023",    "95.99",         "Office Stuff",    "supplies",
  "Thompson III",      "3/05/2023",    "200",           "travel",          "conference",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    NA,
  "mike green jr.",  "2023/02/15",     "500",           "OFFICE SUPPLIES",    NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",       "03/22/2023",   "$156.00",      "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "Office Stuff",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",        "3/14/24",      "500",         "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",
)

as.data.frame(expenses)
```

The following are many of the inconsistencies

- Inconsistencies with the name entries
  * some names are in "first last" format and some names are in "last, first" format. 
  * some names are capitalized, some aren't
  * some names have middle include middle names, some don't
  * some names include titles, some don't
  * some names include a title, e.g. Dr. , while some don't
  * some of the titles contain a period after the title, some don't
  * some names include a suffix (e.g. Jr), some don't

- Inconsistencies with dates
  * some dates are in mm/dd/yyyy format (USA style)
  * some dates are in dd/mm/yyyy format (European style)
  * some dates are in yyyy-mm-dd format (standard for many computer systems)
  * some dates use 4 digit years, some use 2 digit years
  * some dates have a leading zero for single digit days or months

- There are additional inconsistencies with the amount, category and comments
  columns
  
# What you need to do 

Write a function named cleanUp that takes the dataframe as 
an argument and returns a cleaned dataframe. Do not lose any data.

Split the `employee` column into the following fields (and remove the 
original single `employee` column):

- **employeeFullName** — a paste of all the name components below (title, first, middle, last, suffix)
- **employeeTitle** — e.g. Dr, Ms (no period)
- **employeeFirst** — first name or initial
- **employeeMiddle** — middle name or initial (NA if none)
- **employeeLast** — last name or initial
- **employeeSuffix** — e.g. Sr, Jr (no period; NA if none)

Leave out periods from all name parts (e.g. `J.` becomes `J`, `Dr.` becomes `Dr`).  
All other columns (date, amount, category, comments) should be cleaned to a 
consistent structure as before.

<style>
/* Scrollable box for wide table output: horizontal slider, no wrapping */
.cleanup-table-scroll {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: visible;
  border: 1px solid #dee2e6;
  border-radius: 4px;
  padding: 0.5rem;
  margin: 0.5rem 0;
}
.cleanup-table-scroll pre {
  white-space: pre;
  margin: 0;
}
</style>

::: {.callout-note collapse="true"}
## click here to see one solution

The solution below uses base R (grep, grepl, sub, gsub, strsplit, toupper, tolower, paste, paste0, nchar) where possible; any stringr functions are called with the `stringr::` prefix.

```{r}
#| label: cleanUp-solution
#| class-output: cleanup-table-scroll

# =============================================================================
# STRING HELPERS
# =============================================================================

# strip_periods(text)
#   Arguments:  text  = character vector (e.g. "Dr.", "J. Smith")
#   Returns:    character vector of same length with periods removed and
#               leading/trailing spaces trimmed.
#   Example:    strip_periods("  J. ")  ->  "J"
strip_periods <- function(text) {
  gsub("\\.", "", trimws(text))
}

# to_sentence_case(text)
#   Arguments:  text = single string (e.g. "mike david", "O'BRIEN")
#   Returns:    string with first letter of each word uppercase, rest lowercase (Sentence case).
#               Returns NA unchanged if text is NA; empty string returned as-is.
#   Example:    to_sentence_case("mike david harris")  ->  "Mike David Harris"
to_sentence_case <- function(text) {
  if (is.na(text) || nchar(trimws(text)) == 0) return(text)
  words <- strsplit(trimws(text), "\\s+")[[1]]
  capped <- paste0(toupper(substring(words, 1, 1)), tolower(substring(words, 2)))
  paste(capped, collapse = " ")
}

# =============================================================================
# NAME PARSING — split into small functions by format
# =============================================================================

# has_suffix(namePart)
#   Arguments:  namePart = single string, no leading/trailing title (e.g. "Jones Jr")
#   Returns:    TRUE if name ends with a suffix like Jr, Sr, III, II, IV.
#   Example:    has_suffix("Jones Jr")  ->  TRUE
# Suffix pattern: one of Jr/Sr/III/II/IV at end. Require at least one space before suffix
# so that "Thompson III" correctly captures "III" (R's regex can otherwise match "II" inside "III").
suffix_pattern <- "(Jr\\.?|Sr\\.?|III|II|IV)"
suffix_at_end_regex <- paste0("\\s+", suffix_pattern, "\\s*$")
has_suffix <- function(namePart) {
  grepl(suffix_at_end_regex, namePart, ignore.case = TRUE)
}

# extract_suffix(namePart)
#   Arguments:  namePart = string that may end with a suffix
#   Returns:    list of two: suffix (string or NA), and nameWithSuffixRemoved (string).
#   Example:    extract_suffix("Conway, Ella Sr.")  ->  list(suffix = "Sr", nameWithSuffixRemoved = "Conway, Ella")
#   Uses two groups (name)(suffix) so the suffix is captured correctly; space before suffix required.
extract_suffix <- function(namePart) {
  if (!has_suffix(namePart)) {
    return(list(suffix = NA, nameWithSuffixRemoved = trimws(namePart)))
  }
  # Two groups: everything before the suffix, and the suffix. Requiring \s+ before suffix fixes III vs II.
  nameWithoutSuffix <- sub(paste0("^(.*)\\s+", suffix_pattern, "\\s*$"), "\\1", namePart, ignore.case = TRUE)
  suffix <- sub(paste0("^.*\\s+", suffix_pattern, "\\s*$"), "\\1", namePart, ignore.case = TRUE)
  list(suffix = strip_periods(suffix), nameWithSuffixRemoved = trimws(nameWithoutSuffix))
}

# normalize_suffix(suffix)
#   Arguments:  suffix = string (e.g. "jr", "SR", "iii") or NA
#   Returns:    correctly cased suffix: "Jr", "Sr", "II", "III", "IV", or NA if not recognized.
#   Example:    normalize_suffix("jr")  ->  "Jr"
normalize_suffix <- function(suffix) {
  if (is.na(suffix) || trimws(suffix) == "") return(NA)
  s <- tolower(trimws(suffix))
  if (s %in% c("jr", "jr.")) return("Jr")
  if (s %in% c("sr", "sr.")) return("Sr")
  if (s == "ii") return("II")
  if (s == "iii") return("III")
  if (s == "iv") return("IV")
  suffix
}

# has_trailing_title(namePart)
#   Arguments:  namePart = string that might end with "Ph.D" or "PhD"
#   Returns:    TRUE if it ends with that.
trailing_phd_regex <- paste0(
  "\\s+",           # space before Ph.D
  "Ph\\.?D\\.?",    # Ph.D or PhD (periods optional)
  "\\s*$"           # optional spaces at end
)
has_trailing_title <- function(namePart) {
  grepl(trailing_phd_regex, namePart, ignore.case = TRUE)
}

# extract_trailing_phd(namePart)
#   Returns:  list(titlePart = "PhD" or NA, nameWithTitleRemoved = string).
extract_trailing_phd <- function(namePart) {
  if (!has_trailing_title(namePart)) {
    return(list(titlePart = NA, nameWithTitleRemoved = trimws(namePart)))
  }
  tit <- sub(paste0(".*", trailing_phd_regex), "\\1", namePart, ignore.case = TRUE)
  nameWithout <- sub(trailing_phd_regex, "", namePart, ignore.case = TRUE)
  list(titlePart = strip_periods(tit), nameWithTitleRemoved = trimws(nameWithout))
}

# has_leading_title(namePart)
#   TRUE if name starts with Dr, Mr, Ms, Mrs (with optional period).
leading_title_regex <- paste0(
  "^(Dr\\.?|Mr\\.?|Ms\\.?|Mrs\\.?)",  # title at start
  "\\s+"                               # at least one space after
)
has_leading_title <- function(namePart) {
  grepl(leading_title_regex, namePart, ignore.case = TRUE)
}

# extract_leading_title(namePart)
#   Returns:  list(titlePart = "Dr" or NA, nameWithTitleRemoved = string).
extract_leading_title <- function(namePart) {
  if (!has_leading_title(namePart)) {
    return(list(titlePart = NA, nameWithTitleRemoved = trimws(namePart)))
  }
  tit <- sub(paste0(leading_title_regex, ".*"), "\\1", namePart, ignore.case = TRUE)
  nameWithout <- sub(leading_title_regex, "", namePart, ignore.case = TRUE)
  list(titlePart = strip_periods(tit), nameWithTitleRemoved = trimws(nameWithout))
}

# parse_name_last_comma_first(namePart)
#   For names like "Last, First" or "Last, First Middle".
#   Arguments:  namePart = string containing exactly one comma
#   Returns:    list(first =, middle =, last =). Periods stripped from each.
parse_name_last_comma_first <- function(namePart) {
  parts <- strsplit(namePart, ",\\s*")[[1]]
  last <- strip_periods(trimws(parts[1]))
  firstAndMiddle <- strsplit(trimws(parts[2]), "\\s+")[[1]]
  first <- strip_periods(firstAndMiddle[1])
  middle <- if (length(firstAndMiddle) >= 2) strip_periods(firstAndMiddle[2]) else NA
  list(first = first, middle = middle, last = last)
}

# parse_name_first_last(namePart)
#   For names like "First Last", "First Middle Last", or just "Last".
#   Arguments:  namePart = string with no comma
#   Returns:    list(first =, middle =, last =).
parse_name_first_last <- function(namePart) {
  words <- strsplit(trimws(namePart), "\\s+")[[1]]
  if (length(words) == 1) {
    list(first = NA, middle = NA, last = strip_periods(words[1]))
  } else if (length(words) == 2) {
    list(first = strip_periods(words[1]), middle = NA, last = strip_periods(words[2]))
  } else {
    first <- strip_periods(words[1])
    middle <- strip_periods(words[2])
    last <- paste(strip_periods(words[3:length(words)]), collapse = " ")
    list(first = first, middle = middle, last = last)
  }
}

# parse_one_employee(employeeName)
#   Arguments:  employeeName = one raw employee string (e.g. "Dr. A. Davis", "Schwartz, Joe")
#   Returns:    list with employeeTitle, employeeFirst, employeeMiddle, employeeLast,
#               employeeSuffix, employeeFullName. All name parts without periods.
#   Example:    parse_one_employee("Conway, Ella Sr.")  ->  first=Ella, last=Conway, suffix=Sr, title=NA
parse_one_employee <- function(employeeName) {
  raw <- trimws(employeeName)
  outSuffix <- extract_suffix(raw)
  raw <- outSuffix$nameWithSuffixRemoved
  outPhd <- extract_trailing_phd(raw)
  title <- outPhd$titlePart
  raw <- outPhd$nameWithTitleRemoved
  outLead <- extract_leading_title(raw)
  if (!is.na(outLead$titlePart)) {
    title <- if (is.na(title) || title == "") outLead$titlePart else paste0(outLead$titlePart, " ", title)
  }
  raw <- outLead$nameWithTitleRemoved
  if (is.na(title) || title == "") title <- NA

  if (length(strsplit(raw, ",\\s*")[[1]]) >= 2) {
    parsed <- parse_name_last_comma_first(raw)
  } else {
    parsed <- parse_name_first_last(raw)
  }
  first <- to_sentence_case(parsed$first)
  middle <- to_sentence_case(parsed$middle)
  last <- to_sentence_case(parsed$last)
  suffix <- normalize_suffix(outSuffix$suffix)

  fullParts <- c(title, first, middle, last, suffix)
  fullParts <- fullParts[!is.na(fullParts) & fullParts != ""]
  employeeFullName <- paste(fullParts, collapse = " ")

  list(employeeTitle = title, employeeFirst = first, employeeMiddle = middle,
       employeeLast = last, employeeSuffix = suffix, employeeFullName = employeeFullName)
}

# parse_employee_column(employeeNames)
#   Arguments:  employeeNames = character vector of raw employee strings
#   Returns:    data frame with columns employeeFullName, employeeTitle, employeeFirst,
#               employeeMiddle, employeeLast, employeeSuffix.
parse_employee_column <- function(employeeNames) {
  rows <- lapply(employeeNames, parse_one_employee)
  out <- as.data.frame(do.call(rbind, lapply(rows, unlist)), stringsAsFactors = FALSE)
  out <- out[c("employeeFullName", "employeeTitle", "employeeFirst", "employeeMiddle", "employeeLast", "employeeSuffix")]
  out
}

# =============================================================================
# DATE PARSING
# =============================================================================

# is_iso_date(dateStr)
#   Returns TRUE if string is already yyyy-mm-dd (e.g. "2023-01-19").
iso_date_regex <- paste0(
  "^[0-9]{4}",      # 4 digits at start (year)
  "-",
  "[0-9]{1,2}",     # 1 or 2 digits (month)
  "-",
  "[0-9]{1,2}$"     # 1 or 2 digits at end (day)
)
is_iso_date <- function(dateStr) {
  grepl(iso_date_regex, trimws(dateStr))
}

# is_month_name_date(dateStr)
#   Returns TRUE for strings like "Jan 19, 2023" or "January 19, 2023".
month_name_date_regex <- paste0(
  "^[A-Za-z]+",     # month name (letters)
  "\\s+",
  "[0-9]{1,2}",     # day
  ",\\s*",
  "[0-9]{2,4}$"     # 2- or 4-digit year
)
is_month_name_date <- function(dateStr) {
  grepl(month_name_date_regex, trimws(dateStr))
}

# parse_one_date(dateStr)
#   Arguments:  dateStr = one date string in any of the supported formats
#   Returns:    string "yyyy-mm-dd" or NA if unparseable.
#   Example:    parse_one_date("1/19/2023")  ->  "2023-01-19"
parse_one_date <- function(dateStr) {
  s <- trimws(dateStr)
  if (is_iso_date(s)) return(s)
  if (is_month_name_date(s)) {
    d <- as.Date(s, format = "%b %d, %Y")
    if (is.na(d)) d <- as.Date(s, format = "%B %d, %Y")
    return(format(d, "%Y-%m-%d"))
  }
  date_parts <- strsplit(s, "[/-]")[[1]]
  if (length(date_parts) != 3) return(NA)
  n1 <- as.integer(date_parts[1])
  n2 <- as.integer(date_parts[2])
  n3 <- as.integer(date_parts[3])
  len1 <- nchar(date_parts[1])
  len3 <- nchar(date_parts[3])
  if (len1 == 4) {
    y <- n1; m <- n2; d <- n3
  } else if (len3 == 4) {
    y <- n3
    if (n1 > 12) { d <- n1; m <- n2 } else if (n2 > 12) { m <- n1; d <- n2 } else { m <- n1; d <- n2 }
  } else {
    y <- 2000 + (n3 - 2000) %% 100
    if (n1 > 12) { d <- n1; m <- n2 } else if (n2 > 12) { m <- n1; d <- n2 } else { m <- n1; d <- n2 }
  }
  sprintf("%04d-%02d-%02d", y, m, d)
}

# parse_dates(dateStrings)
#   Arguments:  dateStrings = character vector of date strings
#   Returns:    character vector of same length, each element "yyyy-mm-dd" or NA.
parse_dates <- function(dateStrings) {
  sapply(trimws(dateStrings), parse_one_date)
}

# =============================================================================
# AMOUNT, CATEGORY, COMMENTS
# =============================================================================

# word_to_number(word)
#   Arguments:  word = string with a number in words (e.g. "five hundred", "twenty-one",
#               "one thousand two hundred thirty-four"). Hyphens and "and" are allowed.
#   Returns:    numeric value, or NA if the string is not a recognized number phrase.
#   Examples:   word_to_number("five hundred") -> 500; word_to_number("twenty-one") -> 21;
#               word_to_number("one thousand two hundred thirty four") -> 1234.
#   Supports:   zero–nine, ten–nineteen, twenty–ninety, hundred, thousand, million.
word_to_number <- function(word) {
  w <- tolower(trimws(word))
  w <- gsub("-", " ", w)
  w <- gsub("\\band\\b", " ", w)
  words <- strsplit(w, "\\s+")[[1]]
  words <- words[words != ""]
  if (length(words) == 0) return(NA)

  ones <- c(zero = 0, one = 1, two = 2, three = 3, four = 4, five = 5, six = 6, seven = 7, eight = 8, nine = 9)
  teens <- c(ten = 10, eleven = 11, twelve = 12, thirteen = 13, fourteen = 14, fifteen = 15,
             sixteen = 16, seventeen = 17, eighteen = 18, nineteen = 19)
  tens <- c(twenty = 20, thirty = 30, forty = 40, fifty = 50, sixty = 60, seventy = 70, eighty = 80, ninety = 90)

  total <- 0
  current <- 0
  i <- 1
  while (i <= length(words)) {
    x <- words[i]
    if (x %in% names(ones)) {
      current <- current + ones[x]
      i <- i + 1
    } else if (x %in% names(teens)) {
      current <- current + teens[x]
      i <- i + 1
    } else if (x %in% names(tens)) {
      current <- current + tens[x]
      i <- i + 1
    } else if (x == "hundred") {
      if (current == 0) current <- 1
      current <- current * 100
      i <- i + 1
    } else if (x == "thousand") {
      if (current == 0) current <- 1
      total <- total + current * 1000
      current <- 0
      i <- i + 1
    } else if (x == "million") {
      if (current == 0) current <- 1
      total <- total + current * 1000000
      current <- 0
      i <- i + 1
    } else {
      return(NA)
    }
  }
  total + current
}

# clean_amounts(amountStrings)
#   Arguments:  amountStrings = character vector (e.g. "$27.00", "1,234.56", "five hundred")
#   Returns:    numeric vector. Removes $ and commas; converts word amounts when possible.
clean_amounts <- function(amountStrings) {
  x <- trimws(amountStrings)
  result <- numeric(length(x))
  for (i in seq_along(x)) {
    s <- gsub("[$,]", "", x[i])
    s <- trimws(s)
    if (grepl("^[0-9.]+$", s)) {
      result[i] <- as.numeric(s)
    } else {
      result[i] <- word_to_number(x[i])
    }
  }
  result
}

# clean_categories(categoryStrings)
#   Arguments:  categoryStrings = character vector (e.g. "FOOD", "office suppl.")
#   Returns:    character vector, lowercase, with abbreviations normalized
#               (e.g. "office suppl." and "Office Stuff" -> "office supplies", "LUNCH" -> "food").
office_suppl_regex <- paste0(
  "\\boffice\\s+suppl\\.?\\s*$"   # "office suppl" or "office suppl." at end of string
)
office_stuff_regex <- "\\boffice\\s+stuff\\s*$"
lunch_regex <- "\\blunch\\s*$"
clean_categories <- function(categoryStrings) {
  x <- tolower(trimws(categoryStrings))
  x <- sub(office_suppl_regex, "office supplies", x)
  x <- sub(office_stuff_regex, "office supplies", x)
  x <- sub(lunch_regex, "food", x)
  x
}

# clean_comments(commentStrings)
#   Arguments:  commentStrings = character vector; can contain placeholders like "---", "NA"
#   Returns:    character vector; placeholders and empty strings replaced with NA.
placeholder_regex <- "^(---|NA|n/a)$"
clean_comments <- function(commentStrings) {
  x <- trimws(as.character(commentStrings))
  x[grepl(placeholder_regex, x, ignore.case = TRUE)] <- NA
  x[x == ""] <- NA
  x
}

# =============================================================================
# MAIN CLEANUP
# =============================================================================

# cleanUp(expensesDF)
#   Arguments:  expensesDF = data frame with columns employee, date, amount, category, comments
#   Returns:    data frame with employee replaced by employeeFullName, employeeTitle,
#               employeeFirst, employeeMiddle, employeeLast, employeeSuffix; and with
#               date, amount, category, comments cleaned to consistent formats.
#   Example:    cleanUp(expenses)  ->  wide table with cleaned columns
cleanUp <- function(expensesDF) {
  df <- as.data.frame(expensesDF, stringsAsFactors = FALSE)
  emp <- parse_employee_column(df$employee)
  df$employee <- NULL
  df <- cbind(emp, df)
  df$date     <- parse_dates(df$date)
  df$amount   <- clean_amounts(df$amount)
  df$category <- clean_categories(df$category)
  df$comments <- clean_comments(df$comments)
  df
}

cleanUp(expenses)
```

:::
