# <yrChapterNumber>25. </yrChapterNumber> "Cleaning" data

Often data comes to us in a "messy" format. One of the "less glamorous" but
very important jobs of anyone
working with data is to know how to "clean up" "messy" data. In other words
to put the data into a common format. 

For example, suppose that employees in a company submit expense reports
every month for travel and other business expenses. It would be best if the
data entry system (e.g. a website) enforced a standard format for the data.
However, this doesn't always happen. Sometimes the system tries to enforce
a specific format but people "misuse" the system leading to 
inconsistencies between different entries. Sometimes data from different
data entry systems are combined into a single data set.

The data below represents expenses from various employees. There are many
inconsistencies between different entries. For example, 
some names are in "first last" format and some names are in "last, first" format. 
There are many other 
formatting inconsistencies that are listed below.

```{r}
library(tibble)
expenses = tribble(
  ~employee,           ~date,          ~amount,          ~category,         ~comments,
  "Sue Smith",         "1/19/2023",    "59.99",         "food",            "lunch with client",
  "Schwartz, Joe",     "01/19/2023",   "$27.00",        "office supplies", "paper for printer",
  "mike david harris", "2023-01-19",   "25",            "Office Stuff",    NA,
  "Dr. A. Davis",      "19/1/2023",    "five hundred",  "FOOD",            "NA",
  "Dr Jones",          "1/19/23",      "1,234.56",      "office suppl.",   "chairs",
  "S. Jones Jr",       "19/1/23",      "1000",          "Office supplies", "desk",
  "Conway, Ella Sr.",  "Jan 19, 2023", "$35.23",        "LUNCH",           "---",
  "Pat Williams",      "2/20/2023",    "150",           "travel",          "train ticket",
  "Chen, Wei",         "20/02/2023",   "$89.50",        "Travel",          "hotel",
  "jennifer ann martinez", "2023-02-21", "42",          "FOOD",            NA,
  "Ms. B. Kim",        "21/2/2023",    "two hundred",  "office supplies", "stapler",
  "J. Thompson",      "2/21/23",      "2,500.00",      "TRAVEL",          "flight",
  "R. Patel Ph.D",     "21/2/23",      "75",            "office suppl.",   "notebooks",
  "Garcia, Maria Jr.", "Feb 22, 2023", "$120.00",       "food",            "team dinner",
  "Dr. Evans",         "3/01/2023",    "95.99",         "Office Stuff",    "supplies",
)

as.data.frame(expenses)
```

The following are many of the inconsistencies

- Inconsistencies with the name entries
  * some names are in "first last" format and some names are in "last, first" format. 
  * some names are capitalized, some aren't
  * some names have middle include middle names, some don't
  * some names include titles, some don't
  * some names include a title, e.g. Dr. , while some don't
  * some of the titles contain a period after the title, some don't
  * some names include a suffix (e.g. Jr), some don't

- Inconsistencies with dates
  * some dates are in mm/dd/yyyy format (USA style)
  * some dates are in dd/mm/yyyy format (European style)
  * some dates are in yyyy-mm-dd format (standard for many computer systems)
  * some dates use 4 digit years, some use 2 digit years
  * some dates have a leading zero for single digit days or months

- There are additional inconsistencies with the amount, category and comments
  columns
  
# What you need to do 

Write a function named cleanUp that takes the dataframe as 
an argument and returns a cleaned dataframe. Do not lose any data.

Split the `employee` column into the following fields (and remove the 
original single `employee` column):

- **employeeFullName** — a paste of all the name components below (title, first, middle, last, suffix)
- **employeeTitle** — e.g. Dr, Ms (no period)
- **employeeFirst** — first name or initial
- **employeeMiddle** — middle name or initial (NA if none)
- **employeeLast** — last name or initial
- **employeeSuffix** — e.g. Sr, Jr (no period; NA if none)

Leave out periods from all name parts (e.g. `J.` becomes `J`, `Dr.` becomes `Dr`).  
All other columns (date, amount, category, comments) should be cleaned to a 
consistent structure as before.

<style>
/* Scrollable box for wide table output: horizontal slider, no wrapping */
.cleanup-table-scroll {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: visible;
  border: 1px solid #dee2e6;
  border-radius: 4px;
  padding: 0.5rem;
  margin: 0.5rem 0;
}
.cleanup-table-scroll pre {
  white-space: pre;
  margin: 0;
}
</style>

::: {.callout-note collapse="true"}
## click here to see one solution

The solution below uses base R (grep, grepl, sub, gsub, strsplit, toupper, tolower, paste, paste0, nchar) where possible; any stringr functions are called with the `stringr::` prefix.

```{r}
#| label: cleanUp-solution
#| class-output: cleanup-table-scroll

# --- Helper: remove periods and trim; returns character of same length ---

# gsub(pattern, replacement, x): returns char vector with all matches replaced
strip_periods <- function(x) gsub("\\.", "", trimws(x))

# --- Helper: parse one employee string into title, first, middle, last, suffix ---

# Returns list with employeeTitle, employeeFirst, employeeMiddle, employeeLast, employeeSuffix, employeeFullName.

# Names are stored without periods (e.g. J not J.).
parse_one_employee <- function(s) {
  s <- trimws(s)
  title <- NA_character_
  suffix <- NA_character_

  # Extract suffix from end (Jr, Sr, III, II, IV)
  if (grepl("\\s+(Jr\\.?|Sr\\.?|III|II|IV)\\s*$", s, ignore.case = TRUE)) {

    # sub(pattern, replacement, x): replaces first match; \\1 is first capture group
    suffix <- sub(".*\\s+(Jr\\.?|Sr\\.?|III|II|IV)\\s*$", "\\1", s, ignore.case = TRUE)
    suffix <- strip_periods(suffix)
    s <- sub("\\s+(Jr\\.?|Sr\\.?|III|II|IV)\\s*$", "", s, ignore.case = TRUE)
    s <- trimws(s)
  }

  # Extract trailing title (e.g. Ph.D)
  if (grepl("\\s+Ph\\.?D\\.?\\s*$", s, ignore.case = TRUE)) {
    if (!is.na(title) && title != "") title <- paste0(title, " ") else title <- ""
    title <- paste0(title, sub(".*\\s+(Ph\\.?D\\.?)\\s*$", "\\1", s, ignore.case = TRUE))
    title <- strip_periods(title)
    s <- sub("\\s+Ph\\.?D\\.?\\s*$", "", s, ignore.case = TRUE)
    s <- trimws(s)
  }

  # Extract leading title (Dr, Mr, Ms, Mrs)
  if (grepl("^(Dr\\.?|Mr\\.?|Ms\\.?|Mrs\\.?)\\s+", s, ignore.case = TRUE)) {
    tit <- sub("^(Dr\\.?|Mr\\.?|Ms\\.?|Mrs\\.?)\\s+.*", "\\1", s, ignore.case = TRUE)
    tit <- strip_periods(tit)
    title <- if (is.na(title) || title == "") tit else paste0(tit, " ", title)
    s <- sub("^(Dr\\.?|Mr\\.?|Ms\\.?|Mrs\\.?)\\s+", "", s, ignore.case = TRUE)
    s <- trimws(s)
  }
  if (is.na(title) || title == "") title <- NA_character_

  # Parse name part: "Last, First" or "First Middle Last"

  # strsplit(x, split): returns list of char vectors
  parts <- strsplit(s, ",\\s*")[[1]]
  if (length(parts) >= 2) {
    last <- strip_periods(trimws(parts[1]))
    first_mid <- strsplit(trimws(parts[2]), "\\s+")[[1]]
    first <- if (length(first_mid) >= 1) strip_periods(first_mid[1]) else NA_character_
    middle <- if (length(first_mid) >= 2) strip_periods(first_mid[2]) else NA_character_
  } else {
    words <- strsplit(s, "\\s+")[[1]]
    if (length(words) == 1) {
      first <- NA_character_; middle <- NA_character_; last <- strip_periods(words[1])
    } else if (length(words) == 2) {
      first <- strip_periods(words[1]); middle <- NA_character_; last <- strip_periods(words[2])
    } else if (length(words) >= 3) {
      first <- strip_periods(words[1]); middle <- strip_periods(words[2])
      last <- paste(strip_periods(words[3:length(words)]), collapse = " ")
    } else {
      first <- NA_character_; middle <- NA_character_; last <- NA_character_
    }
  }

  # Build full name (paste non-NA components)
  full_parts <- c(title, first, middle, last, suffix)
  full_parts <- full_parts[!is.na(full_parts) & full_parts != ""]

  # paste(..., collapse): concatenates into one string; returns character
  employeeFullName <- paste(full_parts, collapse = " ")
  list(employeeTitle = title, employeeFirst = first, employeeMiddle = middle,
       employeeLast = last, employeeSuffix = suffix, employeeFullName = employeeFullName)
}

# --- Parse employee column into 6 columns (no data lost) ---

# lapply(X, FUN): applies FUN to each element; returns list. do.call(rbind, ...): binds rows.
parse_employee_column <- function(x) {
  rows <- lapply(x, parse_one_employee)

  # unlist(r) keeps names; rbind gives matrix with correct colnames
  out <- as.data.frame(do.call(rbind, lapply(rows, unlist)), stringsAsFactors = FALSE)

  # Reorder so employeeFullName is first (paste of the rest)
  out <- out[c("employeeFullName", "employeeTitle", "employeeFirst", "employeeMiddle", "employeeLast", "employeeSuffix")]
  out
}

# --- Helper: parse date string to "yyyy-mm-dd" ---

# Handles m/d/y, d/m/y, yyyy-mm-dd, "Mon dd, yyyy". Returns NA if unparseable.
parse_date_flex <- function(x) {
  x <- trimws(x)
  out <- character(length(x))
  for (i in seq_along(x)) {
    s <- x[i]

    # Already ISO
    if (grepl("^[0-9]{4}-[0-9]{1,2}-[0-9]{1,2}$", s)) {
      out[i] <- s
      next
    }

    # Month name: "Jan 19, 2023"

    # grepl(pattern, x): returns logical; TRUE where pattern matches
    if (grepl("^[A-Za-z]+\\s+[0-9]{1,2},\\s*[0-9]{2,4}$", s)) {

      # as.Date(character, format): returns Date; %b = abbreviated month
      d <- as.Date(s, format = "%b %d, %Y")
      if (is.na(d)) d <- as.Date(s, format = "%B %d, %Y")

      # format(x, format): formats (e.g. Date) as character; returns character vector
      out[i] <- format(d, "%Y-%m-%d")
      next
    }

    # Numeric: split on / or - into day, month, year components

    # strsplit(x, split): returns list of char vectors; [[1]] gets first element
    date_parts <- strsplit(s, "[/-]")[[1]]
    if (length(date_parts) != 3) { out[i] <- NA_character_; next }
    b1 <- as.integer(date_parts[1]); b2 <- as.integer(date_parts[2]); b3 <- as.integer(date_parts[3])

    # Year: 4-digit value, or 2-digit in last position (USA/European)
    if (nchar(date_parts[1]) == 4) {
      y <- b1; m <- b2; d <- b3
    } else if (nchar(date_parts[3]) == 4) {
      y <- b3
      if (b1 > 12) { d <- b1; m <- b2 } else if (b2 > 12) { m <- b1; d <- b2 } else { m <- b1; d <- b2 }
    } else {
      y <- 2000 + (b3 - 2000) %% 100
      if (b1 > 12) { d <- b1; m <- b2 } else if (b2 > 12) { m <- b1; d <- b2 } else { m <- b1; d <- b2 }
    }

    # sprintf(fmt, ...): returns character vector with formatted output
    out[i] <- sprintf("%04d-%02d-%02d", y, m, d)
  }
  out
}

# --- Helper: convert amount string to numeric ---

# Removes $ and commas; handles written "five hundred", "two hundred".
word_to_number <- function(w) {
  w <- tolower(trimws(w))

  # Simple lookup for common words in this data
  if (grepl("^five\\s*hundred", w)) return(500)
  if (grepl("^two\\s*hundred", w)) return(200)
  if (grepl("^one\\s*hundred", w)) return(100)
  NA_real_
}

clean_amount <- function(x) {
  x <- trimws(x)
  out <- numeric(length(x))
  for (i in seq_along(x)) {
    s <- x[i]

    # gsub(pattern, replacement, x): returns char vector, all matches replaced
    s <- gsub("[$,]", "", s)
    s <- trimws(s)
    if (grepl("^[0-9.]+$", s)) {
      out[i] <- as.numeric(s)
    } else {
      out[i] <- word_to_number(x[i])
    }
  }
  out
}

# --- Helper: standardize category (lowercase, fix abbreviations) ---

clean_category <- function(x) {

  # tolower(x): returns character vector in lowercase
  x <- tolower(trimws(x))

  # sub(pattern, replacement, x): replaces first match
  x <- sub("\\boffice\\s+suppl\\.?\\s*$", "office supplies", x)
  x <- sub("\\boffice\\s+stuff\\s*$", "office supplies", x)
  x <- sub("\\blunch\\s*$", "food", x)
  x
}

# --- Helper: standardize comments (NA for empty/placeholder) ---

clean_comments <- function(x) {
  x <- trimws(as.character(x))

  # Replace placeholder strings with NA
  x[grepl("^(---|NA|n/a)$", x, ignore.case = TRUE)] <- NA_character_
  x[x == ""] <- NA_character_
  x
}

# --- Main function ---

# cleanUp(df): expects a dataframe with columns employee, date, amount, category, comments;

# returns a dataframe with employee split into employeeFullName, employeeTitle, employeeFirst,
# employeeMiddle, employeeLast, employeeSuffix (no periods in names), plus cleaned date, amount, category, comments.
cleanUp <- function(df) {
  df <- as.data.frame(df, stringsAsFactors = FALSE)
  emp <- parse_employee_column(df$employee)
  df$employee <- NULL
  df <- cbind(emp, df)
  df$date     <- parse_date_flex(df$date)
  df$amount   <- clean_amount(df$amount)
  df$category <- clean_category(df$category)
  df$comments <- clean_comments(df$comments)
  df
}

# Example: clean the expenses data

cleanUp(expenses)
```

:::
