---
toc: true
toc-depth: 4
format:
  html:
    anchor-sections: false
---
# <yrChapterNumber>24. </yrChapterNumber> Practice - Searching and Cleaning Data (with regex in R)

<style>
/* Style all level-4 headings */
h4 {
  background-color: #f2f2f2;   /* light grey */
  padding: 0.4em 0.6em;
  margin-left: -1rem;          /* extend to container edges */
  margin-right: -1rem;
  padding-left: 1rem;
  padding-right: 1rem;
}
</style>

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

## Introduction - about the data

In this assignment, you'll practice using regular expressions to identify and clean messy data. 
The dataset below contains employee expense records with various inconsistencies in formatting. ***Some*** of the inconsistencies are described below
the data.

### Employee expenses data

::: {.content-hidden}
COMMENT: create the dataframe but don't display it yet
:::

```{r}
#| label: make-data
#| echo: false
#| include: false

library(tibble)

expenses = tribble(
  ~employee,           ~date,          ~amount,          ~category,         ~comments,
  "Sue Smith",         "1/19/2023",    "59.99",         "food",            "lunch with client",
  "  schwartz, joe",   "01/19/2023",   "$27.00",        "office supplies", "paper for printer",
  "mike david harris", "2023-01-19",   "25",            "Office Stuff",    NA,
  "Dr. A. Davis",      "19/1/2023",    "five hundred",  "FOOD",            "NA",
  "Dr Jones",          "1/19/23",      "1,234.56",      "office suppl.",   "chairs",
  "S. Jones Jr",       "19/01/23",     "1000",          "Office supplies", "desk",
  "Conway, Ella Sr.",  "Jan 19, 2023", "$35.23",        "LUNCH",           "---",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    "",
  "mike green jr.",    "2023/02/15",   "500",           "OFFICE SUPPLIES", NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",      "03/22/2023",   "$156.00",       "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "OFFICE STUFF",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",      "3/14/24",      "500",           "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",
  "Pat Williams",      "2/20/2023",    "150",           "travel",          "train ticket",
  "Chen, Wei",         "20/02/2023",   "$89.50",        "Travel",          "hotel",
  "jennifer ann martinez","2023-02-21","42",           "FOOD",            NA,
  "Ms. B. Kim",        "21/2/2023",    "12.50",         "office supplies", "stapler",
  "J. Thompson",       "2/21/23",      "2,500.00",      "TRAVEL",          "flight",
  "R. Patel Ph.D",     "21/2/23",      "75",            "office suppl.",   "notebooks",
  "Garcia, Maria Jr.", "Feb 22, 2023", "$120.00",       "food",            "team dinner",
  "Dr. Evans",         "3/01/2023",    "95.99",         "Office Stuff",    "supplies",
  "Thompson III",      "3/05/2023",    "200",           "travel",          "conference",
  "Brown, Tom",        "3/10/2023",    "65.00",         "food",            "client lunch",
  "  Williams, Kate  ","03/01/2023",   "$45.00",        "office supplies", "paper restock",
  "robert james lee",  "2023/03/01",   "25",            "Office Stuff",    "mouse pad",
  "mike green jr.",    "2023/03/05",   "150",           "OFFICE SUPPLIES", "desk lamp",
  "Dr. B. Wilson",     "20/2/2023",    "85.00",         "FOOD",            "team lunch",
  "Dr Smith",          "3/01/23",      "800.00",        "office suppl.",   "tablet",
  "R. Brown Jr",       "March 10, 2023","300",          "Office supplies", "office chair",
  "Martinez, Ana Sr.", "Feb 20, 2023", "$30.00",        "LUNCH",           "coffee meeting",
  "Johnson,Mary",      "3/25/2023",    "22.00",         "food",            "snacks",
  "Cooper,  Sam",      "03/25/2023",   "$75.00",        "office supplies", "whiteboard markers",
  "peter michael wu",  "2023-03-25",   "35",            "Office Stuff",    "mouse",
  "Dr. C. Taylor",     "3/10/2023",    "120.00",        "FOOD",            "staff lunch",
  "Drake Wilson",      "4/01/24",      "250",           "Office supplies", "scanner",
  "Dr Chen",           "3/10/23",      "900.00",        "office suppl.",   "monitor",
  "T. Wilson Jr",      "Jul 5 2024",   "200",           "Office supplies", "router",
  "Rodriguez, Eva Sr.","Apr 1, 2023",  "$55.00",        "LUNCH",           "team outing",

  "Mike Turchini   ",  "2024-01-31",   "100",           "lunch",           ""
)
expenses = as.data.frame(expenses)
```

```{r}
#| echo: false
#| include: false

# Save a copy of the dataframe.
# If the answer to a question modifies the data, we will 
# restore the data after the question is over.
saveExpenses = expenses
```

::: {.content-hidden}
COMMENT: original broken data
```r
expenses = tribble(
  ~employee,           ~date,          ~amount,          ~category,         ~comments,
  "Sue Smith",         "1/19/2023",    "59.99",         "food",            "lunch with client",
  "  schwartz, joe",     "01/19/2023",   "$27.00",        "office supplies", "paper for printer",
  "mike david harris", "2023-01-19",   "25",            "Office Stuff",    NA,
  "Dr. A. Davis",      "19/1/2023",    "five hundred",  "FOOD",            "NA",
  "Dr Jones",          "1/19/23",      "1,234.56",      "office suppl.",   "chairs",
  "S. Jones Jr",       "19/01/23",     "1000",          "Office supplies", "desk",
  "Conway, Ella Sr.",  "Jan 19, 2023", "$35.23",        "LUNCH",           "---",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    "",
  "mike green jr.",  "2023/02/15",     "500",           "OFFICE SUPPLIES", NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",       "03/22/2023",   "$156.00",      "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "OFFICE STUFF",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",        "3/14/24",      "500",         "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",
    "Pat Williams",      "2/20/2023",    "150",           "travel",          "train ticket",
  "Chen, Wei",         "20/02/2023",   "$89.50",        "Travel",          "hotel",
  "jennifer ann martinez", "2023-02-21", "42",          "FOOD",            NA,
  "Ms. B. Kim",        "21/2/2023",    "two hundred",  "office supplies", "stapler",
  "J. Thompson",      "2/21/23",      "2,500.00",      "TRAVEL",          "flight",
  "R. Patel Ph.D",     "21/2/23",      "75",            "office suppl.",   "notebooks",
  "Garcia, Maria Jr.", "Feb 22, 2023", "$120.00",       "food",            "team dinner",
  "Dr. Evans",         "3/01/2023",    "95.99",         "Office Stuff",    "supplies",
  "Thompson III",      "3/05/2023",    "200",           "travel",          "conference",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    NA,
  "mike green jr.",  "2023/02/15",     "500",           "OFFICE SUPPLIES",    NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",       "03/22/2023",   "$156.00",      "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "Office Stuff",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",        "3/14/24",      "500",         "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",
  "Mike Turchini   ",     "2024-01-31",   "100",           "lunch",           "",
)
expenses = as.data.frame(expenses)
```
:::

::: {.content-hidden}
COMMENT: Display the 1st few rows.
:::

```{r}
#| echo: false

numrows = 10
cat("There are a total of ", nrow(expenses), " rows. The following are ", 
    "just the first ", numrows, " rows.", sep="")
head(expenses, 10)
```

Click on the link below to see the full dataset.


::: {.content-hidden}
COMMENT: show the full dataset/code
:::

::: {.callout-note collapse="true"}
## click here to see the full data set

Copy the following code into R and run it to 
recreate the full dataset.

```{r}
#| ref.label: make-data
#| echo: true
```
:::

- Inconsistencies with the `employee` column entries

  * some names are in "first last" (e.g. "Sue Smith") format and 
    some names are in "last, first" format (e.g. "Brown, Tom")
  * some names are capitalized, some aren't
  * some names have middle include middle names, some don't
  * some names include titles, some don't
  * some names include a title, e.g. Dr. , while some don't
  * some of the titles contain a period after the title, some don't
  * some names include a suffix (e.g. Jr), some don't
  * some names are missing first names
  * some names use periods (e.g. J. Thompson), some don't
  * possibly other inconsistencies ...

  
- Inconsistencies with `date` column

  * some dates are in mm/dd/yyyy format (USA style)
  * some dates are in dd/mm/yyyy format (European style)
  * some dates are in yyyy-mm-dd format (standard for many computer systems)
  * some dates use 4 digit years, some use 2 digit years
  * some dates have a leading zero for single digit days or months
  * possibly other inconsistencies ...

- Inconsistencies in the `amount` column

  * some values have $ signs, some don't, 
  * some have commas (e.g. 1,234.56) some don't
  * some are missing the cents (e.g. "75")
  * some are written in words (e.g. "five hundred")
  * possibly other inconsistencies ...

- Inconsistencies in the `category` column

  * the same categories are written differently (e.g. "food" vs "FOOD", 
    "office supplies" vs "Office suppl." vs "OFFICE STUFF")
  * possibly other inconsistencies ...

- Inconsistencies in the `comments` column

  * blanks are expressed differently: NA vs "NA" vs "---" vs "".
  
    Note that in this column, other differences are more forgivable since 
    comments are by their very nature written by people differently and are 
    not expected to be in a standard form. However, there should be a standard
    way to express that there are no comments for a given row.


## Questions - PART I  (writing code)

### Instructions

Your task is to write R code using regular expressions to identify different patterns in the data.

#### Each question starts with a fresh copy of the data

For each question, assume that you start with a fresh copy of the data.
In other words, if question #5 asks you to change the data, assume that
question #6 starts again with the original data shown above.

#### Focus on regex code, but use other R code if necessary

You may use any of the regex functions we covered in class:

- `grep()`: Find patterns and return matching indices or values
- `grepl()`: Find patterns and return TRUE/FALSE for each element
- `sub()`: Replace the first occurrence of a pattern
- `gsub()`: Replace all occurrences of a pattern
- `strsplit()`: Split strings based on a pattern

In addition you can use other R code to help.

#### Advice on approaching the questions

For each question:

1. First try to solve it yourself
2. If you get stuck, try to break down the pattern you're looking for into smaller pieces
3. Test your solution with the sample data to make sure it catches all cases

#### The answer must work even if the data changes.

You solution MUST work, even if the actual data changes. For example 

QUESTION: Display all entries in the comments column that contain the word "for".

**CORRECT ANSWER**: The following answer uses the `\b` metacharacter to indicate the start and
the end of the word.

```{r}
grep("\\bfor\\b", expenses$comments, value=TRUE)
```

**WRONG ANSWER**: The following is obviously a wrong answer, even though it produces the same results.
The answer refers to a specific row.
The output would be different if the order of the rows changed.
The answer must assume no knowledge of the actual data in the dataframe.

```{r}
expenses[2,]
```

**ANOTHER WRONG ANSWER**: The following is not totally correct either even though the output looks right.
Can you see why?

```{r}
grep("for", expenses$comments, value=TRUE)

# THE ANSWER IS WRONG BECAUSE ...
#
# Suppose someone were to add a new row that contained the comment 
# "printer forms", the code
#
#           grep("for", expenses$comments, value=TRUE) 
#
# simply searches for the 
# letters "for" anywhere in the comment and would incorrectly match
# the "for" in the word "forms". The question specifically said that 
# the comment must contain the word for. Even though none of the words
# "form" or "forms" or  "format" or "before" or "information", etc.
# appear in the data, the question asked for the word "for" and you 
# answer should only pick up the word "for".
#
# Let's try it

newExpense = data.frame(employee = "Joe Smith",
                        date = "1/1/2026",
                        amount = "$100.00",
                        category = "office supplies",
                        comments = "printer forms")
expenses = rbind(expenses, newExpense)

grep("for", expenses$comments, value=TRUE)
```


**THERE COULD BE MORE THAN ONE CORRECT ANSWER**: Suppse you forgot about the `\b` 
metacharacter (which indicates a 
word break), you could have answered the question using the following code:

```{r}
pattern = paste0("^for$",                  # "for" is the only text in the comment
                 "|^for[^A-Za-z]",         # or "for" is at the beginning and followed by a non-letter  
                 "|[^A-Za-z]for$",         # or "for" is at the end and preceded by a non-letter
                 "|[^A-Za-z]for[^A-Za-z]"  # or "for" is in the middle and preceded and followed by non-letters
                 )
grep(pattern, expenses$comments, value=TRUE)
```



```{r}
#| echo: false

# RESET expenses back to the original values
expenses = saveExpenses
```



### Warm up Questions 

#### Question 1a

Write R code that outputs all entries in the employee column where the very 
last character in the entry is a lowercase "i" (Do not show the entry 
if the ending is i followed by a space or a period, etc)

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-wu-1a
#| include: false
#| echo: false
#| output: false

answer = 
 grep("i$", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:

<!-- show the values output by the answer code -->
```{r}
#| echo: false
print(answer)
```

#### Question 1b

Modify your answer to the previous part so that the code outputs
all entries where the last LETTER is lowercase "i" (even if it might be followed
by digits or spaces or periods, etc)

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-wu-1b
#| include: false
#| echo: false
#| output: false

answer = 
 grep("i[^a-zA-Z]*$", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:

<!-- show the values output by the answer code -->
```{r}
#| echo: false
print(answer)
```

::: {.callout-note collapse="true"}
### Click here for answer
```{r}
```
:::

#### Question 2a

Show all names that have whitespace (spaces, tabs, newlines) at the beginning
or end of the entry.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-wu-2a
#| include: false
#| echo: false
#| output: false

answer = 
 grep("^\\s+|\\s+$", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:

<!-- show the values output by the answer code -->
:::
```{r}
#| echo: false
print(answer)
```

#### Question 2b

Use regular expressions to remove all whitespace that appears before or 
after an entry in the 
employee column. You can do this in more than one command if you like.

Prove that your code worked by comparing the names that you found in 
previous question to the new versions from this question.

  
::: {.callout-note collapse="true"}
### Click here for answer
```{r}
###########################
# Some preparation
###########################

# Get the rows that contain leading or trailing whitespace
rowNumbers = grep("^\\s+|\\s+$", expenses$employee)
rowNumbers

# Take a copy of the original data in order to modify it without affecting the original data
newExpenses = expenses

############################
# Modify the new dataframe
############################

# Remove any spaces BEFORE the names
newExpenses$employee = sub("^\\s+", "", expenses$employee)

# Remove any spaces AFTER the names
newExpenses$employee = sub("\\s+$", "", expenses$employee)


###############################################################################
# Compare the rows that were changed in the old dataframe and the new dataframe
###############################################################################

data.frame ( oldEntries = paste0('"' , expenses$employee[rowNumbers] , '"'),
             fixedEntries = paste0('"' , newExpenses$employee[rowNumbers] , '"') )


```
:::

### Name Format Questions

#### Question 1a

Write R code using regex to identify all names that are in "last, first" 
format (e.g., "Schwartz, Joe", including those with suffixes like "Sr." or
"Jr."). Make sure that you allow for any number of spaces between the 
comma and the first name (including if there aren't any spaces).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-1a
#| include: false
#| echo: false
#| output: false

answer = 
 grep("^[a-zA-Z]+,.+$", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:

<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-1a
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 1b

Modify your answer to show only names that are strictly in  
last, first form (without a suffix) 

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-1b
#| include: false
#| echo: false
#| output: false

answer = 
 grep("^[a-zA-Z]+, *[a-zA-Z]+$", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:

<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-1b
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

   
#### Question 2

Write R code using regex to display only names that begin with "Dr" (with or without a period).
Make sure that you don't display names (e.g. "Drake Wilson") that begin with Dr 
but who aren't doctors.


<!-- generate answer but don't show this code -->
```{r}
#| label: answer-2
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("^Dr[. ]", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-2
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::


#### Question 3

Write R code using regex to identify all names that are completely lowercase (no capitalization).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-3
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("^[a-z][a-z. ,]*$", trimws(expenses$employee), value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-3
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 4

Write R code using regex to identify all names that include a middle name or middle initial.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-4
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("( [A-Za-z]\\. |([^ ]+ ){2,}[^ ]+)", expenses$employee, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-4
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 5

Write R code using regex to identify all names that have a suffix (Sr. or Jr., with or without the period).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-5
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("(Jr\\.?|Sr\\.?) *$", trimws(expenses$employee), value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-5
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

### Date Format Questions

#### Question 6

Write R code using regex to identify dates in the MM/DD/YYYY format (with or without leading zeros), ensuring valid months (1-12) and days (1-31).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-6
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("^(0?[1-9]|1[0-2])/(0?[1-9]|[12][0-9]|3[01])/([0-9]{4}|[0-9]{2})$", expenses$date, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-6
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 7

Write R code using regex to identify dates in the European format (DD/MM/YYYY or DD/MM/YY), ensuring valid days and months.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-7
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("^(0?[1-9]|[12][0-9]|3[01])/(0?[1-9]|1[0-2])/([0-9]{4}|[0-9]{2})$", expenses$date, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-7
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 8

Write R code using regex to identify dates in ISO format (YYYY-MM-DD).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-8
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("^[0-9]{4}-(0[1-9]|1[0-2])-(0[1-9]|[12][0-9]|3[01])$", expenses$date, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-8
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 9

Write R code using regex to identify dates that use written month names (e.g., "Jan 19, 2023", "Feb 15, 2023").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-9
#| include: false
#| echo: false
#| output: false
  
answer = 
 grep("(Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec)[a-z]* [0-9]{1,2},? [0-9]{4}", expenses$date, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-9
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

### Amount Format Questions

#### Question 10

Write R code using regex to identify amounts that include a 
dollar sign (e.g., "$27.00", "$156.00").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-10
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("^\\$", expenses$amount, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-10
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 11

Write R code using regex to identify amounts that are 
written out in words (e.g., "five hundred", "one thousand").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-11
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("^[a-z ]+$", expenses$amount, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-11
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 12

Write R code using regex to identify amounts that include 
thousand separators (commas) (e.g., "1,234.56", "2,345.67").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-12
#| include: false
#| echo: false
#| output: false
   
answer = 
  grep(",[0-9]{3}", expenses$amount, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-12
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::


#### Question 13

Write R code using regex to identify amounts that are 
whole numbers without decimal places (e.g., "25", "1000").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-13
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("^[0-9]+$", expenses$amount, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-13
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

### Category Format Questions

#### Question 14

Write R code using regex to identify categories that are 
entirely uppercase (e.g., "FOOD", "LUNCH").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-14
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("^[A-Z ]+$", expenses$category, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-14
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 15

Write R code using regex to identify all variations of 
"office supplies" (including abbreviated forms and different capitalizations).

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-15
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("office (suppl|stuff|supplies)", expenses$category, ignore.case=TRUE, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-15
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

### Comment Format Questions

#### Question 16

Write R code using regex to identify comments that are either 
NA, "NA", or consist entirely of dashes (e.g "---" and "-----") or 
are the empty string, i.e. "".

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-16
#| include: false
#| echo: false
#| output: false
  
tf = is.na(expenses$comments) | 
     grepl("^NA$|^-+$|^$", expenses$comments) 
answer = expenses$comments[tf]
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-16
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 17

Write R code using regex to identify comments that describe 
meetings (containing words like "lunch", "dinner", "client", "team").

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-17
#| include: false
#| echo: false
#| output: false
  
answer = 
  grep("(lunch|dinner|client|team)", expenses$comments, ignore.case=TRUE, value=TRUE)
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-17
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

### Challenge Questions

#### Question 18

Write R code using regex to extract just the last name from any 
name format in the employee column. Your solution should work 
for all name formats in the dataset.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-18
#| include: false
#| echo: false
#| output: false
   
name = trimws(expenses$employee)
# Strip trailing suffixes (Jr, Sr, Ph.D, III, II, IV) so "last word" is the surname (case-insensitive)
suffix_pat = "\\s*(Jr\\.?|Sr\\.?|Ph\\.?D\\.?|III|IV|II)\\s*$"
name_clean = trimws(sub(suffix_pat, "", name, ignore.case = TRUE))
last_name = ifelse(grepl("^[^,]+,", name),
  trimws(sub("^([^,]+),.*", "\\1", name)),
  sub(".*\\s+([^ ]+)$", "\\1", name_clean))
# Remove any suffix from last_name (e.g. "Smith Jr" in comma format)
answer = trimws(sub(suffix_pat, "", last_name, ignore.case = TRUE))
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-18
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 19

Write R code using regex to convert all dates to YYYY-MM-DD format,
regardless of their original format. Your solution should handle all 
date formats in the dataset.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-19
#| include: false
#| echo: false
#| output: false
  
answer = format(
  as.Date(expenses$date,
    tryFormats = c("%Y-%m-%d", "%m/%d/%Y", "%m/%d/%y", "%d/%m/%Y", "%d/%m/%y",
      "%B %d, %Y", "%b %d, %Y", "%B %d %Y", "%b %d %Y", "%Y/%m/%d")),
      "%Y-%m-%d")
```

Given the data above, your output should show the following:
   
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-19
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

#### Question 20

Write R code using regex to standardize all amounts to a decimal number 
format (e.g., "1234.56"), removing dollar signs, commas, and converting 
written amounts to numbers.

<!-- generate answer but don't show this code -->
```{r}
#| label: answer-20
#| include: false
#| echo: false
#| output: false
  
amt = gsub("[$,]", "", expenses$amount)
amt = gsub("one thousand", "1000", amt, ignore.case=TRUE)
amt = gsub("five hundred", "500", amt, ignore.case=TRUE)
amt = gsub("two hundred", "200", amt, ignore.case=TRUE)
answer = format(round(as.numeric(amt), 2), nsmall=2)
```

Given the data above, your output should show the following:
    
<!-- show the values output by the answer code -->
```{r}
#| echo: false
cat(answer, sep = "\n")
```

::: {.callout-note collapse="true"}
### Click here for answer
<!-- show how to generate answer -->
```{r}
#| ref.label: answer-20
#| echo: true
```
<!-- show the output -->
```{r}
cat(answer, sep="\n")
```
:::

## Questions - PART II   (project)

Write a function named cleanup that takes the dataframe as 
an argument and returns a cleaned dataframe. 
Each column should have a consistent format for all columns.
Do not lose any data in the process.

Split the `employee` column into the following fields (and remove the 
original single `employee` column):

- **employeeFullName** — a paste of all the name components below (title, first, middle, last, suffix)
- **employeeTitle** — e.g. Dr, Ms (no period)
- **employeeFirst** — first name or initial
- **employeeMiddle** — middle name or initial (NA if none)
- **employeeLast** — last name or initial
- **employeeSuffix** — e.g. Sr, Jr (no period; NA if none)

Leave out periods from all name parts (e.g. `J.` becomes `J`, `Dr.` becomes `Dr`).  
All other columns (date, amount, category, comments) should be cleaned to a 
consistent structure as before.

<style>
/* Scrollable box for wide table output: horizontal slider, no wrapping */
.cleanup-table-scroll {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: visible;
  border: 1px solid #dee2e6;
  border-radius: 4px;
  padding: 0.5rem;
  margin: 0.5rem 0;
}
.cleanup-table-scroll pre {
  white-space: pre;
  margin: 0;
}
</style>

```{r}
#| label: cleanUp-solution
#| class-output: cleanup-table-scroll
#| echo: false
#| output: false

# =============================================================================
# STRING HELPERS
# =============================================================================

# strip_periods(text)
#   Arguments:  text  = character vector (e.g. "Dr.", "J. Smith")
#   Returns:    character vector of same length with periods removed and
#               leading/trailing spaces trimmed.
#   Example:    strip_periods("  J. ")  ->  "J"
strip_periods <- function(text) {
  gsub("\\.", "", trimws(text))
}

# to_sentence_case(text)
#   Arguments:  text = single string (e.g. "mike david", "O'BRIEN")
#   Returns:    string with first letter of each word uppercase, rest lowercase (Sentence case).
#               Returns NA unchanged if text is NA; empty string returned as-is.
#   Example:    to_sentence_case("mike david harris")  ->  "Mike David Harris"
to_sentence_case <- function(text) {
  if (is.na(text) || nchar(trimws(text)) == 0) return(text)
  words <- strsplit(trimws(text), "\\s+")[[1]]
  capped <- paste0(toupper(substring(words, 1, 1)), tolower(substring(words, 2)))
  paste(capped, collapse = " ")
}

# =============================================================================
# NAME PARSING — split into small functions by format
# =============================================================================

# has_suffix(namePart)
#   Arguments:  namePart = single string, no leading/trailing title (e.g. "Jones Jr")
#   Returns:    TRUE if name ends with a suffix like Jr, Sr, III, II, IV.
#   Example:    has_suffix("Jones Jr")  ->  TRUE
# Suffix pattern: one of Jr/Sr/III/II/IV at end. Require at least one space before suffix
# so that "Thompson III" correctly captures "III" (R's regex can otherwise match "II" inside "III").
suffix_pattern <- "(Jr\\.?|Sr\\.?|III|II|IV)"
suffix_at_end_regex <- paste0("\\s+", suffix_pattern, "\\s*$")
has_suffix <- function(namePart) {
  grepl(suffix_at_end_regex, namePart, ignore.case = TRUE)
}

# extract_suffix(namePart)
#   Arguments:  namePart = string that may end with a suffix
#   Returns:    list of two: suffix (string or NA), and nameWithSuffixRemoved (string).
#   Example:    extract_suffix("Conway, Ella Sr.")  ->  list(suffix = "Sr", nameWithSuffixRemoved = "Conway, Ella")
#   Uses two groups (name)(suffix) so the suffix is captured correctly; space before suffix required.
extract_suffix <- function(namePart) {
  if (!has_suffix(namePart)) {
    return(list(suffix = NA, nameWithSuffixRemoved = trimws(namePart)))
  }
  # Two groups: everything before the suffix, and the suffix. Requiring \s+ before suffix fixes III vs II.
  nameWithoutSuffix <- sub(paste0("^(.*)\\s+", suffix_pattern, "\\s*$"), "\\1", namePart, ignore.case = TRUE)
  suffix <- sub(paste0("^.*\\s+", suffix_pattern, "\\s*$"), "\\1", namePart, ignore.case = TRUE)
  list(suffix = strip_periods(suffix), nameWithSuffixRemoved = trimws(nameWithoutSuffix))
}

# normalize_suffix(suffix)
#   Arguments:  suffix = string (e.g. "jr", "SR", "iii") or NA
#   Returns:    correctly cased suffix: "Jr", "Sr", "II", "III", "IV", or NA if not recognized.
#   Example:    normalize_suffix("jr")  ->  "Jr"
normalize_suffix <- function(suffix) {
  if (is.na(suffix) || trimws(suffix) == "") return(NA)
  s <- tolower(trimws(suffix))
  if (s %in% c("jr", "jr.")) return("Jr")
  if (s %in% c("sr", "sr.")) return("Sr")
  if (s == "ii") return("II")
  if (s == "iii") return("III")
  if (s == "iv") return("IV")
  suffix
}

# has_trailing_title(namePart)
#   Arguments:  namePart = string that might end with "Ph.D" or "PhD"
#   Returns:    TRUE if it ends with that.
trailing_phd_regex <- paste0(
  "\\s+",           # space before Ph.D
  "Ph\\.?D\\.?",    # Ph.D or PhD (periods optional)
  "\\s*$"           # optional spaces at end
)
has_trailing_title <- function(namePart) {
  grepl(trailing_phd_regex, namePart, ignore.case = TRUE)
}

# extract_trailing_phd(namePart)
#   Returns:  list(titlePart = "PhD" or NA, nameWithTitleRemoved = string).
extract_trailing_phd <- function(namePart) {
  if (!has_trailing_title(namePart)) {
    return(list(titlePart = NA, nameWithTitleRemoved = trimws(namePart)))
  }
  tit <- sub(paste0(".*", trailing_phd_regex), "\\1", namePart, ignore.case = TRUE)
  nameWithout <- sub(trailing_phd_regex, "", namePart, ignore.case = TRUE)
  list(titlePart = strip_periods(tit), nameWithTitleRemoved = trimws(nameWithout))
}

# has_leading_title(namePart)
#   TRUE if name starts with Dr, Mr, Ms, Mrs (with optional period).
leading_title_regex <- paste0(
  "^(Dr\\.?|Mr\\.?|Ms\\.?|Mrs\\.?)",  # title at start
  "\\s+"                               # at least one space after
)
has_leading_title <- function(namePart) {
  grepl(leading_title_regex, namePart, ignore.case = TRUE)
}

# extract_leading_title(namePart)
#   Returns:  list(titlePart = "Dr" or NA, nameWithTitleRemoved = string).
extract_leading_title <- function(namePart) {
  if (!has_leading_title(namePart)) {
    return(list(titlePart = NA, nameWithTitleRemoved = trimws(namePart)))
  }
  tit <- sub(paste0(leading_title_regex, ".*"), "\\1", namePart, ignore.case = TRUE)
  nameWithout <- sub(leading_title_regex, "", namePart, ignore.case = TRUE)
  list(titlePart = strip_periods(tit), nameWithTitleRemoved = trimws(nameWithout))
}

# parse_name_last_comma_first(namePart)
#   For names like "Last, First" or "Last, First Middle".
#   Arguments:  namePart = string containing exactly one comma
#   Returns:    list(first =, middle =, last =). Periods stripped from each.
parse_name_last_comma_first <- function(namePart) {
  parts <- strsplit(namePart, ",\\s*")[[1]]
  last <- strip_periods(trimws(parts[1]))
  firstAndMiddle <- strsplit(trimws(parts[2]), "\\s+")[[1]]
  first <- strip_periods(firstAndMiddle[1])
  middle <- if (length(firstAndMiddle) >= 2) strip_periods(firstAndMiddle[2]) else NA
  list(first = first, middle = middle, last = last)
}

# parse_name_first_last(namePart)
#   For names like "First Last", "First Middle Last", or just "Last".
#   Arguments:  namePart = string with no comma
#   Returns:    list(first =, middle =, last =).
parse_name_first_last <- function(namePart) {
  words <- strsplit(trimws(namePart), "\\s+")[[1]]
  if (length(words) == 1) {
    list(first = NA, middle = NA, last = strip_periods(words[1]))
  } else if (length(words) == 2) {
    list(first = strip_periods(words[1]), middle = NA, last = strip_periods(words[2]))
  } else {
    first <- strip_periods(words[1])
    middle <- strip_periods(words[2])
    last <- paste(strip_periods(words[3:length(words)]), collapse = " ")
    list(first = first, middle = middle, last = last)
  }
}

# parse_one_employee(employeeName)
#   Arguments:  employeeName = one raw employee string (e.g. "Dr. A. Davis", "Schwartz, Joe")
#   Returns:    list with employeeTitle, employeeFirst, employeeMiddle, employeeLast,
#               employeeSuffix, employeeFullName. All name parts without periods.
#   Example:    parse_one_employee("Conway, Ella Sr.")  ->  first=Ella, last=Conway, suffix=Sr, title=NA
parse_one_employee <- function(employeeName) {
  raw <- trimws(employeeName)
  outSuffix <- extract_suffix(raw)
  raw <- outSuffix$nameWithSuffixRemoved
  outPhd <- extract_trailing_phd(raw)
  title <- outPhd$titlePart
  raw <- outPhd$nameWithTitleRemoved
  outLead <- extract_leading_title(raw)
  if (!is.na(outLead$titlePart)) {
    title <- if (is.na(title) || title == "") outLead$titlePart else paste0(outLead$titlePart, " ", title)
  }
  raw <- outLead$nameWithTitleRemoved
  if (is.na(title) || title == "") title <- NA

  if (length(strsplit(raw, ",\\s*")[[1]]) >= 2) {
    parsed <- parse_name_last_comma_first(raw)
  } else {
    parsed <- parse_name_first_last(raw)
  }
  first <- to_sentence_case(parsed$first)
  middle <- to_sentence_case(parsed$middle)
  last <- to_sentence_case(parsed$last)
  suffix <- normalize_suffix(outSuffix$suffix)

  fullParts <- c(title, first, middle, last, suffix)
  fullParts <- fullParts[!is.na(fullParts) & fullParts != ""]
  employeeFullName <- paste(fullParts, collapse = " ")

  list(employeeTitle = title, employeeFirst = first, employeeMiddle = middle,
       employeeLast = last, employeeSuffix = suffix, employeeFullName = employeeFullName)
}

# parse_employee_column(employeeNames)
#   Arguments:  employeeNames = character vector of raw employee strings
#   Returns:    data frame with columns employeeFullName, employeeTitle, employeeFirst,
#               employeeMiddle, employeeLast, employeeSuffix.
parse_employee_column <- function(employeeNames) {
  rows <- lapply(employeeNames, parse_one_employee)
  out <- as.data.frame(do.call(rbind, lapply(rows, unlist)), stringsAsFactors = FALSE)
  out <- out[c("employeeFullName", "employeeTitle", "employeeFirst", "employeeMiddle", "employeeLast", "employeeSuffix")]
  out
}

# =============================================================================
# DATE PARSING
# =============================================================================

# is_iso_date(dateStr)
#   Returns TRUE if string is already yyyy-mm-dd (e.g. "2023-01-19").
iso_date_regex <- paste0(
  "^[0-9]{4}",      # 4 digits at start (year)
  "-",
  "[0-9]{1,2}",     # 1 or 2 digits (month)
  "-",
  "[0-9]{1,2}$"     # 1 or 2 digits at end (day)
)
is_iso_date <- function(dateStr) {
  grepl(iso_date_regex, trimws(dateStr))
}

# is_month_name_date(dateStr)
#   Returns TRUE for strings like "Jan 19, 2023" or "January 19, 2023".
month_name_date_regex <- paste0(
  "^[A-Za-z]+",     # month name (letters)
  "\\s+",
  "[0-9]{1,2}",     # day
  ",\\s*",
  "[0-9]{2,4}$"     # 2- or 4-digit year
)
is_month_name_date <- function(dateStr) {
  grepl(month_name_date_regex, trimws(dateStr))
}

# parse_one_date(dateStr)
#   Arguments:  dateStr = one date string in any of the supported formats
#   Returns:    string "yyyy-mm-dd" or NA if unparseable.
#   Example:    parse_one_date("1/19/2023")  ->  "2023-01-19"
parse_one_date <- function(dateStr) {
  s <- trimws(dateStr)
  if (is_iso_date(s)) return(s)
  if (is_month_name_date(s)) {
    d <- as.Date(s, format = "%b %d, %Y")
    if (is.na(d)) d <- as.Date(s, format = "%B %d, %Y")
    return(format(d, "%Y-%m-%d"))
  }
  date_parts <- strsplit(s, "[/-]")[[1]]
  if (length(date_parts) != 3) return(NA)
  n1 <- as.integer(date_parts[1])
  n2 <- as.integer(date_parts[2])
  n3 <- as.integer(date_parts[3])
  len1 <- nchar(date_parts[1])
  len3 <- nchar(date_parts[3])
  if (len1 == 4) {
    y <- n1; m <- n2; d <- n3
  } else if (len3 == 4) {
    y <- n3
    if (n1 > 12) { d <- n1; m <- n2 } else if (n2 > 12) { m <- n1; d <- n2 } else { m <- n1; d <- n2 }
  } else {
    y <- 2000 + (n3 - 2000) %% 100
    if (n1 > 12) { d <- n1; m <- n2 } else if (n2 > 12) { m <- n1; d <- n2 } else { m <- n1; d <- n2 }
  }
  sprintf("%04d-%02d-%02d", y, m, d)
}

# parse_dates(dateStrings)
#   Arguments:  dateStrings = character vector of date strings
#   Returns:    character vector of same length, each element "yyyy-mm-dd" or NA.
parse_dates <- function(dateStrings) {
  sapply(trimws(dateStrings), parse_one_date)
}

# =============================================================================
# AMOUNT, CATEGORY, COMMENTS
# =============================================================================

# word_to_number(word)
#   Arguments:  word = string with a number in words (e.g. "five hundred", "twenty-one",
#               "one thousand two hundred thirty-four"). Hyphens and "and" are allowed.
#   Returns:    numeric value, or NA if the string is not a recognized number phrase.
#   Examples:   word_to_number("five hundred") -> 500; word_to_number("twenty-one") -> 21;
#               word_to_number("one thousand two hundred thirty four") -> 1234.
#   Supports:   zero–nine, ten–nineteen, twenty–ninety, hundred, thousand, million.
word_to_number <- function(word) {
  w <- tolower(trimws(word))
  w <- gsub("-", " ", w)
  w <- gsub("\\band\\b", " ", w)
  words <- strsplit(w, "\\s+")[[1]]
  words <- words[words != ""]
  if (length(words) == 0) return(NA)

  ones <- c(zero = 0, one = 1, two = 2, three = 3, four = 4, five = 5, six = 6, seven = 7, eight = 8, nine = 9)
  teens <- c(ten = 10, eleven = 11, twelve = 12, thirteen = 13, fourteen = 14, fifteen = 15,
             sixteen = 16, seventeen = 17, eighteen = 18, nineteen = 19)
  tens <- c(twenty = 20, thirty = 30, forty = 40, fifty = 50, sixty = 60, seventy = 70, eighty = 80, ninety = 90)

  total <- 0
  current <- 0
  i <- 1
  while (i <= length(words)) {
    x <- words[i]
    if (x %in% names(ones)) {
      current <- current + ones[x]
      i <- i + 1
    } else if (x %in% names(teens)) {
      current <- current + teens[x]
      i <- i + 1
    } else if (x %in% names(tens)) {
      current <- current + tens[x]
      i <- i + 1
    } else if (x == "hundred") {
      if (current == 0) current <- 1
      current <- current * 100
      i <- i + 1
    } else if (x == "thousand") {
      if (current == 0) current <- 1
      total <- total + current * 1000
      current <- 0
      i <- i + 1
    } else if (x == "million") {
      if (current == 0) current <- 1
      total <- total + current * 1000000
      current <- 0
      i <- i + 1
    } else {
      return(NA)
    }
  }
  total + current
}

# clean_amounts(amountStrings)
#   Arguments:  amountStrings = character vector (e.g. "$27.00", "1,234.56", "five hundred")
#   Returns:    numeric vector. Removes $ and commas; converts word amounts when possible.
clean_amounts <- function(amountStrings) {
  x <- trimws(amountStrings)
  result <- numeric(length(x))
  for (i in seq_along(x)) {
    s <- gsub("[$,]", "", x[i])
    s <- trimws(s)
    if (grepl("^[0-9.]+$", s)) {
      result[i] <- as.numeric(s)
    } else {
      result[i] <- word_to_number(x[i])
    }
  }
  result
}

# clean_categories(categoryStrings)
#   Arguments:  categoryStrings = character vector (e.g. "FOOD", "office suppl.")
#   Returns:    character vector, lowercase, with abbreviations normalized
#               (e.g. "office suppl." and "Office Stuff" -> "office supplies", "LUNCH" -> "food").
office_suppl_regex <- paste0(
  "\\boffice\\s+suppl\\.?\\s*$"   # "office suppl" or "office suppl." at end of string
)
office_stuff_regex <- "\\boffice\\s+stuff\\s*$"
lunch_regex <- "\\blunch\\s*$"
clean_categories <- function(categoryStrings) {
  x <- tolower(trimws(categoryStrings))
  x <- sub(office_suppl_regex, "office supplies", x)
  x <- sub(office_stuff_regex, "office supplies", x)
  x <- sub(lunch_regex, "food", x)
  x
}

# clean_comments(commentStrings)
#   Arguments:  commentStrings = character vector; can contain placeholders like "---", "NA"
#   Returns:    character vector; placeholders and empty strings replaced with NA.
placeholder_regex <- "^(---|NA|n/a)$"
clean_comments <- function(commentStrings) {
  x <- trimws(as.character(commentStrings))
  x[grepl(placeholder_regex, x, ignore.case = TRUE)] <- NA
  x[x == ""] <- NA
  x
}

# =============================================================================
# MAIN CLEANUP
# =============================================================================

# cleanUp(expensesDF)
#   Arguments:  expensesDF = data frame with columns employee, date, amount, category, comments
#   Returns:    data frame with employee replaced by employeeFullName, employeeTitle,
#               employeeFirst, employeeMiddle, employeeLast, employeeSuffix; and with
#               date, amount, category, comments cleaned to consistent formats.
#   Example:    cleanUp(expenses)  ->  wide table with cleaned columns
cleanup <- function(expensesDF) {
  df <- as.data.frame(expensesDF, stringsAsFactors = FALSE)
  emp <- parse_employee_column(df$employee)
  df$employee <- NULL
  df <- cbind(emp, df)
  df$date     <- parse_dates(df$date)
  df$amount   <- clean_amounts(df$amount)
  df$category <- clean_categories(df$category)
  df$comments <- clean_comments(df$comments)
  df
}

cleanup(expenses)
```

Given the data above, your output should show the following:
       
<!-- show the values output by the answer code -->
```{r}
cleanup(expenses)
```

::: {.callout-note collapse="true"}
### Click here for one possible solution
<!-- show how to generate answer -->
```{r}
#| ref.label: cleanUp-solution
#| class-output: cleanup-table-scroll
#| echo: true
```
:::


```{=html}
<script>
document.addEventListener('DOMContentLoaded', function () {

  const firstH1 = document.querySelector('main h1') || document.querySelector('h1');
  const sections = document.querySelectorAll('main section');

  function getHeading(section) {
    return section.querySelector(':scope > h1, :scope > h2, :scope > h3, :scope > h4, :scope > h5, :scope > h6');
  }

  function getChildSections(section) {
    return section.querySelectorAll(':scope > section');
  }

  // ===== toggle =====
  function setSection(section, open) {
    section.dataset.open = open;

    const heading = getHeading(section);
    const icon = heading?.querySelector('.toggle-icon');

    const children = Array.from(section.children).filter(el => !/^H[1-6]$/.test(el.tagName));
    children.forEach(el => {
      el.style.display = open ? '' : 'none';
    });

    if (icon) icon.textContent = open ? '▼ ' : '▶ ';
  }

  function toggleSection(section) {
    const open = section.dataset.open !== 'false';
    setSection(section, !open);
  }

  // ===== build UI =====
  sections.forEach(section => {

    const heading = getHeading(section);
    if (!heading) return;
    if (heading === firstH1) return;

    const children = getChildSections(section);
    const hasSubsections = children.length > 0;

    // ===== triangle =====
    const icon = document.createElement('span');
    icon.textContent = '▼ ';
    icon.className = 'toggle-icon';
    icon.style.cursor = 'pointer';
    icon.setAttribute('tabindex', '0');

    heading.prepend(icon);
    heading.style.cursor = 'pointer';

    // initial state
    section.dataset.open = 'true';

    heading.addEventListener('click', function(e){
      if (e.target.tagName === 'A') return;
      toggleSection(section);
    });

    icon.addEventListener('keydown', function(e){
      if (e.key === 'Enter' || e.key === ' ') {
        e.preventDefault();
        toggleSection(section);
      }
    });

    // ===== local buttons =====
    if (hasSubsections) {

      const controls = document.createElement('div');
      controls.style.margin = '0.25rem 0 0.75rem 0';

      const collapseBtn = document.createElement('button');
      collapseBtn.textContent = 'Collapse Sections';
      collapseBtn.style.marginRight = '0.5rem';

      const expandBtn = document.createElement('button');
      expandBtn.textContent = 'Expand Sections';

      controls.appendChild(collapseBtn);
      controls.appendChild(expandBtn);
      heading.after(controls);

      collapseBtn.onclick = () => {
        children.forEach(s => setSection(s, false));
      };

      expandBtn.onclick = () => {
        children.forEach(s => setSection(s, true));
      };
    }

  });

  // ===== global buttons =====
  function setAll(open) {
    sections.forEach(section => {
      const heading = getHeading(section);
      if (!heading || heading === firstH1) return;
      setSection(section, open);
    });
  }

  const globalControls = document.createElement('div');
  globalControls.style.margin = '0.5rem 0 1rem 0';

  const collapseAll = document.createElement('button');
  collapseAll.textContent = 'Collapse Sections';
  collapseAll.style.marginRight = '0.5rem';

  const expandAll = document.createElement('button');
  expandAll.textContent = 'Expand Sections';

  collapseAll.onclick = () => setAll(false);
  expandAll.onclick = () => setAll(true);

  globalControls.appendChild(collapseAll);
  globalControls.appendChild(expandAll);

  if (firstH1) {
    firstH1.after(globalControls);
  }

});
</script>
```


