# Regular Expression Practice - Cleaning Data


```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

## Introduction - about the data

In this assignment, you'll practice using regular expressions to identify and clean messy data. The dataset below contains employee expense records with various inconsistencies in formatting. ***Some*** of the inconsistencies are described below
the data.

### Employee expenses data

Click on the link below to see the full dataset.
The first few rows of data are shown below the link.

::: {.callout-note collapse="true"}
## click here to see the full data set
```{r}
library(tibble)
expenses = tribble(
  ~employee,           ~date,          ~amount,          ~category,         ~comments,
  "Sue Smith",         "1/19/2023",    "59.99",         "food",            "lunch with client",
  "schwartz, joe",     "01/19/2023",   "$27.00",        "office supplies", "paper for printer",
  "mike david harris", "2023-01-19",   "25",            "Office Stuff",    NA,
  "Dr. A. Davis",      "19/1/2023",    "five hundred",  "FOOD",            "NA",
  "Dr Jones",          "1/19/23",      "1,234.56",      "office suppl.",   "chairs",
  "S. Jones Jr",       "19/01/23",     "1000",          "Office supplies", "desk",
  "Conway, Ella Sr.",  "Jan 19, 2023", "$35.23",        "LUNCH",           "---",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    NA,
  "mike green jr.",  "2023/02/15",     "500",           "OFFICE SUPPLIES", NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",       "03/22/2023",   "$156.00",      "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "OFFICE STUFF",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",        "3/14/24",      "500",         "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",
    "Pat Williams",      "2/20/2023",    "150",           "travel",          "train ticket",
  "Chen, Wei",         "20/02/2023",   "$89.50",        "Travel",          "hotel",
  "jennifer ann martinez", "2023-02-21", "42",          "FOOD",            NA,
  "Ms. B. Kim",        "21/2/2023",    "two hundred",  "office supplies", "stapler",
  "J. Thompson",      "2/21/23",      "2,500.00",      "TRAVEL",          "flight",
  "R. Patel Ph.D",     "21/2/23",      "75",            "office suppl.",   "notebooks",
  "Garcia, Maria Jr.", "Feb 22, 2023", "$120.00",       "food",            "team dinner",
  "Dr. Evans",         "3/01/2023",    "95.99",         "Office Stuff",    "supplies",
  "Thompson III",      "3/05/2023",    "200",           "travel",          "conference",
  "Brown, Tom",        "2/15/2023",    "42.50",         "food",            "dinner meeting",
  "  Williams, Kate  ","02/15/2023",   "$89.99",        "office supplies", "toner cartridge",
  "robert james lee",  "2023/02/15",   "75",            "Office Stuff",    NA,
  "mike green jr.",  "2023/02/15",     "500",           "OFFICE SUPPLIES",    NA,
  "Dr. B. Wilson",     "15/2/2023",    "two hundred",   "FOOD",            "NA",
  "Dr Smith",          "2/15/23",      "2,345.67",      "office suppl.",   "monitor",
  "R. Brown Jr",       "March 2, 2023","750",           "Office supplies", "filing cabinet",
  "Martinez, Ana Sr.", "Feb 15, 2023", "$48.75",        "LUNCH",           "---",
  "Johnson,Mary",      "3/22/2023",    "63.45",         "food",            "team lunch",
  "Cooper,  Sam",       "03/22/2023",   "$156.00",      "office supplies", "office decoration",
  "peter michael wu",  "2023-03-22",   "50",            "Office Stuff",    NA,
  "Dr. C. Taylor",     "2/3/2023",     "one thousand",  "FOOD",            "NA",
  "Drake Wilson",        "3/14/24",      "500",         "Office supplies", "printer",
  "Dr Chen",           "3/2/23",       "3,456.78",      "office suppl.",   "laptop",
  "T. Wilson Jr",      "Jun 27 2024",  "500",           "Office supplies", "printer",
  "Rodriguez, Eva Sr.","Mar 22, 2023", "$92.45",        "LUNCH",           "---",

)
expenses = as.data.frame(expenses)
```
:::


```{r}
numrows = 10
cat("There are a total of ", nrow(expenses), " rows. The following is 
    just the first ", numrows, " rows.", sep="")
head(expenses, 10)
```

- Inconsistencies with the `employee` column entries

  * some names are in "first last" (e.g. "Sue Smith") format and 
    some names are in "last, first" format (e.g. "Brown, Tom")
  * some names are capitalized, some aren't
  * some names have middle include middle names, some don't
  * some names include titles, some don't
  * some names include a title, e.g. Dr. , while some don't
  * some of the titles contain a period after the title, some don't
  * some names include a suffix (e.g. Jr), some don't
  * some names are missing first names
  * some names use periods (e.g. J. Thompson), some don't
  * possibly other inconsistencies ...

  
- Inconsistencies with `date` column

  * some dates are in mm/dd/yyyy format (USA style)
  * some dates are in dd/mm/yyyy format (European style)
  * some dates are in yyyy-mm-dd format (standard for many computer systems)
  * some dates use 4 digit years, some use 2 digit years
  * some dates have a leading zero for single digit days or months
  * possibly other inconsistencies ...

- Inconsistencies in the `amount` column

  * some values have $ signs, some don't, 
  * some have commas (e.g. 1,234.56) some don't
  * some are missing the cents (e.g. "75")
  * some are written in words (e.g. "five hundred")
  * possibly other inconsistencies ...

- Inconsistencies in the `category` column

  * the same categories are written differently (e.g. "food" vs "FOOD", 
    "office supplies" vs "Office suppl." vs "OFFICE STUFF")
  * possibly other inconsistencies ...

- Inconsistencies in the `comments` column

  * blanks are expressed differently: NA vs "NA" vs "---".
  
    Note that in this column, other differences are more forgivable since 
    comments are by their very nature written by people differently and are 
    not expected to be in a standard form. However, there should be a standard
    way to express that there are no comments for a given row.


## Questions - PART I

Your task is to write R code using regular expressions to identify different patterns in the data.

You may use any of the regex functions we covered in class:
- `grep()`: Find patterns and return matching indices or values
- `grepl()`: Find patterns and return TRUE/FALSE for each element
- `sub()`: Replace the first occurrence of a pattern
- `gsub()`: Replace all occurrences of a pattern
- `strsplit()`: Split strings based on a pattern

For each question:
1. First try to solve it yourself
2. If you get stuck, try to break down the pattern you're looking for into smaller pieces
3. Test your solution with the sample data to make sure it catches all cases

### Name Format Questions

1. Answer all parts

    a) Write R code using regex to identify all names that are in "last, first" 
       format (e.g., "Schwartz, Joe", including those with suffixes like "Sr." or
       "Jr."). Make sure that you allow for any number of spaces between the 
       comma and the first name (including if there aren't any spaces).

       <!-- generate answer but don't show this code -->
       ```{r}
       #| label: answer-1a
       #| echo: false
       #| output: false
      
       answer = 
         grep("^[a-zA-Z]+,.+$", expenses$employee, value=TRUE)
       ```

       Given the data above, your output should show the following:
       
       <!-- show the values output by the answer code -->
       ```{r}
       #| echo: false
       cat(answer, sep = "\n")
       ```

       ::: {.callout-note collapse="true"}
       ### Click here for answer
       <!-- show how to generate answer -->
       ```{r}
       #| ref.label: answer-1a
       #| echo: true
       ```
       <!-- show the output -->
       ```{r}
       cat(answer, sep="\n")
       ```
       :::

    b) Modify your answer to show only names that are strictly in  
       last, first form (without a suffix) 

       <!-- generate answer but don't show this code -->
       ```{r}
       #| label: answer-1b
       #| echo: false
       #| output: false
      
       answer = 
         grep("^[a-zA-Z]+, *[a-zA-Z]+$", expenses$employee, value=TRUE)
       ```

       Given the data above, your output should show the following:
       
       <!-- show the values output by the answer code -->
       ```{r}
       #| echo: false
       cat(answer, sep = "\n")
       ```

       ::: {.callout-note collapse="true"}
       ### Click here for answer
       <!-- show how to generate answer -->
       ```{r}
       #| ref.label: answer-1b
       #| echo: true
       ```
       <!-- show the output -->
       ```{r}
       cat(answer, sep="\n")
       ```
       :::

   

2. Write R code using regex to display only names that begin with "Dr" (with or without a period).
   Make sure that you don't display names (e.g. "Drake Wilson") that begin with Dr 
   but who aren't doctors.


   <!-- generate answer but don't show this code -->
   ```{r}
   #| label: answer-2
   #| echo: false
   #| output: false
      
   answer = 
      grep("^Dr[. ]", expenses$employee, value=TRUE)
   ```

   Given the data above, your output should show the following:
       
   <!-- show the values output by the answer code -->
   ```{r}
   #| echo: false
   cat(answer, sep = "\n")
   ```

   ::: {.callout-note collapse="true"}
   ### Click here for answer
   <!-- show how to generate answer -->
   ```{r}
   #| ref.label: answer-2
   #| echo: true
   ```
   <!-- show the output -->
   ```{r}
   cat(answer, sep="\n")
   ```
   :::


3. Write R code using regex to identify all names that are completely lowercase (no capitalization).

4. Write R code using regex to identify all names that include a middle name or middle initial.

5. Write R code using regex to identify all names that have a suffix (Sr. or Jr., with or without the period).

### Date Format Questions

6. Write R code using regex to identify dates in the MM/DD/YYYY format (with or without leading zeros), ensuring valid months (1-12) and days (1-31).

7. Write R code using regex to identify dates in the European format (DD/MM/YYYY or DD/MM/YY), ensuring valid days and months.

8. Write R code using regex to identify dates in ISO format (YYYY-MM-DD).

9. Write R code using regex to identify dates that use written month names (e.g., "Jan 19, 2023", "Feb 15, 2023").

### Amount Format Questions

10. Write R code using regex to identify amounts that include a dollar sign (e.g., "$27.00", "$156.00").

11. Write R code using regex to identify amounts that are written out in words (e.g., "five hundred", "one thousand").

12. Write R code using regex to identify amounts that include thousand separators (commas) (e.g., "1,234.56", "2,345.67").

13. Write R code using regex to identify amounts that are whole numbers without decimal places (e.g., "25", "1000").

### Category Format Questions

14. Write R code using regex to identify categories that are entirely uppercase (e.g., "FOOD", "LUNCH").

15. Write R code using regex to identify all variations of "office supplies" (including abbreviated forms and different capitalizations).

### Comment Format Questions

16. Write R code using regex to identify comments that are either NA, "NA", or 
    consist entirely of dashes (e.g "---" and "-----").

17. Write R code using regex to identify comments that describe meetings (containing words like "lunch", "dinner", "client", "team").

### Challenge Questions

18. Write R code using regex to extract just the last name from any name format in the employee column. Your solution should work for all name formats in the dataset.

19. Write R code using regex to convert all dates to YYYY-MM-DD format, regardless of their original format. Your solution should handle all date formats in the dataset.

20. Write R code using regex to standardize all amounts to a decimal number format (e.g., "1234.56"), removing dollar signs, commas, and converting written amounts to numbers.
