# <yrChapterNumber>27. </yrChapterNumber> Get the data - "books database" and other tables

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

As we learn about databases and SQL we will need to refer to various 
collections of data (i.e. "tables" of data in SQL terminology). 
See below for info on how to download and install
the various collections of data into R.


```{r}
#| include: FALSE

require(sqldf)
file.copy(from="data/grades.csv", to=".")
file.copy(from="data/sqlbolt_tables-v007.RData", to=".")
```

## Prepare to import the data

In order to read the data into R and to access the data using SQL, we will
need to load the R packages "readr" and "sqldf". We do so with
the following commands:

```{r}
#| output: false
#| warning: false

if(!require(readr)){install.packages("readr");require(readr);}
if(!require(sqldf)){install.packages("sqldf");require(sqldf);}
```


## import the data: grades

Download the CSV file: [grades.csv](grades.csv) 

This a .csv file that contains information
about students and their grades.
You can read the contents of the file into R 
with the read.csv function as shown below.

```{r}
grades = read.csv("grades.csv")
```
  
Once you've done so, you can display the data:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

## import the data: customers, orders, movies, north_american_cities, etc

Download the RData file: [sqlbolt_tables-v007.RData](sqlbolt_tables-v007.RData)

A .RData file contains a copy of R's environment variables. You can 
save you environment variables to an RData file with the `save()` function.
(see ?save). Once you have an RData file, you can load those variables
into your R environment with the `load()` function (see ?load).

This RData file 
contains the following different dataframe variables. When using the 
sqldf() function, each dataframe 
variable can used as though it were a Relational Database table.

- customers
- orders
- movies
- north_american_cities
- as well as some others (boxOffice, buildings, employees)

Use the load function as shown below to load the data into R.

```{r}
load("sqlbolt_tables-v007.RData")
```

Below are the contents of these tables:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - movies table
```{r}
sqldf("select * from movies")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - orders table
```{r}
sqldf("select * from orders")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - customers table
```{r}
sqldf("select * from customers")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - north_american_cities table
```{r}
sqldf("select * from north_american_cities")
```
:::

## The "Books Database"

Download the CSV files for the "books database".
These CSV files together comprise the data for
the "books database". 

- [titles.csv](data/booksDatabase/titles.csv)
    
- [authors.csv](data/booksDatabase/authors.csv)
    
- [publishers.csv](data/booksDatabase/publishers.csv)
    
- [title_authors.csv](data/booksDatabase/title_authors.csv)
  
- [royalties.csv](data/booksDatabase/royalties.csv)

Download a tutorial on using Relational databases.

- [booksDatabase description and Questions](data/booksDatabaseDescription-withQuestions-ANSWERS-v024.docx)

This word document contains:

(a) a complete description of the "books database"
(b) a tutorial on how "relational databases" are structured
(c) a tutorial on Entity Relationship Diagrams (ERDs)
(d) a set of questions and answers

Once you've downloaded the various CSV files you can run the following commands to 
import the data into R.

### Import the data (books database)




```{r}
# Read in the data for the books database - see the 
titles = read_csv("data/booksDatabase/titles.csv", na="NULL", show_col_types=FALSE)
authors = read_csv("data/booksDatabase/authors.csv", na="NULL", show_col_types=FALSE)
publishers = read_csv("data/booksDatabase/publishers.csv", na="NULL", show_col_types=FALSE)
title_authors = read_csv("data/booksDatabase/title_authors.csv", na="NULL", show_col_types=FALSE)
royalties = read_csv("data/booksDatabase/royalties.csv", na="NULL", show_col_types=FALSE)
```

### Books Database - Entity Relationship Diagram (ERD)

The following is an "Entity Relationship Diagram (ERD)" that 
describes the relationships between the tables. See the Word document
linked above for more info:

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': { 'primaryColor': '#ffffff', 'primaryTextColor': '#0000ff', 'primaryBorderColor': '#000000', 'lineColor': '#000000', 'secondaryColor': '#ffffff', 'tertiaryColor': '#ffffff', 'mainBkg': '#ffffff', 'background': '#ffffff', 'edgeLabelBackground': '#ffffff' }}}%%
erDiagram
    authors ||--o{ title_authors : ""
    title_authors }|--|| titles : ""
    titles ||--o| royalties : ""
    publishers |o--o{ titles : ""
    
    authors {
        string au_id PK
        string au_fname
        string au_lname
        string phone
        string address
        string city
        string state
        string zip
    }
    
    title_authors {
        string title_id FK1
        string au_id FK2
        int au_order
        float royalty_share
    }
    
    titles {
        string title_id PK
        string title_name
        string type
        string pub_id FK
        int pages
        float price
        int sales
        date pubdate
    }
    
    publishers {
        string pub_id PK
        string pub_name
        string city
        string state
        string country
    }
    
    royalties {
        string title_id PK
        float advance
        float royalty_rate
    }
```

Below are the contents of these tables:


::: {.callout-tip collapse="true"}
#### Click here - titles table
```{r}
sqldf("select * from titles")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - publishers table
```{r}
sqldf("select * from publishers")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - title_authors table
```{r}
sqldf("select * from title_authors")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - authors table
```{r}
sqldf("select * from authors")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - royalties table
```{r}
sqldf("select * from royalties")
```
:::
