# <yrChapterNumber>30. </yrChapterNumber> Get the data - "books database" and other tables

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

As we learn about databases and SQL we will need to refer to various 
collections of data (i.e. "tables" of data in SQL terminology). 
See below for info on how to download and install
the various collections of data into R.


```{r}
#| include: FALSE

require(sqldf)
file.copy(from="data/grades.csv", to=".")
file.copy(from="data/sqlbolt_tables-v007.RData", to=".")
```

## Download all of the files

Download the files by clicking on the links below.
To load the various collections of data into R
see the instructions in the sections below.

- The following file contains the grades table:  

  * [grades.csv](grades.csv)

- The following file contains several tables of data    
  (customers, orders, movies, north_american_cities, boxOffice, 
  buildings, employees)  

  * [sqlbolt_tables-v007.RData](sqlbolt_tables-v007.RData)

- The following files are all associated with the "Books database": 

  * [booksDatabase description and Questions](data/booksDatabaseDescription-withQuestions-ANSWERS-v020.docx) 
  
    This file contains a description of the database as well as a discussion of
    different concepts related to databases in general. You should use this
    file as a tutorial about the books database. The end of this file
    contains many questions/answers to test your knowledge of SQL.
    

  * The following are the files that you can use to import the data
    for the database into R.
    
    - [titles.csv](data/booksDatabase/titles.csv)
    
    - [authors.csv](data/booksDatabase/authors.csv)
    
    - [publishers.csv](data/booksDatabase/publishers.csv)
    
    - [title_authors.csv](data/booksDatabase/title_authors.csv)
  
    - [royalties](data/booksDatabase/royalties.csv)


## Prepare to import the data

In order to read the data into R and to access the data using SQL, we will
need to load the R packages "readr" and "sqldf". We do so with
the following commands:

```{r}
#| output: false
#| warning: false

if(!require(readr)){install.packages("readr");require(readr);}
if(!require(sqldf)){install.packages("sqldf");require(sqldf);}
```


## import the data: grades

[grades.csv](grades.csv) is a .csv file that contains information
about students and their grades.
You can read the contents of the file into R 
with the read.csv function as shown below.

```{r}
grades = read.csv("grades.csv")
```
  
Once you've done so, you can display the data:

```{r}
sqldf("select * from grades")
```


## import the data: customers, orders, movies, north_american_cities, etc

The file [sqlbolt_tables-v007.RData](sqlbolt_tables-v007.RData)
contains the following different collections of information:

- customers
- orders
- movies
- north_american_cities
- as well as some others (boxOffice, buildings, employees)

Use the load function as shown below to load the data into R.

```{r}
load("sqlbolt_tables-v007.RData")
```

Below are the contents of these tables:

```{r}
sqldf("select * from grades")
sqldf("select * from movies")
sqldf("select * from orders")
sqldf("select * from customers")
sqldf("select * from north_american_cities")
```


## import the data: "books database"

The "books database" is a collection of several relational database tables. 
The links to download the files for this database are located at the top
of this webpage.

Pay special attention to the file
[booksDatabase description and Questions](data/booksDatabaseDescription-withQuestions-ANSWERS-v020.docx)
which contains a description of the database as well as a discussion of
different concepts related to databases in general. You should use this
file as a tutorial about the books database. The end of this file
contains many questions/answers to test your knowledge of SQL.

Once you've downloaded the various files you can run the following commands to 
import the data into R.

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The books database contains information about various books, authors,
# publishers, etc.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Read in the data for the books database - see the 
titles = read_csv("data/booksDatabase/titles.csv", na="NULL", show_col_types=FALSE)
authors = read_csv("data/booksDatabase/authors.csv", na="NULL", show_col_types=FALSE)
publishers = read_csv("data/booksDatabase/publishers.csv", na="NULL", show_col_types=FALSE)
title_authors = read_csv("data/booksDatabase/title_authors.csv", na="NULL", show_col_types=FALSE)
royalties = read_csv("data/booksDatabase/royalties.csv", na="NULL", show_col_types=FALSE)
```

Below are the contents of these tables:

```{r}
sqldf("select * from titles")
sqldf("select * from publishers")
sqldf("select * from title_authors")
sqldf("select * from authors")
sqldf("select * from royalties")
```

