# <yrChapterNumber>28. </yrChapterNumber> Get the data for the SQL lessons

## Download the data

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

The following are the different collections of data. Each collection can 
be downloaded in different formats (e.g. .sqlite, .csv, .RData). 
Each format should contain the same information for a given dataset. 
The data is made available in different formats so that you can 
access the data using different tools, specifically:

- a .sqlite database for use with R and many other sqlite tools, including
  **yrSqliteViewer** (see below for more info)

- a set of .csv files for use with R (and many other tools)

- a .RData file for use with R


See the information below for infomation 
about how to use each of the different formats. For now, 
if you're just getting started, I recommend that you download the .sqlite 
versions for each collection of data and then follow the instructions
below for how to download and use the **yrSqliteViewer** tool.




### grades table

A single table named "grades"

  * Download as a sqlite file:   
    [grades.sqlite](data/grades.sqlite) 

  * Download as a CSV file:   
    <a href="data/grades.csv" download>grades.csv</a>
  


### sqlbolt tables

A collection of a few tables that come from some of the lessons on the website
<https://sqlbolt.com/>. These are stored in the file "sqlbolt_tables-v007.RData".

  * Download as a sqlite file:   
    [sqlbolt_tables-v007.sqlite](data/sqlbolt_tables-v007.sqlite) 
  
  * Download as a .RData file:   
    <a href="data/sqlbolt_tables-v007.RData" download>sqlbolt_tables-v007.RData</a>

### movies practice table

A standalone movies table for extra SQL practice questions is available here:

- Practice page: [SQL Practice: Movies Table](sql0016-moviesPractice-v001.qmd)
- Download as a sqlite file: [moviesPractice-v001.sqlite](data/moviesPractice-v001.sqlite)
- Download as a CSV file: [moviesPractice-v001.csv](data/moviesPractice-v001.csv)
 
### The "books" database 

This is  a collection of several tables (titles, publishers, authors, royalties, title_authors)
The following tutorial uses this data to explain relational databases and SQL:

  - [booksDatabaseDescription-withQuestions-ANSWERS-v024.docx](data/booksDatabaseDescription-withQuestions-ANSWERS-v024.docx)

Also, see the [section below](#yrBooksErd) that shows the 
["Entity Relationship Diagram"](#yrBooksErd) and the 
[contents of the tables](#yrBooksData) for this database.

#### Download the data

Download the data here in one of the following formats:
  
  - Download as a single sqlite file that contains all of the tables  
    [books-v025.sqlite](data/books-v025.sqlite) 
  
  - Download as separate csv files for each table:  
    [titles.csv](data/booksDatabase/titles.csv)  
    [authors.csv](data/booksDatabase/authors.csv)  
    [publishers.csv](data/booksDatabase/publishers.csv)  
    [title_authors.csv](data/booksDatabase/title_authors.csv)  
    [royalties.csv](data/booksDatabase/royalties.csv)  
  
  - Download as a single ,RData file that contains all tables as R dataframes:  
    <a href="data/booksDatabase-v004.RData" download>booksDatabase-v004.RData</a>




```{r}
#| include: FALSE

require(sqldf)
file.copy(from="data/grades.csv", to=".")      # original
file.copy(from="data/grades.sqlite", to=".")
file.copy(from="data/grades.RData", to=".")

file.copy(from="data/sqlbolt_tables-v007.RData", to=".")   # original
file.copy(from="data/sqlbolt_tables-v007.sqlite", to=".") 
```


### Converting between formats

The following is an R file that can be used to convert between the various
formats. You do NOT need to use this, but it may be useful to you now or in the future.

::: {.callout-note collapse="true"}
##### Click here for a set of R functions to convert between formats

Click below to download a single R file that contains all the functions. 

- <a href="yrSqliteUtils-v001.R" download>Download </a>

:::: {.callout-note collapse="true"}
##### Click here to see the code
```{r}
#| file: yrSqliteUtils-v001.R
```
::::
:::



## How to access the data using the yrSqliteViewer.html

The yrSqliteViewer.html tool is a single file that can be downloaded 
to your computer that allows you to work with .sqlite files. 

Follow the instructions here 
<https://github.com/y-rosenthal/yrSqliteViewer>
to download yrSqliteViewer.html and use
it to access .sqlite files.


## How to access the data using R

### Prepare to import the data

In order to read the data into R and to access the data using SQL, we will
need to load the R packages "readr" and "sqldf". We do so with
the following commands:

```{r}
#| output: false
#| warning: false

if(!require(readr)){install.packages("readr");require(readr);}
if(!require(sqldf)){install.packages("sqldf");require(sqldf);}
```


### Importing .csv files into R

Download the CSV file: [grades.csv](data/grades.csv) 

This a .csv file that contains information
about students and their grades.
You can read the contents of the file into R 
with the read.csv function as shown below.

```{r}
grades = read.csv("grades.csv")
```
  
Once you've done so, you can display the data:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

### Importing .RData files into R

Download the RData file: [sqlbolt_tables-v007.RData](data/sqlbolt_tables-v007.RData)

A .RData file contains a copy of R's environment variables. You can 
save you environment variables to an RData file with the `save()` function.
(see ?save). Once you have an RData file, you can load those variables
into your R environment with the `load()` function (see ?load).

This RData file 
contains the following different dataframe variables. When using the 
sqldf() function, each dataframe 
variable can used as though it were a Relational Database table.

- customers
- orders
- movies
- north_american_cities
- as well as some others (boxOffice, buildings, employees)

Use the load function as shown below to load the data into R.

```{r}
load("sqlbolt_tables-v007.RData")
```

Below are the contents of these tables:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - movies table
```{r}
sqldf("select * from movies")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - orders table
```{r}
sqldf("select * from orders")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - customers table
```{r}
sqldf("select * from customers")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - north_american_cities table
```{r}
sqldf("select * from north_american_cities")
```
:::

{{< include yrBooksDatabase-common-v001.qmd >}}


