# <yrChapterNumber>28. </yrChapterNumber> Get the data - "books database" and other tables

```{r}
#| echo: false
#| output: false

# dataframes should display in the following width without "wrapping"
options(width = 999)
```

As we learn about databases and SQL we will need to refer to various 
collections of data (i.e. "tables" of data in SQL terminology). 

This section explains how to import data into R for use with SQL and also
provides the various data files that we will be using in the upcoming SQL
lessons.

## Different ways to import data for use with SQL

SQL is a language that is used with Relational Database Management SOftware (RDBMS).




```{r}
#| include: FALSE

require(sqldf)
file.copy(from="data/grades.csv", to=".")
file.copy(from="data/sqlbolt_tables-v007.RData", to=".")
```

## Prepare to import the data

In order to read the data into R and to access the data using SQL, we will
need to load the R packages "readr" and "sqldf". We do so with
the following commands:

```{r}
#| output: false
#| warning: false

if(!require(readr)){install.packages("readr");require(readr);}
if(!require(sqldf)){install.packages("sqldf");require(sqldf);}
```


## import the data: grades

Download the CSV file: [grades.csv](grades.csv) 

This a .csv file that contains information
about students and their grades.
You can read the contents of the file into R 
with the read.csv function as shown below.

```{r}
grades = read.csv("grades.csv")
```
  
Once you've done so, you can display the data:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

## import the data: customers, orders, movies, north_american_cities, etc

Download the RData file: [sqlbolt_tables-v007.RData](sqlbolt_tables-v007.RData)

A .RData file contains a copy of R's environment variables. You can 
save you environment variables to an RData file with the `save()` function.
(see ?save). Once you have an RData file, you can load those variables
into your R environment with the `load()` function (see ?load).

This RData file 
contains the following different dataframe variables. When using the 
sqldf() function, each dataframe 
variable can used as though it were a Relational Database table.

- customers
- orders
- movies
- north_american_cities
- as well as some others (boxOffice, buildings, employees)

Use the load function as shown below to load the data into R.

```{r}
load("sqlbolt_tables-v007.RData")
```

Below are the contents of these tables:

::: {.callout-tip collapse="true"}
#### Click here - grades table
```{r}
sqldf("select * from grades")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - movies table
```{r}
sqldf("select * from movies")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - orders table
```{r}
sqldf("select * from orders")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - customers table
```{r}
sqldf("select * from customers")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - north_american_cities table
```{r}
sqldf("select * from north_american_cities")
```
:::

{{< include yrBooksDatabase-common-v001.qmd >}}


