# regex test

```{r}
#| include: false

# helper function to ensure output is displayed one element per line
yrOnePerLine <- function(x, check = TRUE, debug = FALSE) {
  
  n <- length(x)
  
  if (n == 0) {
    # match print() for empty vectors
    print(x)
    return(invisible(x))
  }
  
  # capture full print output (reference)
  if (check) {
    ref <- capture.output(print(x))
  }
  
  # build output by printing one element at a time
  out <- unlist(lapply(seq_along(x), function(i) {
    capture.output(print(x[i]))
  }))
  
  # flatten (each element gives one line already)
  out <- trimws(out)
  
  if (check) {
    # ALSO normalize ref to per-element lines for fair comparison
    ref_split <- unlist(strsplit(ref, "\n"))
    ref_split <- trimws(ref_split)
    
    # reconstruct expected per-element version from ref
    ref_rebuilt <- unlist(lapply(seq_along(x), function(i) {
      trimws(capture.output(print(x[i])))
    }))
    
    if (!identical(out, ref_rebuilt)) {
      if (debug) {
        cat("---- DEBUG START ----\n")
        cat("\nOriginal print(x):\n"); print(ref)
        cat("\nPer-element print (OUT):\n"); print(out)
        cat("\nRebuilt reference:\n"); print(ref_rebuilt)
        cat("\nclass(x):\n"); print(class(x))
        cat("---- DEBUG END ----\n")
      }
      stop("Output does not match base R print() format.")
    }
  }
  
  cat(out, sep = "\n")
}

test_list <- list(
  simple = c("a", "b", "c"),
  tabs = c("a\tb", "c\td"),
  spaces = c("a  b", "c   d"),
  mixed = c("a \t  b", "\t\tc"),
  numbers = c(1, 2.5, NA),
  factors = factor(c("a", "b", "a")),
  empty = character(0)
)

for (nm in names(test_list)) {
  cat("\n==== TEST:", nm, "====\n")
  try(yrOnePerLine(test_list[[nm]], debug = TRUE))
}
```

## Instructions

1. You may use 1 sheet of notes (double sided) no computer, no other materials.

2. Unless told otherwise, all regex patterns are to be written as they would be used in R.

3. **When writing code:** indicate a space with the word **SPACE** or an **underscore i.e. _** 
   so that I can see it when I'm grading.

4. **When writing code:** your answer **must work even if the data for the question would be different**. 
   The data given is just one example.

# Part I - reading code

For each question write the output of the code EXACTLY as it would appear in R.



--- 

For the following questions, assume that ***stuff*** is defined as:

```{r}
stuff = c("abc", "def", "123.456", "I hope -- you do well.")
```

## Q1

```{r}
answer = 
grep("f", stuff, value=FALSE)   # value=FALSE
answer
yrOnePerLine(answer)
```
