# <yrChapterNumber>40. </yrChapterNumber> Parsing XML using R

```{r}
#| include: false

# do this so we can open the file later without a path
require(fs)
file_copy('data/booksDatabase/books-v025.xml', '.', overwrite=TRUE)
```

This tutorial explains how to parse XML files in R and transform the data into usable dataframes. We'll work with a books database XML file and create dataframes that mirror common relational database tables.

## What is XML?

XML (eXtensible Markup Language) is a markup language that defines a set of rules for encoding documents in a format that is both human-readable and machine-readable. XML files use tags to define elements and their attributes, creating a hierarchical structure.

Key components of XML:

-   **Elements**: Defined by tags (e.g., `<book>...</book>`)
-   **Attributes**: Properties within tags (e.g., `id="B001"`)
-   **Nesting**: Elements can contain other elements
-   **Self-closing tags**: Elements with no content (e.g., `<pulitzerPrizeWinner />`)

## Required Packages

For XML processing in R, we'll use the `xml2` package for parsing and the `tidyverse` for data manipulation:

```{r}
#| message: false
#| warning: false

# Install packages if needed
# install.packages(c("xml2", "tidyverse"))

library(xml2)      # For XML parsing
library(tidyverse) # For data manipulation
library(dplyr)     # For data transformation
```


## Loading the XML File

First, let's load our XML file:

```{r}
# Load the XML file
books_xml <- read_xml("books-v025.xml")

# Quick look at the structure
xml_structure(books_xml)
```

The `read_xml()` function from the `xml2` package reads the XML file into R, creating an XML document object that we can query and manipulate.

## Understanding XML Node Selection

To extract data from XML, we need to understand how to select nodes. The `xml2` package provides functions like:

-   `xml_find_all()`: Find all nodes matching an XPath expression
-   `xml_find_first()`: Find the first node matching an XPath expression
-   `xml_attr()`: Extract an attribute value
-   `xml_text()`: Extract the text content

XPath is a query language for selecting nodes in XML documents. Basic XPath patterns include:

-   `/books/book`: All book elements that are children of the books element
-   `//author`: All author elements anywhere in the document
-   `/books/book[@id="B001"]`: The book element with attribute id="B001"
-   `//book/authors/author[@primary="true"]`: All primary authors

## Creating the Dataframes

Now let's create dataframes that mirror the CSV files mentioned. We'll extract the relevant data from the XML and ensure we don't create duplicate entries.

### 1. Publishers Dataframe

Let's start by creating a dataframe for publishers:

```{r}
# Extract all publisher nodes
publisher_nodes <- xml_find_all(books_xml, "//publisher")

# Extract book IDs to use as identifiers
book_ids <- xml_find_all(books_xml, "//book") %>%
  xml_attr("id")

# Create publishers dataframe
publishers_df <- tibble(
  pub_id = character(),
  pub_name = character(),
  city = character(),
  state = character(),
  country = character()
)

# Populate publishers dataframe
for (i in seq_along(publisher_nodes)) {
  pub_name <- xml_text(publisher_nodes[i])
  city <- xml_attr(publisher_nodes[i], "city")
  state <- xml_attr(publisher_nodes[i], "state")
  country <- xml_attr(publisher_nodes[i], "country")
  
  # Check if this publisher is already in the dataframe
  if (!pub_name %in% publishers_df$pub_name) {
    # Create a publisher ID (using P prefix instead of B)
    pub_id <- paste0("P", sprintf("%03d", nrow(publishers_df) + 1))
    
    # Add to dataframe
    publishers_df <- publishers_df %>%
      add_row(
        pub_id = pub_id,
        pub_name = pub_name,
        city = city,
        state = state,
        country = country
      )
  }
}

# Display the publishers dataframe
publishers_df
```

### 2. Authors Dataframe

Now, let's create the authors dataframe:

```{r}
# Extract all author nodes
author_nodes <- xml_find_all(books_xml, "//author")

# Create authors dataframe
authors_df <- tibble(
  au_id = character(),
  au_fname = character(),
  au_lname = character(),
  phone = character(),
  address = character(),
  city = character(),
  state = character(),
  zip = integer()
)

# Process each author
for (i in seq_along(author_nodes)) {
  # Extract author full name and split into first and last
  author_full_name <- xml_text(author_nodes[i]) %>%
    str_trim() %>%
    str_split(" ") %>%
    unlist()
  
  au_fname <- author_full_name[1]
  au_lname <- ifelse(length(author_full_name) > 1, author_full_name[2], "")
  
  # Extract contact details
  phone <- xml_find_first(author_nodes[i], "./phone") %>%
    xml_text()
  
  address_node <- xml_find_first(author_nodes[i], "./address")
  street <- xml_find_first(address_node, "./street") %>% xml_text()
  city <- xml_find_first(address_node, "./city") %>% xml_text()
  state <- xml_find_first(address_node, "./state") %>% xml_text()
  zip <- xml_find_first(address_node, "./zip") %>% xml_text() %>% as.integer()
  
  # Check if this author is already in the dataframe
  # We'll use name and address as a unique identifier
  already_exists <- FALSE
  for (j in 1:nrow(authors_df)) {
    if (authors_df$au_fname[j] == au_fname && 
        authors_df$au_lname[j] == au_lname && 
        authors_df$city[j] == city && 
        authors_df$state[j] == state) {
      already_exists <- TRUE
      break
    }
  }
  
  if (!already_exists) {
    # Create an author ID
    au_id <- paste0("A", sprintf("%03d", nrow(authors_df) + 1))
    
    # Add to dataframe
    authors_df <- authors_df %>%
      add_row(
        au_id = au_id,
        au_fname = au_fname,
        au_lname = au_lname,
        phone = phone,
        address = street,
        city = city,
        state = state,
        zip = zip
      )
  }
}

# Display the authors dataframe
authors_df
```

### 3. Titles Dataframe

Next, let's create the titles dataframe:

```{r}
# Extract all book nodes
book_nodes <- xml_find_all(books_xml, "//book")

# Create titles dataframe
titles_df <- tibble(
  title_id = character(),
  title_name = character(),
  type = character(),
  pub_id = character(),
  pages = numeric(),
  price = numeric(),
  sales = numeric(),
  pubdate = character()
)

# Process each book
for (i in seq_along(book_nodes)) {
  title_id <- xml_attr(book_nodes[i], "id")
  title_name <- xml_attr(book_nodes[i], "title")
  book_type <- xml_attr(book_nodes[i], "type")
  pages <- xml_attr(book_nodes[i], "pages") %>% as.numeric()
  price <- xml_attr(book_nodes[i], "price") %>% as.numeric()
  sales <- xml_attr(book_nodes[i], "sales") %>% as.numeric()
  pubdate <- xml_attr(book_nodes[i], "pubdate")
  
  # Find the publisher and get its ID from our publishers dataframe
  publisher_name <- xml_find_first(book_nodes[i], "./publisher") %>% xml_text()
  pub_id <- publishers_df %>% 
    filter(pub_name == publisher_name) %>% 
    pull(pub_id)
  
  # Add to dataframe
  titles_df <- titles_df %>%
    add_row(
      title_id = title_id,
      title_name = title_name,
      type = book_type,
      pub_id = pub_id,
      pages = pages,
      price = price,
      sales = sales,
      pubdate = pubdate
    )
}

# Display the titles dataframe
titles_df
```

### 4. Royalties Dataframe

Now, let's extract the royalty information:

```{r}
# Create royalties dataframe
royalties_df <- tibble(
  title_id = character(),
  advance = numeric(),
  royalty_rate = numeric()
)

# Process each book for royalty information
for (i in seq_along(book_nodes)) {
  title_id <- xml_attr(book_nodes[i], "id")
  
  # Extract royalty information
  royalty_node <- xml_find_first(book_nodes[i], "./royalty")
  advance <- xml_attr(royalty_node, "advance") %>% as.numeric()
  royalty_rate <- xml_attr(royalty_node, "rate") %>% as.numeric()
  
  # Add to dataframe
  royalties_df <- royalties_df %>%
    add_row(
      title_id = title_id,
      advance = advance,
      royalty_rate = royalty_rate
    )
}

# Display the royalties dataframe
royalties_df
```

### 5. Title Authors Dataframe

Finally, let's create the title_authors dataframe that links books to authors:

```{r}
# Create title_authors dataframe
title_authors_df <- tibble(
  title_id = character(),
  au_id = character(),
  au_order = integer(),
  royalty_shares = numeric()
)

# Process each book for author information
for (i in seq_along(book_nodes)) {
  title_id <- xml_attr(book_nodes[i], "id")
  
  # Extract all authors for this book
  book_author_nodes <- xml_find_all(book_nodes[i], "./authors/author")
  
  for (j in seq_along(book_author_nodes)) {
    # Extract author information
    au_order <- xml_attr(book_author_nodes[j], "order") %>% as.integer()
    royalty_shares <- xml_attr(book_author_nodes[j], "royalty_shares") %>% as.numeric()
    
    # Get author name
    author_full_name <- xml_text(book_author_nodes[j]) %>%
      str_trim() %>%
      str_split(" ") %>%
      unlist()
    
    au_fname <- author_full_name[1]
    au_lname <- ifelse(length(author_full_name) > 1, author_full_name[2], "")
    
    # Find the author ID from our authors dataframe
    au_id <- authors_df %>% 
      filter(au_fname == au_fname, au_lname == au_lname) %>% 
      pull(au_id)
    
    # Add to dataframe
    title_authors_df <- title_authors_df %>%
      add_row(
        title_id = title_id,
        au_id = au_id,
        au_order = au_order,
        royalty_shares = royalty_shares
      )
  }
}

# Display the title_authors dataframe
title_authors_df
```

## Complete Dataset

Now we have created five dataframes that mirror the structure of the CSV files:

1.  `publishers_df`: Information about publishers
2.  `authors_df`: Information about authors
3.  `titles_df`: Information about book titles
4.  `royalties_df`: Information about royalty agreements
5.  `title_authors_df`: Linking table between books and authors

## Exporting the Dataframes

If you want to save these dataframes as CSV files:

```{r}
# Export dataframes to CSV
write_csv(publishers_df, "publishers_extracted.csv")
write_csv(authors_df, "authors_extracted.csv")
write_csv(titles_df, "titles_extracted.csv")
write_csv(royalties_df, "royalties_extracted.csv")
write_csv(title_authors_df, "title_authors_extracted.csv")
```

## Common Challenges When Parsing XML

When working with XML files, you may encounter these common challenges:

1.  **Namespace issues**: Some XML files use namespaces that can complicate node selection
2.  **Encoding problems**: Character encoding can cause issues with special characters
3.  **Missing data**: Elements or attributes that might be optional
4.  **Complex hierarchies**: Deeply nested structures that require careful navigation

## Handling Missing Data

Let's modify our approach to handle potential missing data:

```{r}
# Function to safely extract node text
safe_xml_text <- function(node, xpath = NULL, default = NA) {
  if (is.null(xpath)) {
    result <- tryCatch(
      xml_text(node),
      error = function(e) default
    )
  } else {
    found_node <- xml_find_first(node, xpath)
    result <- if (length(found_node) == 0) default else xml_text(found_node)
  }
  return(result)
}

# Function to safely extract attribute
safe_xml_attr <- function(node, attr, default = NA) {
  result <- tryCatch(
    xml_attr(node, attr),
    error = function(e) default
  )
  if (is.na(result) || result == "") result <- default
  return(result)
}
```

## Advanced XML Operations

For more complex XML operations, you might want to explore:

-   **XPath expressions**: More complex queries to navigate the XML structure
-   **XML validation**: Validating XML against a schema (DTD or XSD)
-   **XML transformation**: Using XSLT to transform XML into other formats

## Conclusion

This tutorial has demonstrated how to parse an XML file in R and transform its data into structured dataframes. We've created five separate dataframes that mirror the structure of typical relational database tables, making it easy to work with the data using familiar R tools.

XML parsing is a valuable skill for working with various data sources, including web APIs, configuration files, and legacy data systems.

## References and Further Reading

-   [xml2 package documentation](https://cran.r-project.org/web/packages/xml2/xml2.pdf)
-   [Introduction to XPath](https://www.w3schools.com/xml/xpath_intro.asp)
-   [R for Data Science](https://r4ds.had.co.nz/)
