## The "Books Database" {#yrBooksDatabase}

Download the CSV files for the "books database".
These CSV files together comprise the data for
the "books database". 

- [titles.csv](data/booksDatabase/titles.csv)
    
- [authors.csv](data/booksDatabase/authors.csv)
    
- [publishers.csv](data/booksDatabase/publishers.csv)
    
- [title_authors.csv](data/booksDatabase/title_authors.csv)
  
- [royalties.csv](data/booksDatabase/royalties.csv)

Download a tutorial on using Relational databases.

- [booksDatabase description and Questions](data/booksDatabaseDescription-withQuestions-ANSWERS-v024.docx)

This word document contains:

(a) a complete description of the "books database"
(b) a tutorial on how "relational databases" are structured
(c) a tutorial on Entity Relationship Diagrams (ERDs)
(d) a set of questions and answers

Once you've downloaded the various CSV files you can run the following commands to 
import the data into R.

### Import the data (books database)



```{r}
#| include: FALSE

if(!require(sqldf)){install.packages("sqldf");require(sqldf);}

file.copy(from="data/booksDatabase/authors.csv", to=".")
file.copy(from="data/booksDatabase/titles.csv", to=".")
file.copy(from="data/booksDatabase/publishers.csv", to=".")
file.copy(from="data/booksDatabase/title_authors.csv", to=".")
file.copy(from="data/booksDatabase/royalties.csv", to=".")
```


```{r}
if(!require(readr)){install.packages("readr");require(readr);}

# Read in the data for the books database - see the 
titles = read_csv("titles.csv", na="NULL", show_col_types=FALSE)
authors = read_csv("authors.csv", na="NULL", show_col_types=FALSE)
publishers = read_csv("publishers.csv", na="NULL", show_col_types=FALSE)
title_authors = read_csv("title_authors.csv", na="NULL", show_col_types=FALSE)
royalties = read_csv("royalties.csv", na="NULL", show_col_types=FALSE)
```

### Books Database - Entity Relationship Diagram (ERD)

The following is an "Entity Relationship Diagram (ERD)" that 
describes the relationships between the tables. See the Word document
linked above for more info:

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': { 'primaryColor': '#ffffff', 'primaryTextColor': '#0000ff', 'primaryBorderColor': '#000000', 'lineColor': '#000000', 'secondaryColor': '#ffffff', 'tertiaryColor': '#ffffff', 'mainBkg': '#ffffff', 'background': '#ffffff', 'edgeLabelBackground': '#ffffff' }}}%%
erDiagram
    authors ||--o{ title_authors : ""
    title_authors }|--|| titles : ""
    titles ||--o| royalties : ""
    publishers |o--o{ titles : ""
    
    authors {
        string au_id PK
        string au_fname
        string au_lname
        string phone
        string address
        string city
        string state
        string zip
    }
    
    title_authors {
        string title_id FK1
        string au_id FK2
        int au_order
        float royalty_share
    }
    
    titles {
        string title_id PK
        string title_name
        string type
        string pub_id FK
        int pages
        float price
        int sales
        date pubdate
    }
    
    publishers {
        string pub_id PK
        string pub_name
        string city
        string state
        string country
    }
    
    royalties {
        string title_id PK
        float advance
        float royalty_rate
    }
```

Below are the contents of these tables:


::: {.callout-tip collapse="true"}
#### Click here - titles table
```{r}
sqldf("select * from titles")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - publishers table
```{r}
sqldf("select * from publishers")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - title_authors table
```{r}
sqldf("select * from title_authors")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - authors table
```{r}
sqldf("select * from authors")
```
:::

::: {.callout-tip collapse="true"}
#### Click here - royalties table
```{r}
sqldf("select * from royalties")
```
:::
