---
title: "Questions"
output: 
  html_document: 
    smart: no
params:
  answers: TRUE

editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

````{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE
)

qnum=0      # question number
```

# Topic: reading and writing files



# Topic: Regular Expressions

### QUESTION `r qnum=qnum+1` `r qnum` 

df is a dataframe. Write a command that shows all rows from df for
which the 2nd character in the first column is "x".

Hints:

a.  Access a dataframe as you normally would but use grep or lgrep
    to return either the row numbers or TRUE/FALSE values that
    identify the rows to be displayed.

b.  Remember that you are NOT told what the column names are.
    Therefore you must use a number to stipulate the first column
    and NOT a column name.

#### EXAMPLE 1

```{r q1-a, collapse=TRUE}
df = data.frame( partNum = c("ax4321", "az12", "bx1234", "bw987"),
                 partName = c("widget","thingie","gadget","gizmo"),
                 price =    c(0.50, 0.60, 1.70, 0.80),
                 stringsAsFactors = FALSE)
df
```

```{r q1-message, echo=TRUE , include=TRUE}
# YOUR COMMAND GOES HERE
```

```{r q1-answer, include=TRUE, echo=FALSE}
```

```{r, q1-x, results='asis', echo=FALSE, include=params$answers}
cat("#### ANSWER")
```

```{r, q1-answer, echo=params$answers, results='hide'}
df[ grep("^.x",df[[1]]), ] 
```

---

### QUESTION `r qnum=qnum+1` `r qnum` 


```{r show1, ref.label='q1-a', eval=TRUE, echo=TRUE, collapse=TRUE}
```

```{r show2, ref.label='q1-answer', eval=TRUE, echo=TRUE, collapse=TRUE}
```

2.  Write the function:

    getCharcterAtPostion = function ( text , position )

    -   text is expected to be a character vector whose length is 1

    -   position is expected to be a number

    The function should return the the single character (i.e. symbol)
    that is at the specified position in the text. (see the EXAMPLEs)

```{r echo=FALSE}
     # ANSWER
     getCharacterAtPosition = function (text, position){
       split = strsplit(text, "")
       split = split[[1]]   # this is necessary since strsplit returns a list
       split[position]
     }
```

**EXAMPLE 1**

```{r}
    # DEFINE THE FUNCTION HERE

    getCharacterAtPosition("table", 3)
```

2.  Write the function:

    isMiddleVowel = function ( vec )

vec is expected to be a vector. The function returns a vector of those
entries in vec for which the middle character is a vowel.

HINT:

a.  use the strsplit function with "" as the split value.

```{=html}
<!-- -->
```
2.  Write the function:

    middleVowels = function ( vec )

vec is expected to be a vector. The function returns a vector of those
entries in vec for which the middle character is a vowel.

HINT:

a.  use the strsplit function with "" as the split value.

x is a character vector. Write a command that returns the entries in x
for which the middle character is a vowel (i.e. aeiou)

HINT: use the strsplit function with "" as the split value.

use the nchar function.

**EXAMPLE**

```{r}
     x = c("a", "table", "dotcoms", "a dog", "bed", "Hi Joe!")
```

     YOUR COMMAND GOES HERE

```{r echo=FALSE}
     # ANSWER
     x [ nchar(x) / 2 + 1 ]
```

------------------------------------------------------------------------

2.  x is a character vector. Write a command that returns a vector that
    contains the middle character of each entry in the vector. You may
    assume that all entries in x have an odd number of characters
    (spaces and all punctuation count as characters)

    HINT: use the nchar function.

    **EXAMPLE**

```{r}
     x = c("a", "table", "dotcoms", "a dog", "bed", "Hi Joe!")
```

     YOUR COMMAND GOES HERE

```{r echo=FALSE}
     # ANSWER
     x [ nchar(x) / 2 + 1 ]
```

------------------------------------------------------------------------

2.  Write the function

    analyzeText = function( textSource )

    text is a character vector of lines of text from a book.

```{r echo=FALSE}
     # ANSWER
     sentences = function( text ){
       allOneLine = paste0(text, collapse=" ")
       theSentences = strsplit(allOneLine, split="(?<=[?.!]) *", perl=TRUE)[[1]]
       theSentences
       
     }

     words = function( text ){
       allOneLine = paste0(text, collapse=" ")
       theWords = strsplit(allOneLine, split=" ")[[1]]
       #theWords = strsplit(allOneLine, split="\b", perl=TRUE)
       theWords
       
     }

     #book = readLines(url("https://www.gutenberg.org/files/1661/1661-0.txt"),
     #                 encoding="UTF-8")
     
     #book = book[59:length(book)]
     # book
     #words(book)
```

1.  Write a function to reverse each value in a character vector. HINTS:
    split the string into individual characters reverse the character
    vector of symbols - use lapply paste the character vectors back
    together - use sapply

2.  Write a function to replace ONLY the nth character of each entry in
    a character vector with a specified value. HINT: you can construct
    the pattern using paste and other R functions and include
    information about "n". There are many ways to do this

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
