---
title: "Regular Expression Practice Problems"
output: html_document
date: "2023-02-08"
---


```{r showAnswers, include=FALSE}

####################################################################.
####################################################################.
##
## VERY IMPORTANT!!!!
##
## Set the next line to FALSE to hide the answers.
## Set the next line to TRUE for show the answers.
##
####################################################################.
####################################################################.

#showAnswers = FALSE
showAnswers = TRUE  
```

```{css, echo = FALSE}
body { font-family: monospace; }
pre { border: 0; }
pre.yrQuestion { font-size: 1em; font-weight: normal; font-style: normal; background-color: #fff }
pre.yrAnswer { border: 4px solid black; font-size: 1.2em; font-weight: bold; font-style: italic; }
pre.yrInstructions { border: 4px solid green; font-size: 1.2em; font-weight: bold; font-style: italic; }
pre.yrData { border: 1px solid black;  }
```


```{r rmdSetup, include=FALSE}
# Code below this
# - will NOT have the # or + before the actual command.
# - WILL have # before the output of the commands

knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(include = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(prompt = TRUE)
knitr::opts_chunk$set(comment = "#")
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning = TRUE)

q <- 1

library("knitr")
knit_hooks$set(
  prompt = function(before, options, envir) {
    options(prompt = "# > ")
  }
)
```

```{r dataSetup, include=FALSE}
set.seed(1)

#firstNames = readLines("data/firstNames.txt")
#lastNames = readLines("data/lastNames.txt")

#firstMiddleLast = function(n){
  # get n random first names
  
  # get n random last names
  
  # get a different set of n random middle names
  
#}
```


```{r, class.source="yrInstructions", comment="", prompt=FALSE}
###########################################################################.
# After you're finished with the questions below,
# the following websites have some additional regex practice problems
#
#  # Test your knowledge of regex by trying to match specific challenges
#  https://regexone.com/
#  Also see: https://umar-yusuf.blogspot.com/2021/03/regexonecom-alternative-solution.html
#
#  # Make sure to read "how to play"
#  https://regexcrossword.com/
#
#  # Assumes knowledge of technical subjects (e.g. HTML hex color formats)
#  http://regextutorials.com/
#
#  # NOT WORKING: 
#  # http://play.inginf.units.it/#/level/1
#
#  # NOT GREAT
#  https://regexlearn.com/learn/regex101
#
#
###########################################################################.


############################################################
# Use the following data to answer the following questions
############################################################

# The following vector includes names of people. You may assume the
# following about the names:
#
#  - Each element of the vector only includes spaces and letters (no
#    periods or other punctuation)
#
#  - There will be exactly one space between each name in an entry
#    and no spaces at the beginning or at the end of an entry
#    (i.e. you don't have to be concerned with the possibility of
#    leading or trailing spaces)
#
#  - Each element of the vector includes either 1, 2 or 3 names where a 
#    "name" might just be a single letter (e.g. "John C Smith" has a 
#    middle name of "C")

people = c("John C Smith", 
           "J Edgar Hoover", 
           "Mike Albert Jones", 
           "Anne Taylor",
           "A B Randolph",
           "D Smith",
           "Socrates",
           "Confucious",
           "Prince")
people

# All answers must work even if the data would change.
```

```{r}
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q1, class.source="yrQuestion", comment="", prompt=FALSE}
# Use regular expressions to show only those people that have a first,middle and last
# name. 
# Even a single initial counts as a name. 
# Do not show people that don't have a middle name.
```

\# RESULTS
```{r, q1-results, class.source="yrResults", echo=FALSE}
grep("^[A-Za-z]+ [A-Za-z]+ [A-Za-z]+$", people, value=TRUE)
```

```{r, q1-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q1-results>>
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q2, class.source="yrQuestion", comment="", prompt=FALSE}
# Use regular expressions to show only those people who DO NOT have a middle
# name.
```

\# RESULTS
```{r, q2-results, class.source="yrResults", echo=FALSE}
grep("^[A-Za-z]+ [A-Za-z]+$", people, value=TRUE)
```

```{r, q2-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q2-results>>
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q3, class.source="yrQuestion", comment="", prompt=FALSE}
# Show ALL of the names. 
#
# For all people that have a First, Middle and Last name, modify their
# name to be in the format
#
#    "FM Last"
#
# where F is first letter of first name and M is first letter of middle name
# and Last is their last name.
#
# Names of people who do not have a middle name should not be changed
```

\# RESULTS

```{r, q3-results, class.source="yrResults", echo=FALSE}
threeNamesPattern = "^([A-Z])[a-z]* ([A-Z])[a-z]* ([A-Za-z]+)$"
gsub(threeNamesPattern, "\\1\\2 \\3", people)
```

```{r, q3-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q3-results>>
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q4, class.source="yrQuestion", comment="", prompt=FALSE}
# Same as previous question but this time only show those people who have
# 3 names. Show them in the format FM Last, as described above.
# Do NOT show those people who have only one or two names. 
```

\# RESULTS

```{r, q4-results, class.source="yrResults", echo=FALSE}
threeNamesPattern = "^([A-Z])[a-z]* ([A-Z])[a-z]* ([A-Za-z]+)$"
positions = grep(threeNamesPattern, people)
gsub(threeNamesPattern, "\\1\\2 \\3", people)  [ positions ]
```


```{r, q4-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q4-results>>
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q5, class.source="yrQuestion", comment="", prompt=FALSE}
# Show all the names.
# Convert people who have first middle last names to FM Last
# Convert people who have just two names to F Last
# People who only have one name should not be modified at all.
```

\# RESULTS

```{r, q5-results, class.source="yrResults", echo=FALSE}
pattern = "^([A-Z])[a-z]* (([A-Z])[a-z]* )?([A-Za-z]+)$"
gsub(pattern, "\\1\\3 \\4", people) 
```

```{r, q5-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
##########
# answer 1
##########
<<q5-results>>
```

```{r, q5-answer-2, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
##########
# answer 2
##########

# step 1 - deal with 3 word names
threeNamesPattern = "^([A-Z])[a-z]* ([A-Z])[a-z]* ([A-Za-z]+)$"
positions = grep(threeNamesPattern, people)
positions
fixed3names = gsub(threeNamesPattern, "\\1\\2 \\3", people)  [ positions ]
fixed3names

# step 2 - deal with 2 word names
twoNamesPattern = "^([A-Z])[a-z]* ([A-Za-z]+)$"
positions = grep(twoNamesPattern, people)
positions
fixed2names = gsub(twoNamesPattern, "\\1 \\2", people)  [ positions ]
fixed2names
singularNames = grep("^[A-Za-z]+$", people, value=TRUE)
singularNames

# final answer
allnames = c(fixed3names , fixed2names, singularNames)
allnames
```





\##################################################################  
\# The following questions deal with reading data from a book.  
\##################################################################  

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q6, class.source="yrQuestion", comment="", prompt=FALSE}
# The following file contains the contents of the book 
# "The Adventures of Sherlock Holmes" in UTF-8 format.
#
#      https://www.gutenberg.org/files/1661/1661-0.txt
#
# Read the lines of the book into into a variable named sherlock.
# Show the first 10 lines of the book.
```

\# RESULTS

```{r, q6-results, class.source="yrResults", echo=FALSE}
sherlock = readLines ( 'https://www.gutenberg.org/files/1661/1661-0.txt', 
                       encoding="UTF-8")

head(sherlock, 10)
```

```{r, q6-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q6-results>>
```

```{r, class.source="yrInstructions", comment="", prompt=FALSE}
# SOME BACKGROUND ABOUT ASCII FILES VS UTF-8 FILES  FOR THE FOLLOWING QUESTION
#
# The characters (i.e. letters, digits, punctuation symbols, etc) that appear
# on a standard American keyboard are known as ASCII characters.
# ASCII stands for the American Standard Code for Information Interchange.
# Specifically these include the following characters:
#
#     SPACE ! " # $ % & ' ( ) * + , - . / 0 1 2 3 4 5 6 7 8 9 : ; < = > ? @
#         A B C D E F G H I J K L M N O P Q R S T U V W X Y Z [ \ ] ^ _ ` 
#         a b c d e f g h i j k l m n o p q r s t u v w x y z { | } ~
#
# NOTE: You can generate the above list of characters with the following
# R code:
#        x=rawToChar(as.raw(32:126))
#        y=strsplit(x, "")[[1]]
#        cat(y, sep=" ")
#
# Files that store only ASCII characters are known as "ASCII files".  ASCII
# files can only store English text. By contrast UTF-8 files store
# "Unicode" characters. Unicode can represent thousands of different characters
# (as of Feb 9, 2022, the count is 144,697 different characters). These include
# the characters used in all popular languages used today, such as, Hebrew,
# Arabic, Chinese, Japanese, etc. as well as common characters such as:
#
#    - the British pound sign ( £ )
#
#    - “left and right” quotes (as opposed to "regular quotes)
#
#    - the "em dash" character ( — ) (this is longer than a typical dash, - )
#
#    - many many many other characters, eg. : ← ↑ → ↓ ↔ ∆ ∇ √ ∞ ∩ ∪ ∫ ∬
```



\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q7, class.source="yrQuestion", comment="", prompt=FALSE}
# The sherlock holmes file described above is a UTF-8 file.
# Write a function
#
#   getNotAsciiCharacters = function( theText )
#
#   theText is expected to be a vector that contains lines of text.
#
# The function should return a vector of each of the non-ASCII characters
# that appears anywhere in any element of the vector. The characters should
# only be returned one time each, even if the text contains the same 
# character many times. Sort the results.
#
# If you run the function using the entire text of
# "The Adventures of Sherlock Holmes" you will get the following results.
#
#   > getNonAsciiCharacters(sherlock)
#   [1] "—" "‘" "’" "“" "”" "£" "﻿"  "½" "à" "â" "æ" "é" "è" "œ"
```



\# RESULTS

```{r, q7-results, class.source="yrResults", echo=FALSE}
# NOTE - it would be very instructive to go through the debugger
# and look at each variable as you step through the function.
# Pay close attention to the characters in the gsub command - some
# are not obvious, e.g. 
#
#   the following need a SINGLE backslash (why not two?)  \" \' \\ 
#
#   "]" needs to be the first character in the [brackets]
#
#   "-" needs to be the last character in the brackets

getNonAsciiCharacters = function( theText ) {
  cleaned = gsub("[][ 0-9a-zA-Z~!@#$%^&*()_+`={}|`:<>?,./;'\"\\-]","", theText)
  specialChars = strsplit(cleaned, "")
  vec = unlist(specialChars)
  oneOfEach = unique(vec)
  sort(oneOfEach)
}

getNonAsciiCharacters(sherlock)
```


```{r, q7-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q7-results>>
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q8, class.source="yrQuestion", comment="", prompt=FALSE}
# Show the first 10 line numbers from the book that contain any non-ASCII characters.
# Also show the actual lines.
# You may use the function getNonAsciiCharacters that you created for the previous
# question as part of your answer.
```

```{r, q8-results, class.source="yrResults", echo=FALSE}
nac = getNonAsciiCharacters(sherlock)
pattern = paste0(nac, collapse="" )
pattern = paste0("[", pattern, "]")
lineNumbers = grep(pattern, sherlock, value=FALSE)
first10LineNumbers = head(lineNumbers, 10)
first10LineNumbers
sherlock[first10LineNumbers]
```

```{r, q8-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q8-results>>
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q9, class.source="yrQuestion", comment="", prompt=FALSE}
# Convert the vector that contains the lines from Sherlock into an ASCII text
# vector. 
#
# - change all smart quotes into regular quotes.
# - change all smart apostrophes into regular apostrophes.
# - replace the 1/2 symbols with the text " 1/2 "
# - replace the "æ" symbols with the text "ae"
# - replace the "œ" symbols with the text "oe"
# - replace the "â" and "à" symbols with the ASCII letter "a"
# - replace the  "é" and "è" symbols with the ASCII letter "e"
# - replace all "em dashes" with dashes
# - replace the British pound signs with the text " pounds "
#
# Rerun the getNonAsciiCharacters from the preivous questions to prove
# that you have removed all non-ASCII characters. The result should be character(0).

```

\# RESULTS

```{r, q9-results, class.source="yrResults", echo=FALSE}
# Copy the Unicode characters from the text above into the following
# gsub commands.

sherlock = gsub("’", "'", sherlock)
sherlock = gsub("‘", "'", sherlock)
sherlock = gsub("”", "\"", sherlock)
sherlock = gsub("“", "\"", sherlock)
sherlock = gsub("½", " 1/2 ", sherlock)
sherlock = gsub("æ", "ae", sherlock)
sherlock = gsub("œ", "oe", sherlock)
sherlock = gsub("â", "a", sherlock)
sherlock = gsub("à", "a", sherlock)
sherlock = gsub("é", "e", sherlock)
sherlock = gsub("è", "e", sherlock)
sherlock = gsub("—", "-", sherlock)
sherlock = gsub("£", " pounds ", sherlock)

getNonAsciiCharacters(sherlock)
```

```{r, q9-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q9-results>>
```



\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q10, class.source="yrQuestion", comment="", prompt=FALSE}
# How many lines are in the book?
```

\# RESULTS

```{r, q10-results, class.source="yrResults", echo=FALSE}
length(sherlock)
```

```{r, q10-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q10-results>>
```



\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q11, class.source="yrQuestion", comment="", prompt=FALSE}
# Remove all lines in the file that precede the actual text of the book.
# This includes the preamble information about project gutenberg as
# well as the table of contents.
#
# Also remove the lines at the end of the
# text that follow the end of the actual book. This includes lines about
# the gutenberg project.
# 
# We will be using regex to extract the words and the sentences 
# of the book and the extra information at the beginning and end of the file
# will just confuse matters.
#
# Put the results in a variable named sherlock2.
```

\# RESULTS

```{r, q11-results, class.source="yrResults", include=FALSE}
head(sherlock, 40)    # see which lines precede the text of the book
sherlock2 = sherlock[-(1:30)]   # remove those lines
head(sherlock2, 30)
```

```{r, q11-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q11-results>>
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q12, class.source="yrQuestion", comment="", prompt=FALSE}
# Show the last 375 lines of the text. You can see from this output that 
# the actual text of the book ends with the text:
#
#   "*** END OF THE PROJECT GUTENBERG EBOOK THE ADVENTURES OF SHERLOCK HOLMES ***"
```

\# RESULTS

```{r, q12-results, class.source="yrResults", include=FALSE}
tail(sherlock2, 375)
```

```{r, q12-answer, class.source="yrAnswer", echo=showAnswers, include=showAnswers, comment="", prompt=FALSE}
# ANSWER
<<q12-results>>
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q13, class.source="yrQuestion", comment="", prompt=FALSE}
# Determine which entry in the sherlock2 vector contains this line.

# ANSWER
grep("END OF THE PROJECT GUTENBERG EBOOK", sherlock2 )  # 11904
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q14, class.source="yrQuestion", comment="", prompt=FALSE}
# Remove everything from that position to the end of the vector

# ANSWER
sherlock2 = sherlock2[1:11903]

# Make sure that things look right
tail(sherlock2, 20)  
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q15, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION  
#
# The text of the book includes chapter headings. You can see the first 
# chapter heading by just looking at the first few lines of the book.
# For example:
#
#    > head(sherlock2, 7)
#    [1] "I. A SCANDAL IN BOHEMIA"                                               
#    [2] ""                                                                      
#    [3] ""                                                                      
#    [4] "I."                                                                    
#    [5] ""                                                                      
#    [6] "To Sherlock Holmes she is always _the_ woman. I have seldom heard him" 
#    [7] "mention her under any other name. In his eyes she eclipses and" 
#
# From this output it seems like the chapter headings consist of
# a Roman Numeral (in CAPITAL LETTERS) followed by a period 
# (e.g. I. II. III. IV. V. VI. etc.) followed by the chapter title 
# in CAPITAL LETTERS. It also seems like there are some lines that just contain
# roman numerals (e.g. line 4 above). These seem like sub-chapter headings.
# 


# Use grep to find the location of these chapter and sub-chapter heading lines
# then modify the sherlock2 variable to remove these lines. 
# 

# Use grep to find the entries in the sherlock2 vector that contain the 
# chapter headings and sub-chapter headings. Then remove those lines and resave
# the result back to the sherlock2 variable. Do not worry about the blank 
# lines. We will take care of that in a later question.



#########
# ANSWER
#########

# The first 49 Roman numerals are made up of the capital letters I, V and X
# Lets look for lines that start with one or more of these letters and are
# followed by a period.

grep("^[IVX]+\\.", sherlock2, value=TRUE)

# That looks right, now lets get the position numbers
# and remove them from the sherlock2 vector

positions = grep("^[IVX]+\\.", sherlock2, value=FALSE)
positions
sherlock2 = sherlock2[ -positions ]

# This should no longer find those lines
grep("^[IVX]+\\.", sherlock2, value=TRUE)

# And the first chapter heading is obviously gone
sherlock2[1:10]
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q16, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# Use the nchar function to show the number of characters (including
# spaces and punctuation) on each line of sherlock2

# answer
nchar(sherlock2)
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q17, class.source="yrQuestion", comment="", prompt=FALSE}
# Question
#
# The character vector sherlock2 currently contains one entry per
# line of the book.
#
# (a) Create a variable named sherlock3 that contains a character vector that 
#     has all of the lines of the book in a SINGLE entry. Place a space
#     between each line of the book.
#
#     Use the nchar function to count the number of characters in the book.
#     (this includes every character, including spaces and punctuation)
#
# (b) Show that this number equals the sum of the numbers of characters
#     on the lines of sherlock2 plus one space between each line.






# answer (a)

# The following line will put the ENTIRE BOOK in a single value.
#
# WARNING - it is NOT a good idea to type the value sherlock3 to display
# the entire book to the screen - it may slow down your R session a lot.
# If that happens just restart RStudio and you should be back to normal.
# (see below for a little more detail about this ...)

sherlock3 = paste(sherlock2, collapse=" ")

# EXPLANATION - remember how the paste function works.
# Refer to the following examples of using paste to refresh your memory.
first = c("joe", "juan", "shalom")
last = c("smith", "rodriguez", "cohen")
paste(first, last)
paste(first, last, sep="-")
paste(first, last, sep="")
paste(first, last, sep="", collapse="-")
paste(first)
paste(first, collapse="-")

# We can prove that the entire book is in a single character vector. The length
# function returns 1 since the entire book is in one character value

length(sherlock3)  # 1

nchar(sherlock3)   # This is the number of characters in the book

# Warning - it is NOT a good idea to display sherlock3 to the screen.
# Because it is such a long piece of data in a single character entry
# displaying the value will cause R to slow down a lot. This has to do
# with the way R manages the computer's memory - don't worry about the 
# details, but just be aware that if you try to display sherlock3
# to the screen that R can slow down a lot. If you do happen to 
# display sherlock3 to the screen, simply restart RStudio and 
# things should be back to normal.
#
# Instead of displaying the entire value (i.e. the whole book) you can
# instead display parts of the value by using the substring
# function (see ?substring). For example:
#
#   > someData = c("123456789",
#                  "abcdefghijklmnopqrstuvwxyz",
#                  "An apple a day keeps the doctor away.")
#
#   > substring(someData, 4, 8)   # display the 4th through 8th characters
#   [1] "45678"    "defgh"      "apple"
#
#   > substring(someData, 4, 14)   # display the 4th through 14th characters
#   [1] "456789"   "defghijklmn"   "apple a day"


# display the first 1000 characters of the book
substring(sherlock3, 1,1000) 

# display the from character 1001 to the 2000th character of the book
substring(sherlock3, 1001,2000) 


# display the last 1000 characters of the book
substring(sherlock3, nchar(sherlock3) - 999, nchar(sherlock3)) 


# See these online resources for an explanation of the substring
# (and the related function substr) functions.
#
#   https://www.youtube.com/watch?v=FyWG954OytA
#
#   https://www.geeksforgeeks.org/extracting-substrings-from-a-character-vector-in-r-programming-substring-function/







# answer (b)

sum(nchar(sherlock2), length(sherlock2) - 1)

# The above line sums up the total number of characters on each line.
# It also adds one for each space that will be added when you combine 
# the different lines. There are one fewer spaces added than the number
# of lines in the book - since there is no space added after the last line.
```






\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q18, class.source="yrQuestion", comment="", prompt=FALSE}
# Question
#
# Create a character vector named wordsPlusPunctuation that contains
# the values from sherlock3 split up by spaces. 
# Show the first 100 entries of this vector.
#
# hint: use strsplit (see ?strsplit)
#       Don't forget that strsplit returns a LIST and NOT a vector. 
#       Therefore use [[1]] to get the first item from the list
#       i.e. the vector that we want that contains the words.
#       You could also use the unlist function.

# answer 1 - using [[1]]
wordsPlusPunctuation = strsplit(sherlock3 , " ")[[1]]  
head(wordsPlusPunctuation, 300)

# answer 2 - using unlist function
wordsPlusPunctuation = unlist(strsplit(sherlock3 , " "))
head(wordsPlusPunctuation, 300)
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q19, class.source="yrQuestion", comment="", prompt=FALSE}
# Question
#
# The book "The Adventures of Sherlock Holmes" only contains a single space
# between each word of the book. How would your answer to the 
# previous question change if there there were possibly many spaces between
# each word. For example, the variable linesOfPoem, contains lines from
# a poem that has several spaces between some words. (This is apparently
# done for dramatic effect and artistic license).
#
# linesOfPoem = c("Long                pauses               sharpen the mind.",
#                 "They pepper speech with,                    thinking time.",
#                 "Pauses                                                    ",
#                 "Pauses                                                    ",
#                 "They are much about                               nothing.")
#
# Create a variable named, wordsOfPoem, that contains just the individual 
# words of the poem. Make sure that your answer does NOT include empty
# character values. (see below)
#
#   > wordsOfPoem = YOUR COMMAND GOES HERE
#   > wordsOfPoem
#   [1] "Long"     "pauses"   "sharpen"  "the"      "mind."    "They"    
#   [7] "pepper"   "speech"   "with,"    "thinking" "time."    "Pauses"  
#   [13] "Pauses"   "They"     "are"      "much"     "about"    "nothing."

###########
# answer
###########

# The pattern in the strsplit should be " +" instead of just " ".

poem = c("Long                pauses               sharpen the mind.",
         "They pepper speech with,                    thinking time.",
         "Pauses                                                    ",
         "Pauses                                                    ",
         "They are much about                               nothing.")

wordsOfPoem = unlist(strsplit(poem, " +"))    # pattern mean one or more spaces
wordsOfPoem

# note if you don't include the + in the pattern then you will be many
# empty character values in the results
unlist(strsplit(poem, " "))  # this is NOT the correct answer
```



\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q20, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# Create a vector named words that contains the words from the book without
# any additional punctuation marks. Make sure to get rid of any periods, 
# commas, quotation marks, etc. Specifically, get rid of all of the following 
# characters from the words:
#
#   . , ! @ # $ % ^ & * ( ) + = { } " ~ ` ' : ? -
#
# Display the first 300 entries in words.
#
# Hint: use the gsub function


###########
# answer 
###########

# this answer uses a character class, i.e. [...]
# note that you do should NOT type backslashes before regex-meta-characters that
# are in a character class.
#
# In addition:
#   - make sure NOT to put ^ at the beginning.
#   - make sure NOT to put - in the middle
#   - The " has a single backslash in front of it to be able to include it in
#     an R character string that itself is surrounded by "double quotes".

punctuation = "[.,!@#$%^&*()+={}\"~`':?-]"
words = gsub(punctuation, "", wordsPlusPunctuation)
head(words, 300)
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q21, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
# 
# The tolower function converts a character vector to lowercase.
# Create a vector, uniqueWords, that contains a single copy of each word in 
# lowercase form. 
# The vector should be sorted.
# Show the first 200 entries and last 50 entries of the vector.

#########
# ANSWER
#########
words2 = sort(unique(tolower(words)))
head(words2, 200)
tail(words2, 50)
```



\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q22, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# Return to the vector, wordsPlusPunctuation ...
# Create a variable named abbrevs that contains each of the abbreviations
# that are used in the book.
#
# To do so we will do the following:
#
#   a. find all words that start with a capital letter and end with a period
#      (we will assume that in formal writing, most abbreviations follow
#      this pattern) This may include some non-abbreviations such as a name
#      that comes at the end of a sentence, but it will make the list 
#      much shorter to look through.
#
#   b. eliminate duplicates from this list
#
#   c. look through the list and come up with a list of abbreviations

############
# ANSWER
############
potentialAbbreviations = grep ("^[A-Z].*\\.$", wordsPlusPunctuation, value=TRUE)
potentialAbbreviations
potentialAbbreviations2 = sort(unique(potentialAbbreviations))
potentialAbbreviations2

# Looking through the above, the following look like the abbreviations that
# were used. Some of these are roman numerals, some are single letters which
# are probably used as the initials of people, some are recognizable 
# abbreviations such as: Dr. Esq. Mr. MR. Mrs. No. Pa. St. ST. 
#
#    A. C. Dr. Esq. F. H. I. II. III. IV. IX.
#    J. K. Mr. MR. Mrs. No. O. Pa. S. 
#    St. ST. U. V. VI. VII. VIII. X. XI. XII.


abbrevs = c("A", "C", "Dr", "Esq", "F", "H", "I", "II", "III", "IV", "IX",
            "J", "K", "Mr", "MR", "Mrs", "No", "O", "Pa", "S", 
            "St", "ST", "U", "V", "VI", "VII", "VIII", "X", "XI", "XII")

sort(abbrevs)
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q24, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# Show the lines and line numbers from the original book that contain
# the abbreviation "Esq."

# answer
grep("Esq\\.", sherlock, value=TRUE)
grep("Esq\\.", sherlock, value=FALSE)
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q25, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# The single letters A C F H I J. K. O. S. U. V. came up in the results of the last question.
# Show all lines from the original book (stored in sherlock) that contains the text
# "A."   This will allow us to understand if J. is indeed a person's initial or if it 
# is used in some other context.
#
# show the line number that contains this information
#
# Do the same for the other letters, S.   O.   and   U.
#
# Finally write a SINGLE command that shows all of these lines.


# answer
grep("A\\.", sherlock, value=TRUE)
grep("A\\.", sherlock, value=FALSE)

grep("S\\.", sherlock, value=TRUE)
grep("S\\.", sherlock, value=FALSE)

grep("O\\.", sherlock, value=TRUE)
grep("O\\.", sherlock, value=FALSE)

grep("U\\.", sherlock, value=TRUE)
grep("U\\.", sherlock, value=FALSE)

grep("J\\.|S\\.|O\\.|U\\.", sherlock, value=TRUE)
grep("J\\.|S\\.|O\\.|U\\.", sherlock, value=FALSE)
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q26, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
# 
# Create a vector that contains the SENTENCES from the book.
# Do the following steps to accomplish this:
#
#   a. Use the variable sherlock3 that contains the entire book in one
#      character entry. Replace any period that appears after any of
#      the abbreviations with the text "<<PERIOD_PLACE_HOLDER>>"
#
#   b. Insert the text <<BREAK_HERE>> after each remaining period and after
#      each question mark and exclamation point.
#
#   c. use strsplit to split the text into sentences wherever the
#      text "<<BREAK_HERE>>" appears.
#
#   d. Replace the text <<PERIOD_PLACE_HOLDER>> with an actual period.
#
#   e. remove leading spaces from the sentences



# answer (a)
#   a. Use the variable sherlock3 that contains the entire book in one
#      character entry. Replace any period that appears after any of
#      the abbreviations with the text "<<PERIOD_PLACE_HOLDER>>"

# Create the following function to modify all abbreviations in the text.
# Replace all periods that follow one of the abbreviations
# with a unique piece of text. We will use this function on the sherlock
# holmes text below.

modifyAbbrevs = function(text, abbrevs, periodPlaceHolder){
  for (a in abbrevs) {
    pattern = paste0(a, "\\.")
    replacement = paste0(a, periodPlaceHolder)
    text = gsub(pattern, replacement, text)
  }
  text
}

# Use the function to replace all periods that follow one of the
# abbreviations with <<PERIOD_PLACE_HOLDER>>
sherlock4 = modifyAbbrevs(sherlock3, abbrevs, "<<PERIOD_PLACE_HOLDER>>")

# Show the first 1000 characters of the new sherlock4 variable
substring(sherlock4, 1, 1000)



# answer (b)
#   b. Insert the text <<BREAK_HERE>> after each remaining period and after
#      each question mark and exclamation point.


# Add "<<BREAK_HERE>>" after each period, question mark or 
# exclamation point. 

sherlock5 = gsub("([.?!])", "\\1<<BREAK_HERE>>", sherlock4)
substring(sherlock5, 1, 1000)


# answer (c)
#   c. use strsplit to split the text into sentences wherever the
#      text "<<BREAK_HERE>>" appears.

sentences = strsplit(sherlock5, "<<BREAK_HERE>>")[[1]]

# Show the first 10 sentences. Note that there is an issue with the 
# first "sentence". We'll fix that soon ...
head(sentences, 10) 

# The first "sentence" contains the table of contents before the actual
# sentence.

sentences[1]

# answer (d)
#   d. Replace the text <<PERIOD_PLACE_HOLDER>> with an actual period.

sentences2 = gsub("<<PERIOD_PLACE_HOLDER>>", ".", sentences)

#   e. remove leading spaces from the sentences

# All sentences (except the first) have a leading space.
# This is the result of splitting after a .? or ! but not removing the
# following space. Let's remove all leading spaces from the resulting
# sentences.
sentences3 = gsub("^ +", "", sentences2)

head(sentences3, 10)



grep("^[IVX]+\\.$", sherlock, value=TRUE)
grep("^[IVX]+\\.$", sherlock)

grep("The Adventure of the", sherlock, value=TRUE)
grep("The Adventure of the", sherlock, value=TRUE)


grep("^[IVX]+\\. +[A-Z' -]+$", sherlock, value=TRUE)
grep("^[IVX]+\\. +[A-Z' -]+$", sherlock)
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q27, class.source="yrQuestion", comment="", prompt=FALSE}
# Question
# 
# create a variable quotations that pulls out all quotations from the original book, sherlock.
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q28, class.source="yrQuestion", comment="", prompt=FALSE}
# Question
#
# Find the number of times that Holmes is mentioned in the text.
```


```{r}
###########################################
# Use the data for the following questions
###########################################

data = c("abc 12345",
         "abc 1",
         "abc e")
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q30, class.source="yrQuestion", comment="", prompt=FALSE}
#---------------------------------------------------------------------
# Write a command that displays each value of data, one line at a time
#
# Hint: use cat with a sep argument
#---------------------------------------------------------------------

# ANSWER
cat(data, sep="\n")
```


\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q31, class.source="yrQuestion", comment="", prompt=FALSE}
#---------------------------------------------------------------------
# Explain why the following commands produce what they do
#---------------------------------------------------------------------

# Explain why the following command produces the output shown below.

data
cat ( gsub("c \\d?", "X", data ) , sep="\n")

# abX2345
# abX
# abXe


cat ( gsub("c \\d*", "X", data ) , sep="\n")

# abX
# abX
# abXe
```

\##############################<br>
\# QUESTION `r q` `r q=q+1`    <br>
\##############################<br>

```{r, q32, class.source="yrQuestion", comment="", prompt=FALSE}
# QUESTION
#
# Use grep to show the entries in a vector that do not contain
# any digits.

# answer

sampleData = c("hello there", "I eat 3 bananas every day.")
pattern = "^[^0-9]*$"
grep(pattern, sampleData, value=TRUE)
```